A neurotransmitter atlas of the nervous system of C. elegans males and hermaphrodites

Version 1

Abstract

Abstract

Assigning neurotransmitter identity to neurons is key to understanding information flow in a nervous system. It also provides valuable entry points for studying the development and plasticity of neuronal identity features. Neurotransmitter identities have been assigned to most neurons in the C. elegans nervous system through the expression pattern analysis of neurotransmitter pathway genes that encode neurotransmitter biosynthetic enzymes or transporters. However, many of these assignments have relied on multicopy reporter transgenes that may lack relevant cis-regulatory information and therefore may not provide an accurate picture of neurotransmitter usage. We analyzed the expression patterns of 13 CRISPR/Cas9-engineered reporter knock-in strains, which report on the deployment of all main types of neurotransmitters in C. elegans (glutamate, acetylcholine, GABA, serotonin, tyramine, and octopamine) throughout the entire nervous system of both the hermaphrodite and the male. Our analysis reveals novel sites of expression of these neurotransmitter systems within both neurons and glia and defines neurons that may be exclusively neuropeptidergic. Furthermore, we also identified unusual combinations of expression of monoaminergic synthesis pathway genes, suggesting the existence of novel monoaminergic transmitters. Our analysis results in what constitutes the most extensive nervous system-wide map of neurotransmitter usage to date, paving the way for a better understanding of neuronal communication in C. elegans.

Introduction

Understanding information processing in the brain necessitates the generation of precise maps of neurotransmitter deployment. Moreover, comprehending synaptic wiring diagrams is contingent upon decoding the nature of signaling events between anatomically connected neurons. Mapping of neurotransmitter identities onto individual neuron types also presents a valuable entry point for studying how neuronal identity features become genetically specified during development and potentially modified in response to specific external factors (such as the environment) or internal factors (such as sexual identity or neuronal activity patterns).

The existence of complete synaptic wiring diagrams of the compact nervous system of male and hermaphrodite C. elegans nematodes raises questions about the molecular mechanisms by which individual neurons communicate with each other. C. elegans employs the main neurotransmitter systems that are used throughout the animal kingdom, including acetylcholine, glutamate, γ-aminobutyric acid (GABA), and several monoamines (Sulston et al. 1975; Horvitz et al. 1982; Loer and Kenyon 1993; McIntire et al. 1993; Duerr et al. 1999; Lee et al. 1999; Duerr et al. 2001; Alkema et al. 2005; Duerr et al. 2008; Serrano-Saiz et al. 2013; Pereira et al. 2015; Gendrel et al. 2016; Serrano-Saiz et al. 2017b)(Fig. 1A). Efforts to map these neurotransmitter systems to individual cell types throughout the entire nervous system have a long history, beginning with the use of chemical stains that directly detected a given neurotransmitter (dopamine)(Sulston et al. 1975), followed by antibody staining of neurotransmitter themselves (serotonin and GABA)(Horvitz et al. 1982; McIntire et al. 1993) or antibody stains of biosynthetic enzymes or neurotransmitter vesicular transporters (acetylcholine and monoamines)(Loer and Kenyon 1993; Duerr et al. 1999; Duerr et al. 2001; Alkema et al. 2005; Duerr et al. 2008)(see Fig. 1A for an overview of these enzymes and transporters).

Background on genes examined in this paper.

(A) Neurotransmitter synthesis and transport pathway. TH = tyrosine hydroxylase; TDC = tyrosine decarboxylase; TBH = tyramine beta hydroxylase; TPH = tryptophan hydroxylase; GAD = glutamic acid decarboxylase; AAAD = aromatic amino acid decarboxylase; VMAT = vesicular monoamine transporter; VAChT = vesicular acetylcholine transporter; VGAT = vesicular GABA transporter. Taken and modified from (Hobert 2013).

(B) Graphic comparison of scRNA expression data and previously reported reporter expression data. See Table S1 for a more comprehensive version that includes expression of reporter genes in cells that show no scRNA transcripts. Note that scRNA expression values for eat-4 can be unreliable because eat-4 was overexpressed to isolate individual neurons for scRNA analysis (Taylor et al. 2021).

While these early approaches proved successful in revealing neurotransmitter identities, they displayed several technical limitations. Since neurotransmitter-synthesizing or transporting proteins primarily localize to neurites, the cellular identity of expressing cells (usually done by assessing cell body position) often could not be unambiguously established in several, particularly cell-dense regions of the nervous system. One example concerns cholinergic neurons, which are defined by the expression of the vesicular acetylcholine transporter UNC-17/VAChT and choline acetyltransferase CHA-1/ChAT. While mainly neurite-localized UNC-17 and CHA-1 antibody staining experiments could identify a subset of cholinergic neurons (Duerr et al. 2001; Duerr et al. 2008), many remained unidentified (Pereira et al. 2015). In addition, for GABA-producing neurons, it became apparent that antibody-based GABA detection was dependent on staining protocols, leading to the identification of “novel” anti-GABA-positive neurons, i.e. GABAergic neurons, more than 20 years after the initial description of GABAergic neurons (McIntire et al. 1993; Gendrel et al. 2016).

An alternative approach to mapping neurotransmitter usage has been the use of reporter transgenes. This approach has the significant advantage of allowing the fluorophore to either fill the entire cytoplasm of a cell or to be targeted to the nucleus. However, one shortcoming of transgene-based reporter approaches is that one cannot be certain that a chosen genomic region, fused to a reporter gene, indeed contains all cis-regulatory elements of the respective locus. In fact, the first report that described the expression of the vesicular glutamate transporter EAT-4, the key marker for glutamatergic neuron identity, largely underestimated the number of eat-4/VLGUT-positive and, hence, glutamatergic neurons (Lee et al. 1999). The introduction of fosmid-based reporter transgenes has largely addressed such concerns, as these reporters, with their 30-50 Kb size, usually cover entire intergenic regions (Sarov et al. 2012). Indeed, such fosmid-based reporters have been instrumental in describing the entire C. elegans glutamatergic nervous system, defined by the expression of eat-4/VGLUT (Serrano-Saiz et al. 2013), as well as supposedly the complete set of cholinergic (Pereira et al. 2015) and GABAergic neurons (Gendrel et al. 2016).

However, even fosmid-based reporters may not be the final word. In theory, they may still miss distal cis-regulatory elements. Moreover, the multicopy nature of transgenes harbors the risk of overexpression artifacts, such as the titrating of rate-limiting negative regulatory mechanisms. Also, RNAi-based silencing mechanisms triggered by the multicopy nature of transgenic reporter arrays have the potential to dampen the expression of reporter arrays (Nance and Frokjaer-Jensen 2019). One way to get around these limitations, while still preserving the advantages of reporter gene approaches, is to generate reporter alleles in which an endogenous locus is tagged with a reporter cassette, using CRISPR/Cas9 genome engineering. Side-by-side comparisons of fosmid-based reporter expression patterns with those of knock-in reporter alleles indeed revealed several instances of discrepancies in expression patterns of homeobox genes (Reilly et al. 2022).

Another indication that previous neurotransmitter assignments may not have been complete was provided by recent single-cell RNA (scRNA) transcriptomic analyses of the hermaphrodite nervous system (Taylor et al. 2021). As we describe in this paper in more detail, transcripts for several neurotransmitter-synthesizing enzymes or transporters were detected in a few cells beyond those previously described to express the respective genes. This motivated us to use CRISPR/Cas9 engineering to fluorescently tag a comprehensive panel of genetic loci that code for neurotransmitter-synthesizing and -transporting proteins (“neurotransmitter pathway genes”). Using the landmark strain NeuroPAL for neuron identification (Yemini et al. 2021), we identified novel sites of expression of most neurotransmitter pathway genes. Furthermore, we used these reagents to expand and refine neurotransmitter maps of the entire nervous system of the C. elegans male, which contains almost 30% more neurons than the nervous system of the hermaphrodite yet lacks a reported scRNA transcriptome atlas. Together with the NeuroPAL cell identification tool, these reporter alleles allowed us to substantially improve the previously described neurotransmitter map of the male nervous system (Serrano-Saiz et al. 2017b). Our analysis provides insights into the breadth of usage of each individual neurotransmitter system, reveals instances of co-transmitter use, indicates the existence of neurons that may entirely rely on neuropeptides instead of classic neurotransmitters, suggests the likely existence of presently unknown neurotransmitters, and reveals sexual dimorphisms in neurotransmitter usage.

Materials and methods

Transgenic reporter strains

Knock-in reporter alleles were generated either by SunyBiotech (syb alleles) or in-house (ot alleles) using CRISPR/Cas9 genome engineering. A nuclear-targeted gfp-coding sequence (gfp fused to his-44, a histone H2B-encoding gene) was inserted at the 3’ end of the respective locus, as shown in Fig. 2. The gfp reporter was separated from the rest of the coding gene either with a self-cleaving T2A sequence or an SL2 sequence.

Schematics of reporter knock-in alleles.

Reporter alleles were generated by CRISPR/Cas9 genome engineering. The SL2 or T2A-based separation of the reporter from the coding sequence of the respective loci enables targeting of the reporter to the nucleus (via the H2B tag), which in turn facilitates identification of the cell expressing a given reporter.

Reporter alleles generated in this study:

unc-25(ot1372[unc-25::t2a:gfp::h2b]) III

unc-46(syb7278[unc-46::sl2::gfp:h2b]) V

unc-47(syb7566[unc-47::sl2::gfp::h2b]) III

cat-1(syb6486[cat-1::sl2::gfp::h2b]) X

tph-1(syb6451[tph-1::sl2::gfp::h2b]) II

tbh-1(syb7786[tbh-1::sl2::gfp::h2b]) X

tdc-1(syb7768[gfp::linker::h2b::t2a::tdc-1]) II

cat-2(syb8255[cat-2::sl2::gfp::h2b]) II

snf-3(syb7290[snf-3::tagRFP::sl2::gfp::h2b]) II hdl-1(syb1048[hdl-1::gfp]) IV

hdl-1(syb4208[hdl-1::t2a::3xnls::cre]) IV

Since we did not detect fluorophore signals in the hdl-1(syb1048[hdl-1::gfp] strain, we attempted to amplify low level signals, by inserting Cre recombinase at the C-terminus of the hdl-1 locus (hdl-1(syb4208[hdl-1::t2a::3xnls::cre])). We crossed this strain to the recently published “Flexon” strain (arTi361[rps-27p::gfp“flexon”-h2b::unc-54-3’UTR]) (Shaffer and Greenwald 2022). Even low expression of hdl-1 should have led to Cre-mediated excision of the flexon stop cassette, which is designed to abrogate gene expression by a translational stop and frameshift mutation, and subsequently can result in strong and sustained gfp expression under the control of the rps-27 promoter and thereby providing information about cell-specific hdl-1 expression. However, no robust, consistent reporter expression was seen in hdl-1(syb4208[hdl-1::t2a::3xnls::cre]); arTi361[rps-27p::gfp“flexon”-h2b::unc-54-3’UTR] animals.

Three of the reporter alleles that we generated were already previously examined in specific cellular contexts:

unc-17(syb4491[unc-17::t2a::gfp:h2b]) IV (Vidal et al. 2022)

eat-4(syb4257[eat-4::t2a::gfp::h2b]) III (Vidal et al. 2022)

bas-1(syb5923[bas-1::sl2::gfp::h2b]) III (Yu et al. 2023)

Microscopy and image processing

For adult animal imaging, same-sex L4 worms were grouped on NGM plates 6-9 hours prior to imaging to control for accurate staging and avoid mating. Young adult worms were then anesthetized using 50—100 mM sodium azide and mounted on 5% agarose pads on glass slides. Z-stack images were acquired with ZEN software using Zeiss confocal microscopes LSM880 and LSM980 or a Zeiss Axio Imager Z2 and processed with ZEN software or FIJI (Schindelin et al. 2012) to create orthogonal projections. Brightness and contrast, and in some cases gamma values, were adjusted to illustrate dim expression and facilitate neuron identification.

Neuron class identification

Neuron classes were identified by crossing the gfp reporter alleles with the landmark strain “NeuroPAL” (allele otIs669 or otIs696, for bright reporters and dim reporters, respectively) and following published protocols (Tekieli et al. 2021; Yemini et al. 2021) (also see “lab resources” at hobertlab.org). For neuron identification of the eat-4(syb4257), unc-46(syb7278), and unc-47(syb7566) reporter alleles in hermaphrodites, the reporter alleles were also crossed into the fosmid-based reporter transgenes of the same gene [eat-4(otIs518), unc-46(otIs568), and unc-47(otIs564)] as a “first-pass” to identify potential non-overlapping expression of the two alleles. For tph-1(syb6451) analysis, an eat-4 fosmid-based reporter (otIs518) was also used. For identification of VC4, VC5, and HSN, an ida-1p::mCherry integrant (LX2478, lin-15(n765ts) X; vsls269[ida-1::mCherry]) was also used in some cases (Fernandez et al. 2020).

Results

Comparing scRNA data to reporter gene data

To investigate the neurotransmitter identity of neurons throughout the entire C. elegans nervous system of both sexes, we consider here the expression pattern of the following 13 genetic loci (see also Fig. 1A):

eat-4/VGLUT: expression of the vesicular glutamate transporter is alone sufficient to define glutamatergic neuron identity (Lee et al. 1999; Serrano-Saiz et al. 2013).

unc-17/VAChT: expression of the vesicular acetylcholine transporter, located in an operon together with the acetylcholine synthesizing gene cha-1/ChAT (Alfonso et al. 1994), defines cholinergic neurons (Duerr et al. 2001; Duerr et al. 2008; Pereira et al. 2015).

unc-25/GAD, unc-47/VGAT and its sorting co-factor unc-46/LAMP: expression of these three genes defines neurons that synthesize and release GABA (McIntire et al. 1993; McIntire et al. 1997; Jin et al. 1999; Schuske et al. 2007; Gendrel et al. 2016). Additional neurons that we classify as GABAergic are those that do not synthesize GABA (unc-25/GAD-negative), but take up GABA from other neurons (based on anti-GABA antibody staining) and are expected to release GABA based on unc-47/VGAT expression (Gendrel et al. 2016). unc-47/VGAT expression without any evidence of GABA synthesis or uptake (unc-25/GAD- and anti-GABA-negative) is indicative of an unknown inhibitory transmitter being present in these cells and utilizing unc-47/VGAT for vesicular secretion.

tph-1/TPH and bas-1/AAAD: the co-expression of these two biosynthetic enzymes defines, together with the co-expression of the monoamine vesicular transporter cat-1/VMAT, all serotonergic neurons (Fig. 1A)(Horvitz et al. 1982; Duerr et al. 1999; Sze et al. 2000; Hare and Loer 2004).

cat-2/TH and bas-1/AAAD: the co-expression of these two biosynthetic enzymes defines, together with the co-expression of the monoamine vesicular transporter cat-1/VMAT, all dopaminergic neurons (Fig. 1A)(Sulston et al. 1975; Duerr et al. 1999; Lints and Emmons 1999; Hare and Loer 2004).

tdc-1/TDC: defines, together with cat-1/VMAT, all tyraminergic neurons (Fig. 1A)(Alkema et al. 2005).

tbh-1/TBH: expression of this gene, in combination with that of tdc-1/TDC and cat-1/VMAT, defines octopaminergic neurons (Fig. 1A) (Alkema et al. 2005).

cat-1/VMAT: expression of this vesicular monoamine transporter defines all four above-mentioned monoaminergic neurons (serotonin, dopamine, tyramine, octopamine)(Duerr et al. 1999), but as described and discussed below, may also define additional sets of monoaminergic neurons.

hdl-1/AAAD: hdl-1, a previously uncharacterized gene, encodes the only other AAAD with sequence similarity to the bas-1 and tdc-1 AAAD enzymes that produce other bona fide monoamines (Fig. S2)(Hare and Loer 2004). hdl-1 expression may therefore, in combination with cat-1/VMAT, identify neurons that produce and release trace amines of unknown identity.

snf-3/BGT-1/SLC6A12: this gene encodes the functionally validated orthologue of the vertebrate betaine uptake transporter SLC6A12 (i.e. BGT1)(Peden et al. 2013). In combination with the expression of cat-1/VMAT, which synaptically transports betaine (Hardege et al. 2022), snf-3 expression may identify neurons that utilize betaine as a synaptically released neurotransmitter to gate betaine-gated ion channels, such as ACR-23 (Peden et al. 2013) or LGC-41 (Hardege et al. 2022).

For all these 13 genetic loci, we compared scRNA transcriptome data from the CeNGEN scRNA atlas (at all 4 available stringency levels (Taylor et al. 2021)) to previously published reporter and antibody staining data. As shown in Fig. 1B and Table S1, such comparisons reveal the following: (1) scRNA data support the expression of genes in the vast majority of neurons in which those genes were found to be expressed with previous reporter gene approaches. In most cases, this is true even at the highest threshold levels for scRNA detection. (2) Vice versa, reporter gene expression supports scRNA transcriptome data for a specific neurotransmitter system in the great majority of cells. (3) In spite of this congruence, there were several discrepancies between reporter data and scRNA data. Generally, while valuable, scRNA transcriptome data cannot be considered the final word for any gene expression pattern assignments. Lack of detection of transcripts could be a sensitivity issue and, conversely, the presence of transcripts does not necessarily indicate that the respective protein is generated, due to the possibility of posttranscriptional regulation.

Hence, to consolidate and further improve neurotransmitter identity assignment throughout the entire C. elegans nervous system, and to circumvent potential limitations of multicopy, fosmid-based reporter transgenes on which previous neurotransmitter assignments have been based, we engineered and examined expression patterns of 13 reporter knock-in reporter alleles of the neurotransmitter synthesis and vesicular transporter loci listed above (Fig. 1, Fig. 2). For unc-17, eat-4, and unc-25, we knocked-in a t2a::gfp::h2b (his-44) cassette right before the stop codon of the respective gene. For tdc-1, a gfp::h2b::t2a cassette was knocked into the N-terminus of the locus because of different C-terminal splice variants. The self-cleaving T2A peptide frees up GFP::H2B, which will be transported to the nucleus, thereby facilitating cell identification. For other reporter alleles (unc-46, unc-47, tph-1, bas-1, tbc-1, cat-1, cat-2, snf-3), we knocked-in a sl2::gfp::h2b cassette at the C-terminus of the locus. Both types of reporter cassettes should capture posttranscriptional, 3’UTR-mediated regulation of each locus, e.g. by miRNAs (not captured by scRNA data). Since the reporter is targeted to the nucleus, this strategy circumvents shortcomings associated with interpreting antibody staining patterns or dealing with too densely packed cytosolic signals. For all our neuronal cell identification, we utilized the neuronal landmark strain NeuroPAL (Tekieli et al. 2021; Yemini et al. 2021).

Expression of a reporter allele of eat-4/VGLUT, a marker for glutamatergic identity, in the hermaphrodite

37 of the 38 previously reported neuron classes that express an eat-4 fosmid-based reporter (Serrano-Saiz et al. 2013) showed eat-4 transcripts in the CeNGEN scRNA atlas (Taylor et al. 2021) at all 4 thresholds of stringency and 1/38 (PVD neuron) showed it in 3 out of the 4 threshold levels (Fig. 1B, Table S1). However, scRNA transcripts were detected at all 4 threshold levels in three additional neuron classes, RIC, PVN, and DVA for which no previous reporter data provided support. In a recent publication, we had already described that the eat-4 reporter allele syb4257 is expressed in RIC (Reilly et al. 2022)(confirmed in Fig. 3A). We now also confirm expression of this reporter allele, albeit at low levels, in DVA and PVN (Fig. 3B, Table S2).

Expression of eat-4/VGLUT and unc-17/VAChT reporter alleles in the adult hermaphrodite.

Neuronal expression of eat-4(syb4257) and unc-17(syb4491) was characterized with landmark strain NeuroPAL (otIs696 and otIs669, respectively). Only selected neurons are shown for illustrating updates from previous reports. See Table S2 for a complete list of neurons.

(A) Dim expression of eat-4(syb4257) in head neurons ASK and ADL is consistent with previous fosmid-based reporter expression. RIC expression is consistent with previous observation using the same reporter allele (Reilly et al. 2022). In addition, dim expression is detected in pharyngeal neuron M5, previously not detected with eat-4 GFP fosmid-based reporter (otIs388) but visible with eat-4 mCherry fosmid-based reporter (otIs518).

(B) Previously uncharacterized eat-4 expression in PDE and DVA neurons is detected with the eat-4(syb4257) reporter allele. Variable expression in PHA is also occasionally detected. No expression is detected in PVQ. Expression in PVN is detected in both sexes but at a much higher level in the male.

(C) In the head, prominent expression of unc-17(syb4491) in RIP and dim expression in AWA and AFD neurons are detected. There is no visible expression in RIB, FLP, or AVJ. Consistent with previous reports, AIM expresses unc-17 only in males and not hermaphrodites. In addition, very dim expression of AVG can be detected occasionally in hermaphrodites and slightly stronger in males.

(D) In the tail, PVN expresses unc-17(syb4491) in both sexes, consistent with previous reports. Scale bars, 10 μm in A, C, and D; 5 μm in B. Quantification in B is done by normalizing fluorescent intensity of eat-4 GFP to that of the blue channel in the NeuroPAL background. Statistics, Mann-Whitney test.

Another neuron found to have some eat-4 transcripts, but only with the two lower threshold sets, is the I6 pharyngeal neuron. Consistent with our previous fosmid-based reporter data, we detected no I6 expression with our eat-4 reporter allele. The eat-4(syb4257) reporter allele also shows expression in the pharyngeal neuron M5, albeit very weakly (Fig. 3A, Table S2), consistent with CeNGEN data. Weak expression of the eat-4 fosmid-based reporter in ASK and ADL remained weak, but clearly detectable with the eat-4(syb4257) reporter allele (Fig. 3A, Table S2). Extremely dim expression in PHA can be occasionally detected. Whereas the PVQ neuron class display eat-4 scRNA transcript and very dim eat-4 fosmid-based reporter expression, we detected no expression of the eat-4(syb4257) reporter allele in PVQ neurons (Fig. 3B, Table S2). We also did not detect expression of the eat-4(syb4257) reporter allele in the GABAergic AVL and DVB neurons, in which a recent report describes expression using an eat-4 promoter fusion (Li et al. 2023). An absence of eat-4(syb4257) expression in AVL and DVB is also consistent with the absence of scRNA transcripts in these neurons.

A few neurons were found to express eat-4 transcripts by the CeNGEN atlas, but only with lower threshold levels, including, for example, the RMD, PVM, and I4 neurons (Fig. 1B, Table S1). We failed to detect reporter allele expression in RMD or PVM neurons, but occasionally observed very dim expression in I4. Lastly, we identified a novel site of eat-4 expression in the dopaminergic PDE neuron (Fig. 3B, Table S2). While such expression was neither detected with previous reporters nor scRNA transcripts, we detected such expression very consistently but relatively weakly.

Expression of a reporter allele of unc-17/VAChT, a marker for cholinergic identity, in the hermaphrodite

41 of previously described 52 neuron classes that show unc-17 fosmid-based reporter expression (Pereira et al. 2015) showed transcripts in the CeNGEN scRNA atlas at 4 out of 4 threshold levels, another 7 neuron classes at 3 out of 4 threshold levels and 1 at the lowest two threshold levels (Taylor et al. 2021). Only one neuron class, RIP, displayed scRNA levels at all 4 thresholds, but showed no corresponding unc-17 fosmid-based reporter expression (Fig. 1B, Table S1). Using the unc-17(syb4491) reporter allele (Fig. 1A), we confirmed expression in RIP (Fig. 3C, Table S2). Of the additional neuron classes that show unc-17 expression only at the least stringent transcript detection level (Fig. 1B, Table S1), we were able to detect unc-17 reporter allele expression only in AWA (Fig. 3C, Table S2).

Conversely, a few neurons display weak expression with previous multicopy, fosmid-based reporter constructs (RIB, AVG, PVN)(Pereira et al. 2015), but show no scRNA support for such expression (Taylor et al. 2021). The unc-17(syb4491) reporter allele confirmed weak but consistent expression in the PVN neurons as well as variable, borderline expression in AVG (Fig. 3C,D). However, we failed to detect unc-17(syb4491) reporter allele expression in the RIB neurons.

We detected a novel site of unc-17 expression, albeit dim, in the glutamatergic AFD neurons (Fig. 3C, Table S2). This expression was not reported with previous fosmid-based reporter or scRNA data. scRNA transcript reads for cha-1/ChAT, the ACh-synthesizing choline acetyltransferase, were also detected in AFD and PVN (Table S1). Taken together, these observations are consistent with the expectation that although available scRNA data capture the majority of gene expression, it may not necessarily have the required depth to detect lowly expressed genes.

Lastly, another notable observation is the lack of any unc-17 reporter expression or scRNA transcripts in the interneuron AVJ, but presence of scRNA transcript reads for cha-1/ChAT (Table S1), which shares exons with the unc-17/VAChT locus (Alfonso et al. 1994). Although no reporter data is available for cha-1/ChAT, such interesting mismatch between available unc-17 and cha-1/ChAT expression data could provide a hint to potential non-vesicular cholinergic transmission in the AVJ neurons in C. elegans, potentially akin to reportedly non-vesicular release of acetylcholine in the visual system of Drosophila (Yang and Kunes 2004).

Expression of reporter alleles for GABAergic pathway genes in the hermaphrodite

Expression of unc-25/GAD

The most recent analysis of GABAergic neurons identified GABA synthesizing cells by anti-GABA staining and an SL2-based unc-25/GAD reporter allele that monitors expression of the rate-limiting step of GABA synthesis, generated by CRISPR/Cas9 engineering (Gendrel et al. 2016). The scRNA atlas shows robust support for these assignments at all 4 threshold levels (Fig. 1B, Table S1). unc-25 scRNA signals were detected at several orders of magnitude lower levels in 3 additional neuron classes, but only with the least robust threshold level.

In this study we generated another unc-25/GAD reporter allele, using a t2a::gfp::h2b cassette (ot1372) (Fig. 2). This allele showed the same expression pattern as the previously described SL2-based unc-25(ot867) reporter allele (Fig. 4A, Table S2). This includes a lack of expression in a number of neurons that stain with anti-GABA antibodies (SMD, AVA, AVB, AVJ, ALA, and AVF) and GLR glia, corroborating the notion that these neurons take up GABA from other cells (indeed, a subset of those cells do express the GABA uptake reporter SNF-11; (Gendrel et al. 2016)).

Expression of GABAergic reporter alleles in the adult hermaphrodite.

(A) Expression of the unc-25/GAD reporter allele ot1372 is detected in the head, ventral nerve cord and tail neurons. The expression pattern of this new T2A-based reporter allele is similar to that of a previously described SL2-based reporter allele, unc-25(ot867)(Gendrel et al. 2016).

(B) Expression of unc-47/VGAT reporter allele syb7566. Left, the expression pattern of the reporter allele largely matches that a previously described unc-47 mCherry fosmid-based reporter (otIs564) in the head. Right, a close-up view for the characterization of the reporter allele expression with landmark strain NeuroPAL (otIs669). In the head, consistent with previous reports of the unc-47 fosmid-based reporter (otIs564), dim expression of unc-47(syb7566) in SMD, ALA, and very dim and variable expression in IL1 is detected in both sexes, and unc-47(syb7566) is expressed in ADF only in the male and not hermaphrodite. In addition, the reporter allele is also expressed at a very dim level in the pharyngeal neuron I1 whereas no expression is detected in M1. In the tail, consistent with previous reports of the fosmid, sexually dimorphic expression of the unc-47(syb7566) reporter allele is also detected in PDB, AS11, PVN, and PHC only in the male and not the hermaphrodite. In addition, we also detected very dim expression of PLM in both sexes, confirming potential dim expression of the unc-47 mCherry fosmid-based reporter that was only readily visible after anti-mCherry staining in the past (Serrano-Saiz et al. 2017b).

(C) Expression of unc-46/LAMP reporter allele syb7278 is largely similar to that of the previously described unc-46 mCherry fosmid-based reporter (otIs568). We also observed expression of both the reporter allele and fosmid-based reporter in PVW, PVN, and very dimly in PDA. Scale bars, 10 μm.

We carefully examined potential expression in the AMsh glia, which were reported to generate GABA through unc-25/GAD (Duan et al. 2020; Fernandez-Abascal et al. 2022). We did not detect visible unc-25(ot867) or unc-25(ot1372) reporter allele expression in AMsh, consistent with the failure to directly detect GABA in AMsh through highly sensitive anti-GABA staining (Gendrel et al. 2016).

Expression of unc-47/VGAT

While promoter-based transgenes for the vesicular transporter for GABA, unc-47/VGAT, had shown expression that precisely match that of unc-25/GAD (Eastman et al. 1999), we had noted in our previous analysis of the GABA system that a fosmid-based reporter showed much broader expression in many additional neuron classes that show no sign of GABA usage (Gendrel et al. 2016). In several of these neuron classes both the fosmid-based reporter and the scRNA data indicate very robust expression (e.g. AIN, SIA, SDQ) while in many others, scRNA transcripts are only evident at looser thresholds and, correspondingly, fosmid-based reporter expression in these cells is often weak (Table S1). To investigate this matter further, we CRISPR/Cas9-engineered a gfp-based reporter allele for unc-47, syb7566, and first crossed it with an mCherry-based unc-47 fosmid-based reporter (otIs564) as a first-pass assessment for any obvious overlaps and mismatch of expression patterns between the two (Fig. 4B, left side panels). The vast majority of neurons exhibited overlapping expression between syb7566 and otIs564. There were also many notable similarities in the robustness of expression of the fosmid-based reporter and the reporter allele (Table S1). In a few cases where the fosmid-based reporter expression was so dim that it is only detectable via antibody staining against its fluorophore (mCherry) (Gendrel et al. 2016; Serrano-Saiz et al. 2017b), the reporter allele expression was readily visible (Table S1).

The only mismatches of expression of the fosmid-based reporter and the reporter allele were in the pharyngeal neuron M1, which expresses no visible unc-47(syb7566) reporter allele but weak fosmid-based reporter, and the pharyngeal neuron I1, which expresses syb7566 but no fosmid-based reporter (Fig. 4B, right side panels). Similarly, AVJ shows very dim and variable unc-47(syb7566) reporter allele expression but no fosmid-based reporter expression. Since AVJ stains with anti-GABA antibodies (Gendrel et al. 2016), this neuron classifies as an uptake and recycling neuron. Other neurons previously shown to stain with anti-GABA antibodies and to express the unc-47 fosmid-based reporter (ALA and SMD)(Gendrel et al. 2016) still show expression of the unc-47 reporter allele. In contrast, the anti-GABA(+) AVA and AVB neurons express no unc-47 and therefore possibly operate solely in GABA clearance.

In conclusion, while the reporter allele of unc-47, in conjunction with scRNA data, corroborates the notion that while UNC-47/VGAT is expressed in all GABA-synthesizing and most GABA uptaking neurons, there is a substantial number of UNC-47(+) neurons that do not show any evidence of GABA presence. This suggests that UNC-47/VGAT may transport another unidentified neurotransmitter (see Discussion).

Expression of unc-46/LAMP

In all GABAergic neurons, the UNC-47/VGAT protein requires the LAMP-like protein UNC-46 for proper localization (Schuske et al. 2007). A previously analyzed fosmid-based reporter confirmed unc-46/LAMP expression in all “classic” GABAergic neurons (i.e. anti-GABA and UNC-25/GAD(+) neurons), but also showed robust expression in GABA- and UNC-47-negative neurons, such as RIA or RMD (Gendrel et al. 2016). This non-GABAergic neuron expression is confirmed by scRNA (Taylor et al. 2021)(Table S1). We generated an unc-46/LAMP reporter allele, syb7278, and found its expression to be largely similar to the fosmid-based reporter and to the scRNA data (Fig. 4C, Table S1), therefore corroborating the non-GABAergic neuron expression of unc-46/LAMP. We also detected previously unreported expression in the PVW and PVN neurons in both the reporter allele and fosmid-based reporter (Fig. 4C), thereby further corroborating CeNGEN data. In addition, we also detected very dim expression in PDA (Fig. 4C), which show no scRNA transcript reads (Table S1). With one exception (pharyngeal M2 neuron class), the sites of non-GABAergic neuron expression of unc-46/LAMP expression do not show any overlap with the sites of unc-47/VGAT expression, indicating that these two proteins have functions independent of one another.

Expression of reporter alleles for serotonin biosynthetic enzymes, tph-1/TPH and bas-1/AAAD, in the hermaphrodite

tph-1/TPH and bas-1/AAAD code for enzymes required for serotonin (5-HT) synthesis (Fig. 1A). scRNA transcripts for tph-1 and bas-1 are detected in previously defined serotonergic neurons at all 4 threshold levels (HSN, NSM, ADF) (Fig. 1, Table S1). In addition to these well characterized sites of expression, several of the individual genes show scRNA-based transcripts in a few additional cells: tph-1 at all 4 threshold levels in AFD and MI. Neither of these cells display scRNA transcripts for bas-1/AAAD, the enzyme that metabolizes the TPH-1 product 5-HTP into 5-HT (Fig. 1A). To further investigate these observations, we generated reporter alleles for both tph-1 and bas-1 (Fig. 2). Expression of the tph-1 reporter allele syb6451 confirmed expression in the previously well described neurons that stained positive for 5-HT, namely NSM, HSN, and ADF, matching CeNGEN data. While expression in AFD (seen at all 4 threshold levels in the scRNA data) could not be confirmed with the reporter allele, expression in the pharyngeal MI neurons could be confirmed (Fig. 5A, Table S2).

Expression of tph-1/TPH, bas-1/AAAD, and cat-1/VMAT reporter alleles in the adult hermaphrodite.

(A) Dorsoventral view of a hermaphrodite head and midbody expressing tph-1(syb6451). tph-1 expression is detected robustly in the MI neuron and dimly and variably in VC4 and VC5. Neuron identities for MI and VC4 & VC5 were validated using otIs518[eat-4(fosmid)::sl2::mCherry::h2b] and vsls269[ida-1::mCherry], respectively, as landmarks.

(B) Neuronal expression of bas-1(syb5923) characterized with the landmark NeuroPAL (otIs669) strain in the head and midbody regions of young adult hermaphrodites. Dorsoventral view of the adult head shows bas-1/AAAD expression in left-right neuron pairs, including previously unreported expression in the URB neurons. Non-neuronal bas-1/AAAD expression is detected in other non-neuronal cell types as reported previously (Yu et al. 2023) (also see Fig. 11, Fig. S4).

(C) Lateral views of young adult hermaphrodite head and midbody expressing cat-1/VMAT (syb6486). Previously unreported cat-1/VMAT expression is seen in RIR, CAN, and very dimly in AUA. Non-neuronal expression of cat-1/VMAT is detected in a single midbody cell in the gonad (also see Fig. S4), marked with an asterisk. Scale bars, 10 μm.

We detected co-expression of the bas-1 reporter allele, syb5923, with tph-1(syb6451) in NSM, HSN, and ADF, in accordance with previous reporter and scRNA data (Fig. 5B, Table S2). However, bas-1(syb5923) is not co-expressed with tph-1 in MI (Fig. 5A,B), nor is there CeNGEN-transcript evidence for bas-1/AAAD in MI (Fig. 1, Table S1). Hence, TPH-1-synthesized 5-HTP in MI is not metabolized into 5-HT, a notion consistent with the lack of 5-HT-antibody staining in MI (Horvitz et al. 1982; Sze et al. 2000).

We also detected tph-1 reporter allele expression in the serotonergic VC4 and VC5 neurons (Fig. 5A, Table S2), consistent with scRNA data (Fig. 1, Table S1) and previous reporter transgene data (Mondal et al. 2018). However, there is no bas-1(syb5923) expression in VC4 or VC5, consistent with previous data showing that serotonin is taken up, but not synthesized by VC4 and VC5 (Duerr et al. 2001).

As expected from the role of bas-1 in dopamine synthesis (Hare and Loer 2004), bas-1(syb5923) is also expressed in dopaminergic neurons PDE, CEP, and ADE. In addition, it is also expressed weakly in URB, consistent with scRNA data. We did not detect visible expression in PVW or PVT, both of which showed very low levels of scRNA transcripts (Fig. 1, Table S1). Expression of bas-1/AAAD in URB may suggest that URB generates a non-canonical monoamine (e.g. tryptamine, phenylethylamine, or histamine), but since URB expresses no vesicular transporter (CAT-1/VMAT, see below), we consider it unlikely that any such monoamine would be secreted via canonical vesicular synaptic release mechanisms.

Expression of a reporter allele of cat-2/TH, a dopaminergic marker, in the hermaphrodite

The CeNGEN scRNA atlas shows transcripts for the rate-limiting enzyme of dopamine-synthesis cat-2/TH (Fig. 1B, Table S1) at all 4 threshold levels in all 3 previously described dopaminergic neuron classes in the hermaphrodite, ADE, PDE, and CEP (Sulston et al. 1975; Sulston et al. 1980; Lints and Emmons 1999). At lower threshold levels, transcripts can also be detected in the OLL neurons. A CRISPR/Cas9-engineered reporter allele for cat-2/TH, syb8255, confirmed expression in ADE, PDE and CEP in adult hermaphrodites (Fig. 6A, Table S2). As expected and described above, all three neuron classes also expressed bas-1/AAAD (Fig. 5B) and cat-1/VMAT (Fig. 5C, see below) (Table S2). We did not detect visible expression of cat-2(syb8255) in the additional neuron class with low levels of scRNA transcript reads, OLL. The OLL neurons also display no scRNA transcripts (nor reporter allele expression) for bas-1/AAAD and cat-1/VMAT. No additional sites of expression of cat-2(syb8255) were detected in the adult hermaphrodite.

Expression of cat-2/TH, tdc-1/TDC, and tbh-1/TBH reporter alleles in the adult hermaphrodite.

Neuronal expression was characterized with landmark strain NeuroPal (otIs669). Lateral views of young adult hermaphrodites expressing reporter alleles for (A) cat-2(syb8255), (B) tbh-1(syb7786), and (C) tdc-1(syb7768). (A) cat-2 expression in CEPs, ADE, and PDE match previously reported expression. (B) and (C) Head areas are shown, no neuronal expression was detected in other areas. tdc-1 expression matches previous reports; we detected previously unreported expression of tbh-1 in all six IL2 neurons at low levels. Scale bars, 10 μm.

Expression of reporter alleles of tdc-1/TDC and tbh-1/TBH, markers for tyraminergic and octopaminergic neurons, in the hermaphrodite

The invertebrate analogs of adrenaline and noradrenaline, tyramine and octopamine, are generated by tdc-1 and tbh-1 (Fig. 1A)(Alkema et al. 2005). Previous work had identified expression of tdc-1 in the hermaphrodite RIM and RIC neurons and tbh-1 in the RIC neurons (Alkema et al. 2005). Transcripts in the CeNGEN atlas match those sites of expression for both tdc-1 (scRNA at 4 threshold levels in RIM and RIC neurons) and tbh-1 (scRNA at 4 threshold levels in RIC neurons) (Fig. 1B, Table S1). Much lower transcript levels are present in a few additional, non-overlapping neurons (Fig. 1B). CRISPR/Cas9-engineered reporter alleles confirmed tdc-1 expression in RIM and RIC and tbh-1 expression in RIC (Fig. 6B,C, Table S2). In addition, we also detected dim expression of tbh-1(syb7786) in all six IL2 neurons, corroborating scRNA transcript data (Fig. 6C, Table S2). However, IL2 neurons do not exhibit expression of tdc-1 (which acts upstream of tbh-1 in the octopamine synthesis pathway) or cat-1/VMAT, the vesicular transporter for octopamine (Fig. 5C, see below). Hence, the IL2 neurons are unlikely to produce or use octopamine as neurotransmitter, but they may synthesize another monoaminergic signal.

Expression of a reporter allele of cat-1/VMAT, a marker for monoaminergic identity, in the hermaphrodite

As the vesicular monoamine transporter, CAT-1/VMAT is expected to be expressed in all neurons that synthesize serotonin, dopamine, tyramine, and octopamine (Fig. 1A). Both scRNA data and a CRISPR/Cas9-engineered reporter allele, syb6486, confirm expression in all these cells (Fig. 5C, Table S2). In addition, based on antibody staining and previous fosmid-based reporters, CAT-1/VMAT is known to be expressed in neurons that may not synthesize but uptake serotonin (VC4/5, RIH, and ADF) (Duerr et al. 1999; Duerr et al. 2001; Serrano-Saiz et al. 2017b). Again, both scRNA data, as well as a CRISPR/Cas9-engineered reporter allele, syb6486, confirm expression in these cells (Fig. 5C, Table S2).

In addition to these canonical monoaminergic neurons, the scRNA data shows cat-1/VMAT expression at all 4 threshold levels in RIR, CAN, AVM and, at much lower threshold, 8 additional neuron classes (Fig. 1B, Table S1). Our cat-1/VMAT reporter allele, syb6486, corroborates expression in RIR and CAN, but not in AVM (Fig. 5C, Table S2). No cat-1(syb6486) expression was detected in neurons that showed cat-1/VMAT transcripts only at the lowest threshold level (Fig. 1B, Table S1). As previously mentioned, we also detected no cat-1/VMAT expression in the tph-1-positive MI or the cat-2/TH-positive OLL neurons.

The cat-1/VMAT reporter allele revealed expression in an additional neuron class, the AUA neuron pair (Fig. 5C, Table S2). Expression in this neuron is not detected in scRNA data, but we believe that such expression may be consistent with previous CAT-1/VMAT antibody staining data (Duerr et al. 1999). These authors found the same expression pattern as we detected with cat-1/VMAT reporter allele, except for the AIM neuron, which Duerr et al. identified as CAT-1/VMAT antibody-staining positive. However, neither our reporter allele, nor a fosmid-based cat-1/VMAT reporter, nor scRNA data showed expression in AIM, and we therefore think that the neurons identified by Duerr et al as AIM may have been the AUA neurons instead. Lastly, a CAT-1-positive neuron pair in the ventral ganglion, unidentified but mentioned by Duerr and colleagues (Duerr et al. 1999), is likely the tyraminergic RIM neuron pair, based on our reporter allele and CeNGEN scRNA data.

Expression of a reporter allele of the betaine uptake transporter SNF-3 may define betainergic neurons

Notably, three CAT-1/VMAT-expressing neuron classes, CAN, RIR, and AUA do not express biosynthetic enzymes for synthesis of the four conventional monoaminergic transmitters known to be employed in C. elegans (5-HT, dopamine, octopamine, and tyramine). It is possible that these three neuron classes may uptake octopamine or tyramine, but the C. elegans genome codes for no conventional octopamine or tyramine uptaker (SLC6A2/3/4 subfamily) other than the bona fide 5-HT and DA uptake transporters mod-5 and dat-1, which are not expressed in these CAT-1/VMAT-positive neurons (Jayanthi et al. 1998; Ranganathan et al. 2001; Jafari et al. 2011). CAT-1/VMAT is also able to package betaine, another putative neurotransmitter (Peden et al. 2013; Hardege et al. 2022). Betaine is synthesized endogenously, within the nervous system mostly in the CAT-1(+) RIM neuron (Hardege et al. 2022), but it is also provided in the bacterial diet of C. elegans (Peden et al. 2013). In vertebrates, dietary betaine is taken up by the betaine transporter BGT1 (aka SLC6A12). To ask whether CAT-1/VMAT-positive neurons may acquire betaine via BGT1-mediated uptake, we engineered a reporter allele for snf-3/BGT1, syb7290, using CRISPR/Cas9 genome engineering. We detected expression in the betaine-synthesizing (and also tyraminergic) RIM neuron (Fig. S1A, Table S1). In addition, we found that snf-3 is indeed expressed in all the three CAT-1/VMAT(+) neuron classes that do not synthesize a previously known monoaminergic transmitter (CAN, AUA, and variably, RIR) (Fig. S1A,B). These cells may therefore take up betaine and synaptically release it via CAT-1/VMAT. The snf-3(syb7290) reporter allele is also expressed in the serotonergic neuron NSM (albeit variably) (Table S1), thus NSM could also be a betaine-uptake neuron. In addition, we also detected snf-3(syb7290) expression in a number of other neurons that do not express CAT-1/VMAT (Table S1, Fig. S1A-C), as well as a substantial number of other non-neuronal types (Fig. S1E-G). These neurons and non-neuronal cells may serve to clear betaine. snf-3(syb7290) is not expressed in the inner and outer labial neuron classes as previously suggested (Peden et al. 2013); these cells were likely misidentified in the previous study and are in fact inner and outer labial glial cells.

Even though all CAT-1/VMAT-positive neurons in the hermaphrodite can be matched with an aminergic neurotransmitter, we nevertheless wondered whether another presently unknown monoaminergic transmitter, e.g., histamine or other trace amine, could be synthesized by a previously uncharacterized AAAD enzyme encoded in the C. elegans genome, hdl-1 (Fig. S2A)(Hare and Loer 2004). We CRISPR/Cas9-engineered an hdl-1 reporter allele, syb1048, but detected no expression of this reporter in the animal (Fig. S2C,D). Attempts to amplify weak expression signals by insertion of Cre recombinase into the locus failed [hdl-1(syb4208)](see Methods). scRNA data also shows no strong transcript expression in the hermaphrodite nervous system and only detected notable expression in sperm (Taylor et al. 2021).

Reporter alleles and NeuroPAL-facilitated neuron class-identification reveal novel expression patterns of neurotransmitters in the male-specific nervous system

No scRNA data is yet available for the nervous system of the male. Based on the expression of fosmid-based reporters, we had previously assembled a neurotransmitter atlas of the C. elegans male nervous system in which individual neuron classes are notoriously difficult to identity (Serrano-Saiz et al. 2017b). We have since established a NeuroPAL landmark strain that permits more reliable identification of gene expression patterns in both the hermaphrodite and male-specific nervous system (Tekieli et al. 2021; Yemini et al. 2021). We used NeuroPAL to facilitate the analysis of the expression profiles of our CRISPR/Cas9-engineered reporter alleles in the male, resulting in updated expression profiles for 9 of the 13 reporter alleles analyzed. As in the hermaphrodite, reasons for the updates vary. In addition to the improved accuracy of neuron identification provided by NeuroPAL, in some cases there are true differences of expression patterns between the fosmid-based reporters and reporter alleles. We elaborate on these updates for individual reporter alleles below.

Expression of reporter alleles of Glu/ACh/GABA markers in the male-specific nervous system

We analyzed eat-4/VGLUT (syb4257), unc-17/VAChT (syb4491), unc-25/GAD (ot1372), and unc-47/VGAT (syb7566) expression in the male-specific nervous system using NeuroPAL landmark strains (otIs696 for eat-4 and otIs669 for all others). Of all those reporter alleles, unc-25/GAD (ot1372) was the only one with no updated expression. Specifically, in addition to confirming presence of expression of the unc-25/GAD reporter allele in CP9, EF1/2, EF3/4, we also confirmed its lack of expression in anti-GABA(+) neurons R2A, R6A, and R9B (Gendrel et al. 2016; Serrano-Saiz et al. 2017b)(Fig. 8A, Table S3).

Expression of eat-4/VGLUT and unc-17/VAChT reporter alleles in the adult male.

Neuronal expression of eat-4(syb4257) and unc-17(syb4491) was characterized with landmark strain NeuroPAL (otIs696 and otIs669, respectively). Only selected neurons are shown to illustrate updates from previous studies. See Table S3 for a complete list of neurons.

(A) eat-4(syb4257) expression. Top, long panels: CA1, CA2, and CA3 show visible, albeit very dim, novel expression of eat-4 (also expressed in CA4). Bottom panels: CA7 strongly expresses eat-4(syb4257), whereas CP7 does not. Neuron IDs for these two neurons were previously switched (Serrano-Saiz et al. 2017b).

(B) unc-17(syb4491) expression. Top, long panels: ventral view of a male ventral nerve cord showing high levels of expression in CA1, CA2, & CA3 and previously unreported low levels of expression in CP1, CP2, and CP3. Middle panels: low levels of expression in CA7 and CP7. There is no visible expression in CP9. Bottom panels: lateral view of a male tail showing previously unreported dim expression in R1B, R4B, R5B, R7B, R9B, and ventral view of the pre-anal ganglion showing expression in DX3/4. Scale bars, 10 μm.

(C) The updated neurotransmitter atlas underscores the molecular diversity of the male-specific ventral cord neuron class CA and CP. Based on their expression patterns for neurotransmitter genes, these neurons can be potentially grouped into the following 4 CA and 5 CP sub-classes. See Table S3 and Figs. 8-10 for all genes mentioned in the following. CAs: (1) CA1 to CA4: express unc-17, unc-47, very weak eat-4; (2) CA5, 6, 8: express unc-17 and unc-47; (3) CA7: expresses unc-17, strong eat-4, and no unc-47; (4) CA9: expresses weak unc-17. CPs: (1) CP0: only expresses weak eat-4; (2) CP1 to CP4: express unc-17, unc-47, cat-1, tph-1, bas-1, and stain for 5-HT; (3) CP5, 6: express eat-4, unc-47, cat-1, tph-1, bas-1, and stain for anti-5-HT; (4) CP7, 8: express very weak unc-17 and unc-47; (5) CP9: expresses strong unc-25 and unc-47, and stain for anti-GABA. The proposed sub-classification for these neuron classes indicates different functions of individual CA and CP neuron types; it also provides useful resources for ongoing single-cell sequencing efforts for male-specific neurons.

Expression of GABAergic reporter alleles in the adult male.

Neuronal expression of unc-25(ot1372) and unc-47(syb7566) reporter alleles was characterized with landmark strain NeuroPAL (otIs669). Only selected neurons are shown to illustrate updates from previous reports. See Table S3 for a complete list of neurons.

(A) unc-25(ot1372) is expressed in male-specific CP9 and EF neurons as well as a few sex-shared neurons, all consistent with previous reports (Gendrel et al. 2016; Serrano-Saiz et al. 2017b).

(B) unc-47(syb7566) shows expression in male head neuron classes MCM and CEM, the former previously undetected and the latter consistent with fosmid-based reporter otIs564.

(C) unc-47(syb7566) shows expression in a number of ventral cord CA and CP neurons, largely consistent with reported otIs564 fosmid-based reporter expression except for no visible expression of syb7566 in CA7 (due to its initial confusion with CP7, described in Fig. 7) and presence of very dim expression in CP7. The syb7566 reporter allele is also not visible in CA9. Scale bars, 10 μm.

In the preanal ganglion, we observed weak expression of unc-17(syb4491) in DX3/4 (Fig. 7B, Table S3), hence assigning previously unknown neurotransmitter identity to these neurons. Related to DX3/4, we also confirmed expression of unc-17 in DX1/2 in the dorsorectal ganglion, consistent with fosmid-based reporter data (Table S3) (Serrano-Saiz et al. 2017b). In the lumbar ganglion, we detected novel expression of unc-17(syb4491) in 5 pairs of type B ray neurons, namely R1B, R4B, R5B, R7B, and R9B (Fig. 7B, Table S3). Expression in all these neurons is quite dim, with R7B and R9B being the dimmest (Fig. 7B). The low levels of expression may explain why it was not captured earlier by the unc-17 fosmid-based reporter (Serrano-Saiz et al. 2017b).

In the ventral nerve cord, we found additional, very weak expression of eat-4(syb4257) in CA1 to CA4 (Fig. 7A, Table S3), as well as weak expression of unc-17(syb4491) in CP1 to CP4 (Fig. 7B, Table S3), all undetected by previous analysis of fosmid-based reporters (Serrano-Saiz et al. 2017b). Conversely, two neurons lack previously reported expression of fosmid-based reporters; CP9 does not show visible unc-17(syb4491) expression (Fig. 7B) and neither does CA9 show visible expression of unc-47(syb7566) expression (Fig. 8C). We also realized that the neuron identifications of CA7 and CP7 were previously switched (Serrano-Saiz et al. 2017b), due to lack of proper markers for those two neurons. With NeuroPAL, we are now able to clearly distinguish the two and update their classic neurotransmitter reporter expression: CA7 expresses high levels of eat-4(syb4257) (Fig. 7A, Table S3), very low levels of unc-17(syb4491) (Fig. 7B), and no unc-47(syb7566) (Fig. 8C); CP7 expresses no eat-4(syb4257) (Fig. 7A, Table S3), very low levels of unc-17(syb4491) (Fig. 7B), and very low levels of unc-47(syb7566) as well (Fig. 8C). Taken together, the analysis of reporter alleles reveals a remarkable diversity of CA and CP neurons, summarized in Fig. 7C.

In the head, we detected expression of unc-47(syb7566) in the male-specific neuron class MCM (Fig. 8B, Table S3), previously not observed with fosmid-based reporters. Consistent with fosmid-based reporter data, the other male-specific neuron class, CEM, shows expression of unc-17(syb4491) (Table S3) and unc-47(syb7566) (Fig. 8B, Table S3) reporter alleles.

Expression of reporter alleles for monoaminergic neurotransmitter pathway genes in the male-specific nervous system

We analyzed cat-1/VMAT (syb6486), tph-1/TPH (syb6451), cat-2/TH (syb8255), bas-1/AAAD (syb5923), tdc-1/TDC (syb7768), and tbh-1/TBH (syb7786) reporter alleles expression in the male-specific nervous system with the help of the NeuroPAL reporter landmark (otIs669). We found novel expression patterns in neurons in all male-specific ganglia (Fig. 9, Fig. 10, Table S3).

Expression of the cat-1/VMAT, tph-1/TPH, and bas-1/AAAD reporter alleles in the adult male. Neuronal expression was characterized with landmark strain NeuroPAL (otIs669).

(A) Novel cat-1(syb6486) expression is seen in male-specific neurons PDC, PVY, PVX, R2A, and R4B. Consistent with previous reports, it is also expressed in HOA, PGA, R5A, R7A, R9A, R1B, and R8B. Its expression in ventral cord neurons CP1 to CP6 is consistent with earlier studies.

(B) tph1-(syb6451) is expressed in male-specific head neuron class CEM and sex-shared neurons ADF, NSM, and MI. Similar to its expression in hermaphrodites, tph-1 in MI was previously undetected. In the tail, in addition to previously determined expression in R1B, R3B, and R9B, tph-1(syb6451) is also expressed at very low levels in R4B and R7B. Ventral cord expression of tph-1(syb6451) in CP1 to CP6 is consistent with previous reports and thus not shown here.

(C) bas-1(syb5923) is expressed in previously identified NSM, ADE, PDE, and CEP neurons. In addition, we detected weak expression in URB as in the hermaphrodite. We also updated bas-1/AAAD expression in 22 male-specific neuron types and 1 more with variable expression (see Table S3 for complete list). Scale bars, 10 μm. Asterisks, non-neuronal expression, also see Fig. 11 and Fig. S4.

Expression of cat-2/TH, tdc-1/TDC and tbh-1/TBH reporter alleles in the adult male.

Neuronal expression was characterized with landmark strain NeuroPAL (otIs669).

(A) cat-2(syb8255) is expressed in male-specific neurons R4A, R7A, and R9B. This expression, as well as its expression in sex-shared neurons PDE, CEP, and ADE is consistent with previous reports (Sulston et al. 1975; Sulston et al. 1980; Lints and Emmons 1999).

(B) tdc-1(syb7768) is expressed in sex-shared neurons RIM and RIC and male-specific neurons HOA, R8A, and R8B, all consistent with previous studies (Serrano-Saiz et al. 2017b). We also detected weak expression in R7A.

(C) tbh-1(syb7786) is expressed in RIC, consistent with its previously reported expression in hermaphrodites. As in hermaphrodites, we also detected tbh-1(syb7786) in IL2 neurons of the male. In male-specific neurons, previously unreported expression is detected in CEM, HOB, and all type B ray neurons except for R6B. Intriguingly, this expression pattern resembles that of pkd-2 and lov-1, both genes essential for male mating functions (Barr and Sternberg 1999; Barr et al. 2001). Scale bars, 10 μm. Asterisks, non-neuronal expression, also see Fig. 11 and Fig. S4.

Serotonin

Serotonergic identity had been assigned to several male-specific neurons before (CP1 to CP6, R1B, R3B, R9B) and we validated these assignments with our reporter alleles (Fig. 9, Table S3). In addition, we detected previously unreported expression of tph-1 (Fig. 9B) in the male-specific head neuron class CEM, as well as in a subset of B-type ray sensory neurons, R4B and R7B. However, not all of these neurons display additional, canonical serotonergic neuron features: While R4B and R7B express bas-1(syb5923) (with R4B expressing it variably) to generate 5-HT, only CEM displays, albeit weak, 5-HT antibody staining (Serrano-Saiz et al. 2017b). Only R4B expresses cat-1(syb6486) for vesicular release of 5-HT. Conversely, a male-specific neuron in the preanal ganglion, PGA, stains for anti-5-HT (Serrano-Saiz et al. 2017b) but does not express tph-1(syb6451) or bas-1(syb5923) reporter alleles (Table S3). The absence of tph-1/TPH and bas-1/AAAD expression confirms the previous assignment of PGA as a 5-HT uptake neuron. PGA expresses the cat-1/VMAT reporter allele, confirming that PGA can package serotonin into vesicles, therefore constituting an uptake-dependent serotonergic neuron, similar to the sex-shared RIH neuron (Table 1, Table S2).

We had previously reported that the PVW neuron stains with anti-serotonin antibodies exclusively in males but did not detect expression of a fosmid-based reporter for the serotonin-synthesizing enzyme TPH-1 (Serrano-Saiz et al. 2017b). We confirmed the lack of tph-1 expression with our new tph-1 reporter allele in both males and hermaphrodites, and also found that hermaphrodite and male PVW does not express the reporter allele for the other enzyme in the 5-HT synthesis pathway, bas-1. Because of very dim cat-1::mCherry fosmid-based reporter expression, only detected upon anti-mCherry antibody staining, we had assigned PVW as a 5-HT-releasing neuron (Serrano-Saiz et al. 2017b). However, we failed to detect expression of our new cat-1/VMAT reporter allele and hence assign PVW a 5-HT-uptake/clearance function in the male, akin to the AIM, URX, and I5 neurons (Table 1, Table S2).

In the ventral nerve cord, in addition to the expression of cat-1(syb6486) and tph-1(syb6451) in CP1 to CP6, consistent with previous fosmid-based reporter data (Fig. 9A,B; Table S3), we also detected novel expression of bas-1(syb5923) in CP1 to CP4 and strongly in CP5 and CP6 (Fig. 9C, Table S3). This updated expression supports the serotonergic identities of these neurons, which had been determined previously based only on their expression of cat-1/VMAT and positive staining for 5-HT.

Dopamine

We found that the expression of the dopamine-synthesizing cat-2(syb8255) reporter allele precisely matched previous assignments of dopaminergic identity (Sulston et al. 1975; Sulston et al. 1980; Lints and Emmons 1999), i.e. expression was detected exclusively in R5A, R7A, and R9A (Fig. 10A, Table S3), in addition to all sex-shared dopaminergic neurons. All these neurons show matching expression of bas-1/AAAD, the other essential enzyme for dopamine synthesis, and cat-1/VMAT, the vesicular transporter for dopamine (Fig. 9A,C; Table S3).

Tyramine & Octopamine

Reporter alleles for the two diagnostic enzymes, tdc-1/TDC and tbh-1/TBH, confirm the previously reported assignment of HOA as tyraminergic (Serrano-Saiz et al. 2017b), based on the presence of tdc-1(syb7768) expression, but absence of expression of the tbh-1(syb7786) reporter allele (Fig. 10B,C). The tdc-1 reporter allele reveals a novel site of expression in R7A. Due to lack of tbh-1 expression, R7A therefore classifies as another tyraminergic neuron. Both HOA and R7A also co-express cat-1/VMAT for vesicular release of tyramine.

We detected no neurons in addition to the sex-shared RIC neuron class that shares all features of a functional octopaminergic neuron, i.e. co-expression of tbh-1/TBH, tdc-1/TDC, and cat-1/VMAT. While one male-specific neuron, R8B, shows an overlap of expression of tdc-1(syb7768) and tbh-1(syb7786), indicating that these neurons can synthesize octopamine, R8B does not express cat-1(syb6486), indicating that these neurons cannot engage in vesicular release of octopamine.

Curiously, while there are no other male-specific neurons that co-express tdc-1 and tbh-1, there are several male-specific neurons that express tbh-1, but not tdc-1 (Fig. 10B,C; Table 1, Table S3). The absence of expression of the TDC-1/AAAD protein, which produces tyramine, the canonical substrate of the TBH-1 enzyme, indicates that TBH-1 must be involved in the synthesis of a compound other than octopamine. Moreover, bas-1/AAAD is expressed in several of the TBH-1(+); TDC-1(-) neurons (R1B, R2B, R3B, R4B, and R7B) (Fig. 9C, Table 1, Table S3). Rather than using L-Dopa or 5-HTP as substrate, BAS-1/AAAD may decarboxylate aromatic amino acids, which then may serve as a substrate for TBH-1. We consider the trace amine phenylethanolamine (PEOH) as a candidate end product (see Discussion).

Other monoaminergic neurons

In the preanal ganglion, we detected novel expression of cat-1(syb6486) in the cholinergic PDC, PVX, and PVY neurons (Fig. 9A). Intriguingly, these neurons express no other serotonergic, dopaminergic, tyraminergic, or octopaminergic pathway gene, but we did find PDC (but not PVX or PVY) to express the betaine uptake transporter reporter allele snf-3(syb7290) (Fig. S1C). This indicates that PDC, just like the sex-shared neurons CAN, AUA, and RIR neuron classes, could potentially take up and synaptically release betaine. PVX and PVY may synthesize or uptake another aminergic transmitter. Such presumptive transmitter is not likely to be synthesized by hdl-1/AAAD since we detected no expression of the hdl-1 reporter allele syb4208 in the male nervous system (Fig. S2C,D).

In addition to PDC, we also observed snf-3 expression in MCM, PHD and variably, in PVV (Fig. S1C). Since these neurons do not express cat-1/VMAT they likely constitute betaine clearance neurons.

The expression pattern of the bas-1/AAAD, which had not been previously analyzed in the male-specific nervous system, reveals additional novelties. In addition to the “canonical” serotonergic and dopaminergic neurons described above, we detected bas-1 reporter allele expression in a substantial number of additional neurons, including the tyraminergic HOA and R7A neurons, but also the DVE, DVF, R2A, R3A, R6A, R8A, R2B, R6B, R7B, PCB and SPC neurons (Fig. 9C, Table S3). As described above, a subset of the neurons co-express TBH-1 (most B-type ray neurons), a few co-express TDC-1 (HOA and several A-type ray neurons), and several co-express neither of these two genes. Only a subset of these neurons express CAT-1/VMAT. Taken together, this expression pattern analysis argues for the existence of additional monoaminergic signaling system(s)(Table 1).

Sexually dimorphic neurotransmitter expression in sex-shared neurons

eat-4/VGLUT

We had previously noted that a fosmid-based eat-4/VGLUT reporter is upregulated in the sex-shared neuron PVN, specifically in males (Serrano-Saiz et al. 2017b). Since PVN is also cholinergic (Fig. 3D)(Pereira et al. 2015), this observation indicates a sexually dimorphic co-transmission configuration. As described above (Fig. 3B, Table S2), our eat-4 reporter allele revealed low levels of eat-4/VGLUT expression in hermaphrodites PVN, but in males the eat-4 reporter alleles shows strongly increased expression, compared to hermaphrodites. Hence, rather than being an “on” vs. “off” dimorphism, dimorphic eat-4/VGLUT expression in male PVN resembles the “scaling” phenomenon we had described previously for eat-4/VGLUT in male PHC neurons, compared to hermaphrodite PHC neurons (Serrano-Saiz et al. 2017a). Both PHC and PVN display a substantial increase in the number of synaptic output of these neurons in males compared to hermaphrodites (Cook et al. 2019), providing a likely explanation for such scaling of gene expression. The scaling of eat-4/VGLUT expression in PVN is not accompanied by scaling of unc-17/VAChT expression, which remains comparable in both sexes (Fig. 3D).

We also examined AIM, another neuron class that was previously reported to be sexually dimorphic in that AIM expresses eat-4/VGLUT fosmid-based reporters in juvenile stages in both sexes, whereas upon sexual maturation its neurotransmitter identity is switched from being glutamatergic to cholinergic only in adult males and not hermaphrodites (Pereira et al. 2015; Pereira et al. 2019). With the eat-4(syb4257) reporter allele, we also detected a downregulation of eat-4 expression to low levels in young adult males and almost complete elimination in 2-day-old adult males, while expression in hermaphrodites stays high.

unc-17/VAChT

The unc-17/VAChT reporter allele syb4491 confirms that cholinergic identity is indeed male-specifically turned on in the AIM neurons (Fig. 3C), thereby confirming the previously reported neurotransmitter switch (Pereira et al. 2015). The fosmid-based unc-17 reporter also showed sexually dimorphic expression in the AVG neurons (Serrano-Saiz et al. 2017b). This is also confirmed with the unc-17 reporter allele, which shows slightly stronger, albeit still dim, AVG expression in males compared to hermaphrodites (Fig. 3C, showing a hermaphrodite representing animals with no visible expression and a male with representative dim expression).

unc-47/VGAT

unc-47(syb7566) confirms previously reported sexually dimorphic expression of unc-47/VGAT in several sex-shared neurons, including ADF, PDB, PVN, PHC, AS10, and AS11 (Fig. 4B, right side panels) (Serrano-Saiz et al. 2017b). The assignment of AS10 was not definitive in our last report (we had considered either DA7 or AS10), but with the help of NeuroPAL the AS10 assignment could be clarified. In all these cases expression was only detected in males and not hermaphrodites. It is worth mentioning that expression of the mCherry-based unc-47/VGAT fosmid-based reporter (otIs564) in some of these neurons was so dim that it could only be detected through immunostaining against the mCherry fluorophore and not readily visible with the fosmid-based reporter by itself (Serrano-Saiz et al. 2017b). In contrast, the unc-47/VGAT reporter allele is detected in all cases except the PQR neuron class. In addition, we also detected dim unc-47/VGAT expression in the PLM neurons in both sexes (Fig. 4B).

In conclusion, although there are some updates in the levels of dimorphic gene expression (PVN neuron class), our analysis with reporter alleles does not reveal novel sexual dimorphism in sex-shared neurons compared to those that we previously identified in (Serrano-Saiz et al. 2017b). These sexual dimorphisms are summarized in Table S4.

Neurotransmitter pathway genes in glia

In vertebrates, glia can produce various signaling molecules, including neurotransmitters (Araque et al. 2014; Savtchouk and Volterra 2018). There is some limited evidence for neurotransmitter synthesis in C. elegans glia. In males, it had been reported that the socket glia of spicule neurons synthesize and utilize dopamine, based on their expression of cat-2/TH and bas-1/AAAD (Lints and Emmons 1999; Hare and Loer 2004; Leboeuf et al. 2014). We confirmed this notion with cat-2/TH and bas-1 reporter alleles (Fig. 11A). Additionally, we detected expression of cat-1/VMAT reporter allele expression in these cells (Fig. 11A), indicating that these glia secrete dopamine by canonical vesicular transport. We also observed bas-1(syb5923) reporter allele expression in cells that are likely to be the spicule sheath glia (Fig. 11A), as well as in additional glia cell types in the head and tail (Fig. 11B).

Expression of neurotransmitter pathway genes in non-neuronal cell types.

Multiple neurotransmitter pathway genes show expression in glial cells (A,B) and other non-neuronal cell types (C). Also see Fig. S4 for whole-worm views that capture more non-neuronal expression.

(A) bas-1, cat-2, and cat-1 reporter alleles exhibit expression in the male spicule glial cell types, largely consistent with previous reports.

(B) top 6 panels: bas-1(syb5923) is expressed in additional, multiple glial cell types in the male tail. Left 3 panels: bas-1(syb5923) crossed into a mir-228::tagRFP pan-glial marker, confirming its expression in glial cells; right 3 panels: bas-1(syb5923) shows no overlap with the panneuronal marker component in NeuroPAL (otIs669). Bottom 2 panels: bas-1(syb5923) also shows expression in at least two glial cells in the head. A hermaphrodite head is shown here. Expression is similar in the male.

(C) cat-1, tdc-1, and tbh-1 are expressed in the male somatic gonad. All three have expression in the vas deferens; additionally, cat-1 and tbh-1 are also expressed in the seminal vesicle.

(D) We also detected eat-4(syb4257) expression in muscle cells in both sexes, most prominently in the head.

We detected no expression of other vesicular transporters or neurotransmitter biosynthetic synthesis machinery in glia of either sex. This observation contrasts previous reports on GABA synthesis and release from the AMsh glia cell type (Duan et al. 2020; Fernandez-Abascal et al. 2022). We were not able to detect signals in AMsh with anti-GABA staining, nor with an SL2 or T2A-based GFP-based reporter allele for unc-25. (Gendrel et al. 2016)(M. Gendrel, pers. comm.; this paper).

There is, however, evidence for neurotransmitter uptake by C. elegans glial cells, mirroring this specific function of vertebrate glia (Henn and Hamberger 1971). We had previously shown that one specific glia-like cell type in C. elegans, the GLR glia, take up GABA via the GABA uptake transporter SNF-11 (Gendrel et al. 2016). We did not detect unc-47/VGAT fosmid-based reporter expression in the GLRs (Gendrel et al. 2016) and also detected no expression with our unc-47/VGAT reporter allele. Hence, these glia are unlikely to release GABA via classic vesicular machinery. Other release mechanisms for GABA can of course not be excluded. Aside from SNF-11 expression in GLR glia, we also observed broad glia expression of the betaine uptake transporter SNF-3/BGT1 in glia in the head, midbody, and tail, of both sexes, indicating a betaine clearance role for glia (Fig. S1E,F).

Neurotransmitter pathway gene expression outside the nervous system

We detected expression of a few neurotransmitter pathway genes in cells outside the nervous system. The most prominent sites of reporter allele expression are located within the somatic gonad. We detected co-expression of tdc-1(syb7768) and tbh-1(syb7786) reporter alleles in the gonadal sheath of hermaphrodite as well as tdc-1(syb7768) expression around the vulva, likely in the neuroendocrine uv1 cells (Fig. S4), as previously reported (Alkema et al. 2005). However, we observed no cat-1(syb6486) expression in the gonadal sheath or the uv1 cells, suggesting alternative release mechanisms for tyramine and octopamine. Instead, cat-1(syb6486) is expressed in a midbody gonadal cell posterior to the vulva, likely the distal valve (Fig. 5C, Fig. S4).

In the male, tdc-1(syb7768), tbh-1(syb7786), and cat-1(syb6486) are also expressed in the somatic gonad: while all three genes are co-expressed in the vas deferens, cat-1 and tbh-1, but not tdc-1, are co-expressed in the seminal vesicle (Fig. 11C). A similar pattern of CAT-1(+); TBH-1(+); TDC-1(-) is detected in several male-specific neurons and may indicate the usage of a novel transmitter (e.g. PEOH, see Discussion) by these cells. SNF-3/BGT1 is also expressed in male somatic gonad cells, indicating that these cells could also use betaine for signaling (Fig. S1E).

The AAADs tdc-1, as well as bas-1, are also prominently expressed in the intestine, where bas-1 has been shown to be involved in generating 5-HT-derived glucosides (Yu et al. 2023). bas-1, but not tdc-1, is also expressed in the hypodermis and seam cells, as is the betaine uptake transporter snf-3 (Fig. S1G, S4). The tph-1 reporter allele expresses in a subset of pharyngeal non-neuronal cells during the L1 to L4 larval stages of development (Fig. S3), which is consistent with low levels of tph-1 transcripts detected in pharyngeal muscles in the CeNGEN scRNA dataset. Additionally, we observed previously uncharacterized eat-4/VGLUT expression in muscle cells in both sexes (Fig. 11D).

Discussion

Using CRISPR/Cas9-engineered reporter alleles we have refined and extended neurotransmitter assignment throughout the nervous system of C. elegans. We conclude that in both hermaphrodites and males, about one quarter of neurons are glutamatergic (eat-4/VGLUT-positive), a little more than half are cholinergic (unc-17/VAChT-positive), around 10% are GABAergic (unc-25/GAD-positive), and about another 10% are monoaminergic (cat-1/VMAT-positive). Comprehensive lists for gene expression and identity assignment are provided in Table S2 for hermaphrodites and Table S3 for males. We summarize updates compared to previously reported data and our rationale for assigning neurotransmitter identities in Table 1 and Table S5.

Neurotransmitter synthesis versus uptake

Direct detection of neurotransmitters through antibody staining has shown that at least two neurotransmitters, GABA and 5-HT, are present in some neurons that do not express the synthesis machinery for these transmitters (Table 2). Instead, these neurons acquire GABA and 5-HT through uptaking them via defined uptake transporters, SNF-11 for GABA (Mullen et al. 2006) and MOD-5 for 5-HT (Ranganathan et al. 2001; Jafari et al. 2011). A combination of scRNA transcriptome and our reporter allele data corroborates the absence of synthesis machinery in these presumptive uptake neurons (Table 2). One interesting question that relates to these uptake neurons is whether they serve as “sinks” for clearance of a neurotransmitter or whether the taken-up neurotransmitter is subsequently “recycled” for synaptic release via a vesicular transporter. Previous data, as well as our updated expression profiles, provide evidence for both scenarios: ALA and AVF do not synthesize GABA via UNC-25/GAD, but stain with anti-GABA antibodies in a manner that is dependent on the uptake transporter SNF-11 (Gendrel et al. 2016). ALA expresses both UNC-47 and UNC-46, hence it is likely to synaptically release GABA, but AVF expresses neither and is therefore apparently involved only in GABA clearance. Similarly, RIH and AIM both stain for 5-HT in a MOD-5-uptake transporter-dependent manner (Jafari et al. 2011), but only RIH, not AIM, expresses the vesicular transporter CAT-1/VMAT, making RIH a serotonergic signaling neuron and AIM a clearance neuron.

Some neurons do not obviously fall into the synthesis or uptake category, most notably, the anti-GABA-antibody-positive SMD, AVA, and AVB neurons (all of which conventional cholinergic neurons). None of these neurons express UNC-25/GAD, nor the SNF-11 uptake transporter, yet unc-25/GAD is required for their anti-GABA-positive staining (Gendrel et al. 2016). This suggests that GABA may be acquired by these neurons through non-canonical uptake mechanisms. Also, unlike the SMD neurons, the AVA and AVB neurons do not express UNC-47 (Gendrel et al. 2016; Taylor et al. 2021)(this paper); hence, it is not clear if or how GABA is released from them. A member of the bestrophin family of ion channels has been shown to mediate GABA release from astrocyte glia in vertebrates (Lee et al. 2010); however, while there are more than 20 bestrophin channels encoded in the C. elegans genome (Hobert 2013), they do not appear to be expressed in the AVA or AVB neurons (Taylor et al. 2021).

The co-expression of a specific uptake transporter and a vesicular transporter also leads us to predict the usage of betaine as a potential neurotransmitter. Betaine is known to be synthesized in C. elegans, but is also taken up via its diet (Peden et al. 2013; Hardege et al. 2022). Betaine has documented effects on animal behavior and acts via activation of several betaine-gated ion channels (Peden et al. 2013; Hardege et al. 2022). Expression of biosynthetic enzymes suggests betaine production in at least the RIM neuron class, which also expresses the vesicular transporter CAT-1/VMAT, capable of transporting betaine (Hardege et al. 2022). The expression of the betaine uptake transporter SNF-3/BGT1 in CAN, AUA, RIR, and male-specific neuron PDC, coupled with their co-expression of CAT-1/VMAT, suggests that several distinct neuron classes in different parts of the nervous system may uptake betaine and engage in vesicular betaine release via CAT-1/VMAT to gate betaine-activated ion channels, such as ACR-23 (Peden et al. 2013) or LGC-41 (Hardege et al. 2022).

Evidence for usage of currently unknown neurotransmitters

Novel inhibitory transmitters?

unc-47/VGAT is expressed in a substantial number of non-GABAergic neurons (95 out of 302 total neurons in hermaphrodites, plus 61 out of 93 male-specific neurons). However, expression in many of these non-GABAergic neurons is low and variable and such expression may not lead to sufficient amounts of a functional gene product. Yet, in some neurons (e.g. the SIA neurons) expression of unc-47 is easily detectable and robust (based on fosmid-based reporter, reporter allele, and scRNA data), indicating that VGAT may transport another neurotransmitter. In vertebrates, VGAT transports both GABA and glycine, and the same is observed for UNC-47 in vitro (Aubrey et al. 2007). While the C. elegans genome encodes no easily recognizable ortholog of known ionotropic glycine receptors, it does encode closely related glutamate-gated anion channels (Hobert 2013). Moreover, a recently identified metabotropic glycine receptor, GPR158 (Laboute et al. 2023), has a clear sequence ortholog in C. elegans, F39B2.8. Therefore, glycine may also act as a neurotransmitter in C. elegans. VGAT has also been shown to transport ý-alanine (Juge et al. 2013), another potential, but as yet unexplored, neurotransmitter in C. elegans. However, it needs to be pointed out that most of the additional UNC-47(+) neurons do not co-express the LAMP-type UNC-46 protein, which is important for sorting UNC-47/VGAT to synaptic vesicles in conventional GABAergic neurons (Schuske et al. 2007). In vertebrates, the functional UNC-46 ortholog LAMP5 is only expressed and required for VGAT transport in a subset of VGAT-positive, GABAergic neurons (Tiveron et al. 2016; Koebis et al. 2019), indicating that alternative vesicular sorting mechanisms may exist for UNC-47/VGAT.

Novel monoaminergic transmitters?

Two neuron classes in the male-specific nervous system, PVX and PVY, express cat-1/VMAT but do not express the synthesis machinery for 5-HT, tyramine, octopamine, or dopamine. Neither do they show evidence for uptake of known monoamines. There are also several CAT-1/VMAT-positive male-specific neurons that express only a subset of the biosynthetic machinery involved in the biosynthesis of known aminergic transmitters in the worm. That is, some neurons express CAT-1/VMAT and BAS-1/AAAD, but none of the previously known enzymes that produce the substrate for BAS-1, i.e. CAT-2 or TPH-1 (Fig. 1A). In these neurons, BAS-1/AAAD may decarboxylate an unmodified (i.e. non-hydroxylated) aromatic amino acid as substrate to produce, for example, the trace amine phenylethylamine (PEA) from phenylalanine (Table 1, Fig. S2A). A subset of these neurons (all being B-type ray sensory neurons) co-express TBH-1, which may use PEA as a substrate to produce the trace amine, phenylethanolamine (PEOH). PEOH is a purported neurotransmitter in Aplysia (Saavedra et al. 1977) and the vertebrate brain (Saavedra and Axelrod 1973) and can indeed be detected in C. elegans extracts (F. Schroeder, pers. comm.). bas-1/AAAD may also be responsible for the synthesis of histamine, an aminergic neurotransmitter that can be found in extracts of C. elegans (Pertel and Wilson 1974). The only other AAAD that displays reasonable sequence similarity to neurotransmitter-producing AAADs is the hdl-1 gene (Hare and Loer 2004; Hobert 2013)(Fig. S2B), for which we, however, did not detect any expression in the C. elegans nervous system (Fig. S2D). Lastly, based on the expression of tph-1, but concurrent absence of bas-1/AAAD, the pharyngeal MI neuron may produce 5-HTP (Fig. S2A). 5-HTP may either be used directly as a signaling molecule or it may be metabolized into some other serotonin derivative, an interesting possibility in light of serotonin-derivatives produced elsewhere in the body (Yu et al. 2023).

Taken together, monoaminergic pathway genes are expressed in non-conventional combinations in several neuron classes, indicating the existence of monoaminergic signaling systems that have yet to be uncovered.

Neurons devoid of canonical neurotransmitter pathway genes may define neuropeptide-only neurons

We identified neurons that do not express any conventional, well-characterized vesicular neurotransmitter transporter families, namely UNC-17/VAChT, CAT-1/VMAT (the only SLC18 family members), UNC-47/VGAT (only SLC32 family member), or EAT-4/VGLUT (an SLC17 family member). Eight sex-shared neurons (ASI, AVF, AVH, BDU, PVM, PVQ, PVW, RMG) and one male-specific neuron (SPD) fall into this category. Most of these neurons exhibit features that are consistent with them being entirely neuropeptidergic. First, electron microscopy has revealed a relative paucity of clear synaptic vesicles in most of these neurons (White et al. 1986; Cook et al. 2019; Witvliet et al. 2021). Second, not only do these neurons express a multitude of neuropeptide-encoding genes (Taylor et al. 2021), but they also display a dense interconnectivity in the “wireless” neuropeptidergic connectome (Ripoll-Sanchez et al. 2022). In addition, one of these neurons, ASI, stands out for its expression of more than a dozen insulin-like peptides (Taylor et al. 2021), indicating a primarily neuroendocrine function of this neuron.

That said, electron microscopy shows that some of the neurons devoid of conventional neurotransmitter pathway genes generate synapses with small, clear synaptic vesicles, indicative of the use of non-peptidergic transmitters (e.g. the sex-shared RMG and PVM neurons or the male-specific SPD neurons) (White et al. 1986; Cook et al. 2019; Witvliet et al. 2021). It is therefore conceivable that either conventional neurotransmitters utilize non-conventional neurotransmitter synthesis and/or release pathways, or that completely novel neurotransmitter systems remain to be discovered. Although the C. elegans genome does not encode additional members of the SLC18A2/3 (cat-1/VMAT, unc-17/VAChT) or SLC32A1 (unc-47/VGAT) family of vesicular neurotransmitter transporters, it does contain a number of additional members of the SLC17A6/7/8 (VLGUT) family (Hobert 2013). These may serve as non-canonical vesicular transporters of more uncommon neurotransmitters or, alternatively, may be involved in modulating release of glutamate (Serrano-Saiz et al. 2020; Choi et al. 2021). Uncharacterized paralogs of bona fide neurotransmitter uptake transporters (SLC6 superfamily) may also have functions in neurotransmitter release rather than uptake. However, based on CeNGEN scRNA data, no robust or selective expression of these SLC17 or SLC6 family members is observed in these “orphan neurons”.

Co-transmission of multiple neurotransmitters

Our analysis expands the repertoire of neurons that co-transmit multiple neurotransmitters (Fig. 12). Neurotransmitter co-transmission has been observed in multiple combinations in the vertebrate brain (Wallace and Sabatini 2023). In C. elegans, the most frequent co-transmission configurations are a classic, fast transmitter (acetylcholine or glutamate) with a monoamine. Co-transmission of two distinct monoaminergic systems also exists. In several cases it is, however, not clear whether the second neurotransmitter is indeed used for communication or whether its presence is merely a reflection of this neuron being solely an uptake neuron. For example, the glutamatergic AIM neuron stains positive for serotonin, which it uptakes via the uptake transporter MOD-5, but it does not express the vesicular monoamine transporter CAT-1/VMAT (Fig. 12, Table 1, Table 2).

Neurons that co-transmit multiple neurotransmitters.

ACh = acetylcholine Glu = glutamate; DA = dopamine; TA? = bas-1-dependent trace amine (histamine, tryptamine, PEA; also see Fig. S2); 5-HT = 5-hydroxytryptamine, or serotonin; 5-HTP = 5-hydroxytryptophan; PEOH = β-hydroxyphenethylamine, or phenylethanolamine; * Variable expression of respective identity gene detected. This could be due to one of the following reasons: (1) the gene is expressed in some but not all animals; (2) the gene is expressed in every animal but the level of expression is below detection threshold in some. In the case of “GABA*” for SMD and “5-HT*” for VC4/5, the anti-GABA and anti-5-HT staining, respectively, was variable (Serrano-Saiz et al. 2017b). ** indicates that R4B and R7B express 5-HT-synthesis machinery (tph-1 and bas-1), but do not stain with 5-HT antibodies.

Co-transmission of small, fast-acting neurotransmitters (glutamate, GABA, acetylcholine) does exist, but it is rare (Fig. 12). The most prominent co-transmission configuration is acetylcholine with glutamate, but acetylcholine can also be co-transmitted with GABA. There are no examples of co-transmission of glutamate and GABA, as observed in several regions of the vertebrate brain (Wallace and Sabatini 2023).

Interestingly, co-transmission appears to be much more prevalent in the male-specific nervous system, compared to the sex-shared nervous system (Fig. 12, Table S3). Remarkably, several male-specific neuron classes may utilize three co-transmitters. Such extensive co-transmission may relate to male-specific neurons displaying a greater degree of anatomical complexity compared to the hermaphrodite nervous system, both in terms of branching patterns and extent of synaptic connectivity (Jarrell et al. 2012; Cook et al. 2019). Given that all co-transmitting neurons display multiple synaptic outputs (Cook et al. 2019), it appears possible that each individual neurotransmitter secretory system is distributed to distinct synapses. Based on vertebrate precedent (Wallace and Sabatini 2023), co-release from the same vesicle is also possible.

Sexual dimorphisms in neurotransmitter usage

The observation of sexual dimorphisms in neurotransmitter abundance in specific regions of the mammalian brain has been one of the earliest molecular descriptors of neuronal sex differences in mammals (Mccarthy et al. 1997). However, it has remained unclear whether such differences are the result of the presence of sex-specific neurons or are indications of distinctive neurotransmitter usage in sex-shared neurons. Using C. elegans as a model, we have been able to precisely investigate (a) whether sex-specific neurons display a bias in neurotransmitter usage and (b) whether there are neurotransmitter dimorphisms in sex-shared neurons (Pereira et al. 2015; Gendrel et al. 2016; Serrano-Saiz et al. 2017b)(this paper). We found that male-specific neurons display a roughly similar proportional usage of individual neurotransmitter systems and note that male specific neurons display substantially more evidence of co-transmission, a possible reflection of their more elaborate morphology. We also confirmed evidence for sexual dimorphisms in neurotransmitter usage in sex-shared neurons (Table S4). These changes in neurotransmitter usage are usually correlated with changes in synaptic connectivity of these sex-shared neurons (Cook et al. 2019).

Neurotransmitter pathway genes in glia and gonad

Neurotransmitter uptake is a classic function of glial cells across animal phylogeny (Henn and Hamberger 1971), and such uptake mechanisms are observed in C. elegans as well. Previous reports demonstrated glutamate uptake by CEPsh (Katz et al. 2019) and GABA uptake by GLR glia (Gendrel et al. 2016). We now add to this list betaine uptake by most glia, as inferred from the expression pattern of SNF-3/BGT1 (Fig. S1, Table S1).

Studies in vertebrates have also suggested that specific glial cell types synthesize and release several neurotransmitters (Araque et al. 2014; Savtchouk and Volterra 2018). For example, astrocytes were recently shown to express VGLUT1 to release glutamate (De Ceglia et al. 2023). Evidence of neurotransmitter synthesis and release also exists in C. elegans. Previous work indicated that glia associated with male-specific spicule neurons synthesize (through cat-2/TH and bas-1/AAAD) the monoaminergic transmitter dopamine to control sperm ejaculation (Leboeuf et al. 2014). Our identification of cat-1/VMAT expression in these glia indicate that dopamine is released via the canonical vesicular monoamine transporter. We also detected expression of bas-1/AAAD in additional male and hermaphrodite glia, indicating the production of other signaling substances released by these glia. bas-1 has indeed recently been shown to be involved in the synthesis of a class of unconventional serotonin derivates (Yu et al. 2023).

We observed no additional examples of neurotransmitter synthesis and release by glia, based on the apparent absence of detectable expression of neurotransmitter synthesizing enzymes or any vesicular transporter (unc-17/VAChT, unc-47/VGAT, eat-4/VGLUT, cat-1/VMAT) in glia. Both observations are particularly notable in the context of previous reports on GABA synthesis and release from the AMsh glia cell type (Duan et al. 2020; Fernandez-Abascal et al. 2022). We were not able to detect AMsh with anti-GABA staining, nor with reporter alleles of unc-25/GAD. However, since very low levels of unc-25 are observed in AMsh scRNA datasets (Taylor et al. 2021; Purice et al. 2023), the usage of GABA by AMsh does warrant further investigation.

Outside the nervous system, the most prominent and functionally best characterized usage of neurotransmitters lies in the hermaphrodite somatic gonad, which has been shown to synthesize octopamine and use it to control oocyte quiescence (Alkema et al. 2005; Kim et al. 2021). Intriguingly, we also detected tbh-1, tdc-1, and cat-1 expression in the somatic gonad of the male, specifically the vas deferens, which is known to contain secretory granules that are positive for secretory molecular markers (Nonet et al. 1993). The presence of octopamine is unexpected because, unlike oocytes, sperm are not presently known to require monoaminergic signals for any aspect of their maturation. It will be interesting to assess sperm differentiation and function of tbh-1 or tdc-1 mutant animals.

Comparing approaches and caveats of expression pattern analysis

Our analysis also provides an unprecedented and relatively systematic comparison of antibody staining, CeNGEN scRNA transcript data, reporter transgene expression and knock-in reporter allele expression. The bottom-line conclusions of these comparisons are: (1) Reporter alleles reveal more sites of expression than fosmid-based reporters. It is unclear whether this is due to the lack of cis-regulatory elements in fosmid-based reporters or issues associated with the multicopy-nature of these reporters (e.g. RNAi-based gene silencing of multicopy arrays or squelching of regulatory factors). (2) The best possible reporter approaches (i.e. reporter alleles) show very good overlap with scRNA data, thereby validating each approach. However, our comparisons also show that no single approach is perfect. CeNGEN scRNA data can miss transcripts and can also show transcripts in cells in which there is no independent evidence for gene or protein expression. Conversely, antibody staining displays vagaries related to staining protocols and protein localization, which can be overcome with reporter approaches, but the price to pay with reporter alleles is that if they are based on SL2 or T2A strategies, they may fail to detect additional levels of posttranslational gene regulation. Posttranslational regulation may result in protein absence even in the presence of transcript. The existence of such mechanisms may be a possible explanation for cases where the expression of synthesis and/or transport machinery expression does not match up (e.g. TDC-1(-) plus TBH-1(+) neurons).

Our detailed analysis of reporter allele expression has uncovered several cases where fluorescent expression of a neurotransmitter pathway gene in a given neuron class appears very dim and variable from animal to animal. It is not clear whether this reflects on/off variability of gene expression, or, alternatively, whether expression levels merely hover around an arbitrary microscopical detection limit. Taking this notion a step further, we cannot exclude the possibility that expression observed with reporter alleles misses sites of expression. This possibility is raised by our inability to detect unc-25/GAD reporter allele expression in AMsh glia (Duan et al. 2020; Fernandez-Abascal et al. 2022) or eat-4 reporter allele expression in AVL and DVB neurons, in which some (but not other) multicopy reporter transgenes detected expression of the respective genes (Li et al. 2023). Functions of these genes in the respective cell types were corroborated by cell-type specific RNAi experiments and/or rescue experiments; whether there is indeed very low expression of these genes in those respective cells or whether drivers used in these studies for knock-down and/or rescue produce very low expression in other functionally relevant cells remains to be resolved.

Conclusions

In conclusion, we have presented here the most complete neurotransmitter map that currently exists for any animal nervous system. This neurotransmitter map comprises a critical step toward deciphering information flow in the nervous system and provides valuable tools for studying the genetic mechanisms underlying cell identity specification. Moreover, the neurotransmitter map presented here opens new opportunities for investigating sex-specific neuronal differentiation processes, particularly in the male-specific nervous system, where a scarcity of molecular markers has limited the analysis of neuronal identity control. Lastly, our analysis strongly suggests that additional neurotransmitter systems remain to be identified.

eLife Assessment

This compelling study reports the to-date most comprehensive neurotransmitter atlas of any organism, using fluorescent knock-in reporter lines. It represents an extremely useful tool for a broad audience of scientists interested in neuronal cell type differentiation and function.

Reviewer #1 (Public Review):

Wang and colleagues conducted a study to determine the neurotransmitter identity of all neurons in C. elegans hermaphrodites and males. They used CRISPR technology to introduce fluorescent gene expression reporters into the genomic loci of NT pathway genes. This approach is expected to better reflect in vivo gene expression compared to other methods like promoter- or fosmid-based transgenes, or available scRNA datasets. The study presents several noteworthy findings, including sexual dimorphisms, patterns of NT co-transmission, neuronal classes that likely use NTs without direct synthesis, and potential identification of unconventional NTs (e.g. betaine releasing neurons). The data is well-described and critically discussed, including a comparison with alternative methods. Although many of the observations and proposals have been previously discussed by the Hobert lab, the current study is particularly valuable due to its comprehensiveness. This NT atlas is the most complete and comprehensive of any nervous system that I am aware of, making it an extremely useful tool for the community.

Reviewer #2 (Public Review):

Summary:

Together with the known anatomical connectivity of C. elegans, a neurotransmitter atlas paves the way toward a functional connectivity map. This study refines the expression patterns of key genes for neurotransmission by analyzing the expression patterns from CRISPR-knocked-in GFP reporter strains using the color-coded Neuropal strain to identify neurons. Along with data from previous scRNA sequencing and other reporter strains, examining these expression patterns enhances our understanding of neurotransmitter identity for each neuron in hermaphrodites and the male nervous system. Beyond the known neurotransmitters (GABA, Acetylcholine, Glutamate, dopamine, serotonin, tyramine, octopamine), the atlas also identifies neurons likely using betaine and suggests sets of neurons employing new unknown monoaminergic transmission, or using exclusively peptidergic transmission.

Strengths:

The use of CRISPR reporter alleles and of the Neuropal strain to assign neurotransmitter usage to each neuron is much more rigorous than previous analysis and reveals intriguing differences between scRNA seq, fosmid reporter, and CRISPR knock-in approaches. Among other mechanisms, these differences between approaches could be attributed to 3'UTR regulatory mechanisms for scRNA vs. knockin or titration of rate-limited negative regulatory mechanisms for fosmid vs. knockin. It would be interesting to discuss this and highlight the occurrences of these potential phenomena for future studies.

Weaknesses:

For GABAergic transmission, one shortcoming arises from the lack of improved expression pattern by a knockin reporter strain for the GABA recapture symporter snf-11. In its absence, it is difficult to make a final conclusion on GABA recapture vs GABA clearance for all neurons expressing the vesicular GABA transporter neurons (unc-47+) but not expressing the GAD/UNC-25 gene e.g. SIA or R2A neurons. At minima, a comparison of the scRNA seq predictions versus the snf-11 fosmid reporter strain expression pattern would help to better judge the proposed role of each neuron in GABA clearance or recycling.

Considering the complexities of different tagging approaches, like T2A-GFP and SL2-GFP cassettes, in capturing post-translational and 3'UTR regulation is important. The current formulation is simplistic. e.g. after SL2 trans-splicing the GFP RNA lacks the 5' regulatory elements, T2A-GFP self-cleavage has its own issues, and the his-44-GFP reporter protein does certainly have a different post-translational life than vesicular transporters or cytoplasmic enzymes.

Do all splicing variants of neurotransmitter-related genes translate into functional proteins? The possibility that some neurons express a non-functional splice variant, leading to his-74-GFP reporter expression without functional neurotransmitter-related protein production is not addressed. Also, one tagged splice variant of unc-25 is expected to fail to produce a GFP reporter, can this cause trouble?

Reviewer #3 (Public Review):

Summary:

In this paper, Wang et al. provide the most comprehensive description and comparison of the expression of the different genes required to synthesize, transport, and recycle the most common neurotransmitters (Glutamate, Acetylcholine, GABA, Serotonin, Dopamine, Octopamine, and Tyramine) used by hermaphrodite and male C. elegans. This paper will be a seminal reference in the field. Building and contrasting observations from previous studies using fosmid, multicopy reporters, and single-cell sequencing, they now describe CRISPR/Cas-9-engineered reporter strains that, in combination with the multicolor pan-neuronal labeling of all C. elegans neurons (NeuroPAL), allows rigorous elucidation of neurotransmitter expression patterns. These novel reporters also illuminate previously unappreciated aspects of neurotransmitter biology in C. elegans, including sexual dimorphism of expression patterns, co-transmission, and the elucidation of cell-specific pathways that might represent new forms of neurotransmission.

Strengths:

The authors set out to establish neurotransmitter identities in C. elegans males and hermaphrodites via varying techniques, including integration of previous studies, examination of expression patterns, and generation of endogenous CRISPR-labeled alleles. Their study is comprehensive, detailed, and rigorous, and achieves the aims. It is an excellent reference for the field, particularly those interested in biosynthetic pathways of neurotransmission and their distribution in vivo, in neuronal and non-neuronal cells.

Weaknesses:

No weaknesses were noted. The authors do a great job linking their characterizations with other studies and techniques, giving credence to their findings. As the authors note, there are sexually dimorphic differences across animals and varying expression patterns of enzymes. While it is unlikely there will be huge differences in the reported patterns across individual animals, it is possible that these expression patterns could vary developmentally, or based on physiological or environmental conditions. It is unclear from the study how many animals were imaged for each condition, and if the authors noted changes across individuals during development (could be further acknowledged in the discussion?)

Author response:

The following is the authors’ response to the original reviews.

We would like to thank the reviewers and editor for their helpful comments and suggestions. In response, we have revised the manuscript in two main ways:

(1) To address the comments about rearranging figures and tables, we added a new Figure 3 that summarizes neurotransmitter assignments across all neuron classes. Our rationale for this change is detailed below.

(2) To address the comment on clarifying neurotransmitter synthesis versus uptake, we analyzed two additional reporter alleles that tag the monoamine uptake transporters for 5-HT and potentially tyramine. These results are now presented in a new Figure 8 and corresponding sections in the manuscript. Related tables have been updated to include this expression data. Two more authors have been added due to their contributions to these experiments.

For more detailed changes, please see our responses to the specific reviewer's comments as well as the revised manuscript.

Public Reviews:

Reviewer #1 (Public Review):

Wang and colleagues conducted a study to determine the neurotransmitter identity of all neurons in C. elegans hermaphrodites and males. They used CRISPR technology to introduce fluorescent gene expression reporters into the genomic loci of NT pathway genes. This approach is expected to better reflect in vivo gene expression compared to other methods like promoter- or fosmid-based transgenes, or available scRNA datasets. The study presents several noteworthy findings, including sexual dimorphisms, patterns of NT co-transmission, neuronal classes that likely use NTs without direct synthesis, and potential identification of unconventional NTs (e.g. betaine releasing neurons). The data is well-described and critically discussed, including a comparison with alternative methods. Although many of the observations and proposals have been previously discussed by the Hobert lab, the current study is particularly valuable due to its comprehensiveness. This NT atlas is the most complete and comprehensive of any nervous system that I am aware of, making it an extremely useful tool for the community.

Reviewer #2 (Public Review):

Summary:

Together with the known anatomical connectivity of C. elegans, a neurotransmitter atlas paves the way toward a functional connectivity map. This study refines the expression patterns of key genes for neurotransmission by analyzing the expression patterns from CRISPR-knocked-in GFP reporter strains using the color-coded Neuropal strain to identify neurons. Along with data from previous scRNA sequencing and other reporter strains, examining these expression patterns enhances our understanding of neurotransmitter identity for each neuron in hermaphrodites and the male nervous system. Beyond the known neurotransmitters (GABA, Acetylcholine, Glutamate, dopamine, serotonin, tyramine, octopamine), the atlas also identifies neurons likely using betaine and suggests sets of neurons employing new unknown monoaminergic transmission, or using exclusively peptidergic transmission.

Strengths:

The use of CRISPR reporter alleles and of the Neuropal strain to assign neurotransmitter usage to each neuron is much more rigorous than previous analysis and reveals intriguing differences between scRNA seq, fosmid reporter, and CRISPR knock-in approaches. Among other mechanisms, these differences between approaches could be attributed to 3'UTR regulatory mechanisms for scRNA vs. knockin or titration of rate-limited negative regulatory mechanisms for fosmid vs. knockin. It would be interesting to discuss this and highlight the occurrences of these potential phenomena for future studies.

We recognize that readers of this study may be interested in understanding the differences between the three approaches. Therefore, in the Introduction, we addressed the potential risk of overexpression artifacts associated with multicopy transgenes, such as fosmid-based reporters, which can affect rate-limiting negative regulatory mechanisms. Additionally, in the Discussion, we included a section titled 'Comparing approaches and caveats of expression pattern analysis' to further explore these comparative methods and their associated nuances.

Weaknesses:

For GABAergic transmission, one shortcoming arises from the lack of improved expression pattern by a knockin reporter strain for the GABA recapture symporter snf-11. In its absence, it is difficult to make a final conclusion on GABA recapture vs GABA clearance for all neurons expressing the vesicular GABA transporter neurons (unc-47+) but not expressing the GAD/UNC-25 gene e.g. SIA or R2A neurons. At minima, a comparison of the scRNA seq predictions versus the snf-11 fosmid reporter strain expression pattern would help to better judge the proposed role of each neuron in GABA clearance or recycling.

The snf-11 fosmid-based reporter data shows very good overlap with scRNA seq predictions (now included in Supp. Table S1).

But there are two much stronger reasons why we did not seek to further the analysis of expression of the snf-11 GABA uptaker:

(1) Due to available anti-GABA staining data, we do know which neurons have the potential to take up GABA (via SNF-11).

(2) Focusing on SNF-11 function rather than expression, we can ask which neurons lose anti-GABA staining in snf-11 mutants.

Both of these types of analyses have been done in an earlier study from our lab (Gendrel et al., 2016, PMID 27740909), which, among other things, investigated GABA uptake mechanisms via SNF-11. Apart from analyzing the expression of a fosmid-based snf-11 reporter, we immunostained worms for GABA in both snf-11 mutant and wild type backgrounds (results summarized in Tables 1 and 2 of Gendrel et al.). Of the neurons that typically stain for GABA (Table 1, Gendrel et al.), two neuron classes (ALA and AVF) lost the staining in snf-11 mutants, suggesting that these neurons likely uptake GABA via SNF-11. Importantly, one of the neurons the reviewer mentioned, R2A, stains for GABA in both wild type and snf-11 mutants, indicating that it likely does not uptake GABA via SNF-11. The other neuron mentioned, SIA, does not stain for GABA in wild type (Table 2, Gendrel et al.), hence not a GABA uptake neuron. In cases like SIA and other neurons, where a neuron does not express unc-25 but does express unc-47 reporters (either fosmid or CRISPR reporter alleles), we speculate that UNC-47 transport another neurotransmitter.

Considering the complexities of different tagging approaches, like T2A-GFP and SL2-GFP cassettes, in capturing post-translational and 3'UTR regulation is important. The current formulation is simplistic. e.g. after SL2 trans-splicing the GFP RNA lacks the 5' regulatory elements, T2A-GFP self-cleavage has its own issues, and the his-44-GFP reporter protein does certainly have a different post-translational life than vesicular transporters or cytoplasmic enzymes.

Yes, agreed, these points are mentioned in the Introduction and discussed in "Comparing approaches and caveats of expression pattern analysis" in the Discussion.

Do all splicing variants of neurotransmitter-related genes translate into functional proteins? The possibility that some neurons express a non-functional splice variant, leading to his-74-GFP reporter expression without functional neurotransmitter-related protein production is not addressed.

We thank the reviewer for bringing up this really interesting point, which we had not considered. First and foremost, with the exception of unc-25 (discussed in the next point), for all other genes that produce multiple splice forms, we made sure to append our tag (at 5’ or 3’ end) such that the expression of all splice forms is captured. The reviewer raises the interesting point that in an alternative splicing scenario, some of the cells that express the primary transcript may “switch” to an inactive form. While we cannot exclude this possibility, we have confirmed by sequence analysis in WormBase that in five of the six cases where there is alternative splicing, the alternatively spliced exon lies outside the conserved, functionally relevant (enzymatic or structural) domain. In one case, unc-25, a shorter isoform is produced that does cut into the functionally relevant domain; however, since all unc-25 reporter allele expression cells are also staining positive for GABA, this may not be an issue.

Also, one tagged splice variant of unc-25 is expected to fail to produce a GFP reporter, can this cause trouble?

Yes, there is indeed a third splice variant of unc-25 with an alternative C-terminus. To address potential expression of this isoform, we CRISPR-engineered another reporter, unc-25(ot1536[unc-25b.1::t2a::gfp::h2b]), in which the inserted t2a::gfp::h2b sequences are fused to the C-terminus of the alternative splice form, but we did not observe any expression of this reporter. Now included in the manuscript.

Reviewer #3 (Public Review):

Summary:

In this paper, Wang et al. provide the most comprehensive description and comparison of the expression of the different genes required to synthesize, transport, and recycle the most common neurotransmitters (Glutamate, Acetylcholine, GABA, Serotonin, Dopamine, Octopamine, and Tyramine) used by hermaphrodite and male C. elegans. This paper will be a seminal reference in the field. Building and contrasting observations from previous studies using fosmid, multicopy reporters, and single-cell sequencing, they now describe CRISPR/Cas-9-engineered reporter strains that, in combination with the multicolor pan-neuronal labeling of all C. elegans neurons (NeuroPAL), allows rigorous elucidation of neurotransmitter expression patterns. These novel reporters also illuminate previously unappreciated aspects of neurotransmitter biology in C. elegans, including sexual dimorphism of expression patterns, cotransmission, and the elucidation of cell-specific pathways that might represent new forms of neurotransmission.

Strengths:

The authors set out to establish neurotransmitter identities in C. elegans males and hermaphrodites via varying techniques, including integration of previous studies, examination of expression patterns, and generation of endogenous CRISPR-labeled alleles. Their study is comprehensive, detailed, and rigorous, and achieves the aims. It is an excellent reference for the field, particularly those interested in biosynthetic pathways of neurotransmission and their distribution in vivo, in neuronal and non-neuronal cells.

Weaknesses:

No weaknesses were noted. The authors do a great job linking their characterizations with other studies and techniques, giving credence to their findings. As the authors note, there are sexually dimorphic differences across animals and varying expression patterns of enzymes. While it is unlikely there will be huge differences in the reported patterns across individual animals, it is possible that these expression patterns could vary developmentally, or based on physiological or environmental conditions. It is unclear from the study how many animals were imaged for each condition, and if the authors noted changes across individuals during development (could be further acknowledged in the discussion?)

We have updated the Methods section to specify the number of animals used for imaging. We agree with the reviewer that documenting the developmental dynamics of neurotransmitter expression would be interesting. However, except for one gene (tph-1, Fig. S2), we did not analyze the expression during different developmental stages for most genes in this study. Following the reviewer's suggestion, we have included this as a potential future direction in "Conclusions" at the end of the revised manuscript.

Recommendations for the authors:

After the consultation session, a common suggestion from the reviewers is to bring the tables more upfront, perhaps even in the form of legible main Figures and in alphabetical order of neurons; since we believe that the study will be in the long-term often used for these data; while the Figures with fluorescent expression patterns could be moved to the supplemental information.

We appreciate the reviewers' and editor's acknowledgment of the tables' possibly frequent usage by the field. We have considered carefully how to order the data presentation. We prefer to keep most of the fluorescent figures in the main text because they convey important subtleties that we want the reader to be aware of.

To address the suggestions to bring key data more upfront, we have added an entirely new figure (Figure 3) before the ensuing data figures that summarized expression patterns of the fluorescent reporters. This new figure (A) summarizes the neurotransmitter use for all neuron classes and (B) illustrates this information within worm schematics, showing the position of neurons in the whole worm. This figure serves as a good overview of neurotransmitter assignments but also specifically refers to the more extensive data and supplementary tables with detailed notes. We believe this solution effectively balances the need for comprehensive information and ease of reference.

Reviewer #1 (Recommendations for The Authors):

Suggestions:

(1) The study contains up to 10 Figures with gene expression patterns; however, I believe the community will use this paper mostly in the future for its summarizing tables. I wonder if it would be more useful to edit the tables and move them to the main figures while most fluorescent reporter images could be moved to the supplementary part.

Yes, as mentioned above, we made new summary table & schematic upfront. We do prefer to keep primary data in main figure body. Please see above (Public Review & Response).

(2) In the section titled 'Neurotransmitter Synthesis versus Uptake', the author's wording could be more careful. The data rather suggests functions for individual neuronal classes, such as clearance neurons or signaling neurons. However, these functions remain hypotheses until further detailed studies are conducted to test them.

These are fair points. We have made several improvements:

(1) In the referenced section, we added a sentence at the end of the paragraph on betaine to suggest the importance of future functional studies.

(2) We analyzed reporter allele expression for two additional genes: the known uptake transporter for 5-HT (mod-5, reporter allele vlc47) and the predicted uptake transporter for tyramine (oct-1, reporter allele syb8870). The results from these experiments are presented in the new Figure 8 and discussed in Results and Discussion correspondingly. We also collaborated with Curtis Loer, who conducted anti-5-HT staining in wild type and mod-5 mutant animals (results shown in Figure 12). These experiments have enhanced our understanding of 5-HT uptake mechanisms and potential tyramine uptake mechanisms.

(3) At the end of the Conclusions, we emphasized the need for future detailed studies to test the functions of neurotransmitter synthesis and uptake.

(3) Page 21; add to the discussion: neurons could use mainly electrical synapses for communication. Especially for RMG neurons, this might be the case (in addition to neuropeptide communication).

“Main usage” is a difficult term to use. If there were neurons that are clearly devoid of any form of synaptic vesicle (small or DCV; note that RMG has plenty of DCVs), but show robust and reproducible electrical synapses, we would agree that such neurons could primarily be a “coupling” neuron. But this call is very hard to make for any C. elegans neuron (RMG included) and hence we prefer to not add further to an already quite long Discussion section.

(4) Page 23: I believe that multi-copy promoter-based transgenes (despite array suppression mechanisms) could be potentially more sensitive than single-copy insertion of fluorescent reporters. In our lab, we observed this a couple of times. This could be discussed.

We discuss this in "Comparing approaches and caveats of expression pattern analysis" in the Discussion.

We have also added a third possibility (i.e. technical issues related to neuron-ID) in the revised manuscript.

Reviewer #2 (Recommendations For The Authors):

Comment during consultation session: As for my feedback on the lack of an SNF-11 reporter strain, exercising more caution in their conclusions would suffice for me. Other comments are simple edits/discussion.

Please see above.

Several neurotransmitter symporters exist in the C. elegans genome, does any express specifically in the "orphan" UNC-47+ neurons?

Yes, good point, we considered this possibility, but of the >10 SLC6-family of neurotransmitter reporters, only the classic, de-orphanized ones that we discuss here in the paper show robust scRNA signals (as discussed in the paper) and none of those give clues about the orphan unc-47(+) neurons.

Based on UNC-47+ expression the article suggests a "Novel inhibitory neurotransmitter". Why would any new neurotransmitter using UNC-47 be necessarily inhibitory? The presence of one potential glycine-gated anion channel and one GPCR in C. elegans genome sounds poor evidence to suggest a sign of glycine or b-alanine transmission.

Yes, agreed, it does not need to be inhibitory. Fixed in Results and Discussion.

To help readers the expression of the knocked in GFP in neurons should not be reported as binary in table S1 which leads to a feeling of strong discrepancy between scRNA seq and CRISPR GFP, which is not the case.

There might be some misunderstanding regarding the coloring in this table. To clarify, the green-filled Excel cells denote the expression of reporters utilized in prior studies, rather than the CRISPR reporter alleles. Expression of the CRISPR alleles is instead indicated on the left side of the neuron names, marked as "CRISPR+" in green font. For signifying absence of expression, we used "no CRISPR" in red font in the first submission. We have now changed it into "CRISPR-" for greater clarity.

The variable expression of reporter GFP between individuals for the same neuron is intriguing. It is unclear if this is observed only for dim neurons or can be more of an ON/OFF expression.

Variability only occurs for dim expression. We have now clarified this point in Discussion, "Comparing approaches and caveats of expression pattern analysis".

The multiple occurrences of co-transmission, especially in male neurons, are interesting. It will be interesting in the future to establish whether the neurotransmitters are synaptically segregated or coreleased. As the section on sexual dimorphism of neurotransmitter usage does not discuss novel information coming from this study, it is not very necessary.

Agreed. We added this perspective to the Discussion, "Co-transmission of multiple neurotransmitters".

In the abstract, dopamine is missing in the main known transmitter.

Fixed. Thanks for spotting this.

Reviewer #3 (Recommendations For The Authors):

Great article. Minor suggestions to strengthen presentation:

Figure 1B is hard to interpret. There could be more intuitive ways of representing the data and the methodologies that support a given expression pattern. Neurons should also be reordered by alphabetical order rather than expression levels to facilitate finding them.

We considered alternative ways of presenting this data, but, regrettably, did not come up with a better approach. To clarify, the primary focus of Fig. 1B is to compare expression of previously reported reporters and scRNA data, which was quite literally the initial impetus for our analysis, i.e. we noted strong scRNA signals that had not previously been supported by transgenic reporter data. For a comprehensive version of the table that includes more details on the expression of CRISPR reporter alleles, please refer to Table S1, which we referenced in the figure legend.

GFP-only channel images in Figures 3, 4, 5, and 9 sometimes show dim signals that the authors are highlighting as new findings. We recommend using the inverted grayscale version of that channel since the contrast of dim signals is more noticeable to the human eye rather than when the image is colorized.

Good point, we implemented these suggestions in the figures the reviewer mentioned, now re-numbered Figures 4, 5, 6, and 12. For Figure 6 (tph-1, bas-1, and cat-1 expression in hermaphrodites), we used a new cat-1 head image to reflect the newly identified ASI and AVL expression that wasn’t readily visible in the original projection used in the earlier version of this manuscript. We also added grayscale images in Figure 13 to reflect dim tbh-1 expression in IL2 neurons more clearly.

A plan to integrate this new information into WormAtlas. The C. elegans community is characterized by the open sharing of information on platforms that are user-friendly and accessible. Ideally, the new information would not just 'erase' what was observed before but will describe the new observations and will let the community reach their own conclusions since there is no perfect method and even these CRISPR/Cas9 reporter strains are only proxy for gene expression that subject to post-transcriptional regulation since they depend on T2A and SL2 sequences.

We completely agree with the reviewer’s suggestion. We will coordinate with WormAtlas on integrating this new information.

In the case of neurons that were removed from using a specific neurotransmitter, like PVQ. What do the authors conclude overall, if it does not use glutamate, are there any new hypotheses to what it could be using?

Since all neurons express multiple neuropeptides, we hypothesize neurons such as PVQ may be primarily peptidergic. This is included in Discussion, "Neurons devoid of canonical neurotransmitter pathway genes may define neuropeptide-only neurons".

In Table S5, the I4 neuron is listed as a variable for eat-4 expression but in Table S1 it says that there was no CRISPR expression detected. Which one is correct?

Thanks for spotting this. Table S5 is correct, we saw very dim and variable expression of the eat-4 reporter allele in I4. Table S1 is fixed now.

Additional discussion points that might be important for the community:

CRIPSR strains used here should be deposited in the CGC.

Yes, all strains generated in this study have already been deposited to CGC.

It would be great to have an additional discussion point on how the neural clusters in CenGEN were defined based on the fosmid reporter expression, so in a way using the defining factor as one that was already defined by it might make results confusing.

Neural cluster definition in CeNGEN did not rely on isolated data points but on the combination of many expression reagents, each with its own shortcomings, but in combination providing reliable identification. Since one feedback we have gotten from many readers of our manuscript is that it is already very long as is, we prefer not to dilute the discussion further.

It would be important to discuss the rate of neurotransmitter genes that have variable expression patterns. Are any of those genes used in NeuroPAL to define specific neuronal classes? This is important to describe as NeuroPAL labeling is being used to define neuronal identity.

All the reporters used in NeuroPAL are promoter-based, very robust and do not include the full loci of genes, so they are not directly comparable with the CRISPR reporter alleles in this study. However, we recognize that some expression pattern variability could be confusing. We have discussed this more in the section "Comparing approaches and caveats of expression pattern analysis" in the Discussion.

Version 2

Abstract

Abstract

Mapping neurotransmitter identities to neurons is key to understanding information flow in a nervous system. It also provides valuable entry points for studying the development and plasticity of neuronal identity features. In the C. elegans nervous system, neurotransmitter identities have been largely assigned by expression pattern analysis of neurotransmitter pathway genes that encode neurotransmitter biosynthetic enzymes or transporters. However, many of these assignments have relied on multicopy reporter transgenes that may lack relevant cis-regulatory information and therefore may not provide an accurate picture of neurotransmitter usage. We analyzed the expression patterns of 16 CRISPR/Cas9-engineered knock-in reporter strains for all main types of neurotransmitters in C. elegans (glutamate, acetylcholine, GABA, serotonin, dopamine, tyramine, and octopamine) in both the hermaphrodite and the male. Our analysis reveals novel sites of expression of these neurotransmitter systems within both neurons and glia, as well as non-neural cells. The resulting expression atlas defines neurons that may be exclusively neuropeptidergic, substantially expands the repertoire of neurons capable of co-transmitting multiple neurotransmitters, and identifies novel neurons that uptake monoaminergic neurotransmitters. Furthermore, we also observed unusual co-expression patterns of monoaminergic synthesis pathway genes, suggesting the existence of novel monoaminergic transmitters. Our analysis results in what constitutes the most extensive whole-animal-wide map of neurotransmitter usage to date, paving the way for a better understanding of neuronal communication and neuronal identity specification in C. elegans.

Introduction

Understanding information processing in the brain necessitates the generation of precise maps of neurotransmitter deployment. Moreover, comprehending synaptic wiring diagrams is contingent upon decoding the nature of signaling events between anatomically connected neurons. Mapping of neurotransmitter identities onto individual neuron classes also presents a valuable entry point for studying how neuronal identity features become genetically specified during development and potentially modified in response to specific external factors (such as the environment) or internal factors (such as sexual identity or neuronal activity patterns).

The existence of complete synaptic wiring diagrams of the compact nervous system of male and hermaphrodite C. elegans nematodes raises questions about the molecular mechanisms by which individual neurons communicate with each other. C. elegans employs the main neurotransmitter systems that are used throughout the animal kingdom, including acetylcholine, glutamate, γ-aminobutyric acid (GABA), and several monoamines (Sulston et al. 1975; Horvitz et al. 1982; Loer and Kenyon 1993; Mcintire et al. 1993; Duerr et al. 1999; Lee et al. 1999; Duerr et al. 2001; Alkema et al. 2005; Duerr et al. 2008; Serrano-Saiz et al. 2013; Pereira et al. 2015; Gendrel et al. 2016; Serrano-Saiz et al. 2017b)(Fig. 1A). Efforts to map these neurotransmitter systems to individual cell types throughout the entire nervous system have a long history, beginning with the use of chemical stains that directly detected a given neurotransmitter (dopamine)(Sulston et al. 1975), followed by antibody staining of neurotransmitter themselves (serotonin and GABA)(Horvitz et al. 1982; Mcintire et al. 1993) or antibody stains of biosynthetic enzymes or neurotransmitter vesicular transporters (acetylcholine and monoamines)(LOER AND KENYON 1993; Duerr et al. 1999; Duerr et al. 2001; Alkema et al. 2005; Duerr et al. 2008)(see Fig. 1A for an overview of these enzymes and transporters).

Background on genes examined in this paper.

(A) Neurotransmitter synthesis and transport pathways. TH = tyrosine hydroxylase; TDC = tyrosine decarboxylase; TBH = tyramine beta hydroxylase; TPH = tryptophan hydroxylase; GAD = glutamic acid decarboxylase; AAAD = aromatic amino acid decarboxylase; VMAT = vesicular monoamine transporter; VAChT = vesicular acetylcholine transporter; VGAT = vesicular GABA transporter. Taken and modified from (HOBERT 2013).

(B) Graphic comparison of scRNA expression data and previously reported reporter expression data. See Table S1 for a more comprehensive version that includes expression of reporter genes in cells that show no scRNA transcripts. Note that scRNA expression values for eat-4 and unc-47 can be unreliable because they were overexpressed to isolate individual neurons for scRNA analysis (Taylor et al. 2021).

While these early approaches proved successful in revealing neurotransmitter identities, they displayed several technical limitations. Since neurotransmitter-synthesizing or - transporting proteins primarily localize to neurites, the cellular identity of expressing cells (usually determined by assessing cell body position) often could not be unambiguously established in several, particularly cell- and neurite-dense regions of the nervous system. One example concerns cholinergic neurons, which are defined by the expression of the vesicular acetylcholine transporter UNC-17/VAChT and choline acetyltransferase CHA-1/ChAT. While mainly neurite-localized UNC-17 and CHA-1 antibody staining experiments could identify a subset of cholinergic neurons (Duerr et al. 2001; Duerr et al. 2008), many remained unidentified (Pereira et al. 2015). In addition, for GABA-producing neurons, it became apparent that antibody-based GABA detection was dependent on staining protocols, leading to the identification of “novel” anti-GABA-positive neurons, i.e. GABAergic neurons, more than 20 years after the initial description of GABAergic neurons (Mcintire et al. 1993; Gendrel et al. 2016).

An alternative approach to mapping neurotransmitter usage has been the use of reporter transgenes. This approach has the significant advantage of allowing the fluorophore to either fill the entire cytoplasm of a cell or to be targeted to the nucleus, thereby facilitating neuron identification. However, one shortcoming of transgene-based reporter approaches is that one cannot be certain that a chosen genomic region, fused to a reporter gene, indeed contains all cis-regulatory elements of the respective locus. In fact, the first report that described the expression of the vesicular glutamate transporter EAT-4, the key marker for glutamatergic neuron identity, largely underestimated the number of eat-4/VLGUT-positive and, hence, glutamatergic neurons (Lee et al. 1999). The introduction of fosmid-based reporter transgenes has largely addressed such concerns, as these reporters, with their 30-50 Kb size, usually cover entire intergenic regions (SAROV et al. 2012). Indeed, such fosmid-based reporters have been instrumental in describing the entire C. elegans glutamatergic nervous system, defined by the expression of eat-4/VGLUT (Serrano-Saiz et al. 2013), as well as supposedly the complete set of cholinergic (Pereira et al. 2015) and GABAergic neurons (Gendrel et al. 2016).

However, even fosmid-based reporters may not be the final word. In theory, they may still miss distal cis-regulatory elements. Moreover, the multicopy nature of transgenes harbors the risk of overexpression artifacts, such as the titrating of rate-limiting negative regulatory mechanisms. Also, RNAi-based silencing mechanisms triggered by the multicopy nature of transgenic reporter arrays have the potential to dampen the expression of reporter arrays (NANCE AND FROKJAER-JENSEN 2019). One way to get around these limitations, while still preserving the advantages of reporter gene approaches, is to generate reporter alleles in which an endogenous locus is tagged with a reporter cassette, using CRISPR/Cas9 genome engineering. Side-by-side comparisons of fosmid-based reporter expression patterns with those of knock-in reporter alleles indeed revealed several instances of discrepancies in expression patterns of homeobox genes (Reilly et al. 2022).

An indication that previous neurotransmitter assignments may not have been complete was provided by recent single-cell RNA (scRNA) transcriptomic analyses of the hermaphrodite nervous system by the CeNGEN consortium (Taylor et al. 2021). As we describe in this paper in more detail, transcripts for several neurotransmitter-synthesizing enzymes or transporters were detected in a few cells beyond those previously described to express the respective genes. This motivated us to use CRISPR/Cas9 engineering to fluorescently tag a comprehensive panel of genetic loci that code for neurotransmitter-synthesizing, -transporting, and -uptaking proteins (“neurotransmitter pathway genes”). Using the landmark strain NeuroPAL for neuron identification (Yemini et al. 2021), we identified novel sites of expression of most neurotransmitter pathway genes. Furthermore, we used these reagents to expand and refine neurotransmitter maps of the entire nervous system of the C. elegans male, which contains almost 30% more neurons than the nervous system of the hermaphrodite yet lacks a reported scRNA transcriptome atlas. Together with the NeuroPAL cell-identification tool, these reporter alleles allowed us to substantially improve the previously described neurotransmitter map of the male nervous system (Serrano-Saiz et al. 2017b). Our analysis provides insights into the breadth of usage of each individual neurotransmitter system, reveals instances of co-transmitter use, indicates the existence of neurons that may entirely rely on neuropeptides instead of classic neurotransmitters, reveals sexual dimorphisms in neurotransmitter usage, and suggests the likely existence of presently unknown neurotransmitters.

Materials and methods

Transgenic reporter strains

Knock-in reporter alleles were generated either by SunyBiotech (syb alleles) or in-house (ot alleles) using CRISPR/Cas9 genome engineering. Most genes were tagged with a nuclear-targeted gfp sequence (gfp fused to his-44, a histone h2b gene) at the 3’ end of the locus to capture all isoforms, except tdc-1 which was tagged at the 5’ end. For unc-25, both isoforms were individually tagged since a single tag would not capture both. Transgene schematics are shown in Fig. 2.

Schematics of reporter knock-in alleles.

Reporter alleles were generated by CRISPR/Cas9 genome engineering. The SL2 or T2A-based separation of the reporter from the coding sequence of the respective loci enables targeting of the reporter to the nucleus (via the H2B tag), which in turn facilitates identification of the cell expressing a given reporter. Genome schematics are from WormBase (Davis et al. 2022).

Reporter alleles generated in this study:

unc-25(ot1372[unc-25a.1c.1::t2a:gfp::h2b]) III

unc-25(ot1536[unc-25b.1::t2a::gfp::h2b]) III

unc-46(syb7278[unc-46::sl2::gfp:h2b]) V

unc-47(syb7566[unc-47::sl2::gfp::h2b]) III

cat-1(syb6486[cat-1::sl2::gfp::h2b]) X

tph-1(syb6451[tph-1::sl2::gfp::h2b]) II

tbh-1(syb7786[tbh-1::sl2::gfp::h2b]) X

tdc-1(syb7768[gfp::linker::h2b::t2a::tdc-1]) II

cat-2(syb8255[cat-2::sl2::gfp::h2b]) II

snf-3(syb7290[snf-3::TagRFP::sl2::gfp::h2b]) II

oct-1(syb8870[oct-1::sl2::gfp::h2b]) I

hdl-1(syb1048[hdl-1::gfp]) IV

hdl-1(syb4208[hdl-1::t2a::3xnls::cre]) IV

Since we did not detect fluorophore signals in the hdl-1(syb1048[hdl-1::gfp]) strain, we attempted to amplify low level signals, by inserting Cre recombinase at the C-terminus of the hdl-1 locus (hdl-1(syb4208[hdl-1::t2a::3xnls::cre])). We crossed this strain to the recently published “Flexon” strain (arTi361[rps-27p::gfp"flexon"-h2b::unc-54-3’UTR]) (SHAFFER AND GREENWALD 2022). Even low expression of hdl-1 should have led to Cre-mediated excision of the flexon stop cassette, which is designed to abrogate gene expression by a translational stop and frameshift mutation, and subsequently can result in strong and sustained gfp expression under the control of the rps-27 promoter and thereby providing information about cell-specific hdl-1 expression. However, no robust, consistent reporter expression was seen in hdl-1(syb4208[hdl-1::t2a::3xnls::cre]); arTi361[rps-27p::gfp"flexon"-h2b::unc-54-3’UTR] animals.

Three of the reporter alleles that we generated were already previously examined in specific cellular contexts:

unc-17(syb4491[unc-17::t2a::gfp:h2b]) IV (Vidal et al. 2022)

eat-4(syb4257[eat-4::t2a::gfp::h2b]) III (Vidal et al. 2022)

bas-1(syb5923[bas-1::sl2::gfp::h2b]) III (Yu et al. 2023)

One of the reporter alleles was obtained from the Caenorhabditis Genetics Center (CGC):

mod-5(vlc47[mod-5::t2a::mNeonGreen]) I (MAICAS et al. 2021)

Microscopy and image processing

For adult animal imaging, 15-25 (exact number depending on the difficulty of neuron ID) same-sex L4 worms were grouped on NGM plates 6-9 hours prior to imaging to control for accurate staging and avoid mating. Young adult worms were then anesthetized using 50—100 mM sodium azide and mounted on 5% agarose pads on glass slides. Z-stack images were acquired with ZEN software using Zeiss confocal microscopes LSM880 and LSM980 or a Zeiss Axio Imager Z2 and processed with ZEN software or FIJI (Schindelin et al. 2012) to create orthogonal projections. Brightness and contrast, and in some cases gamma values, were adjusted to illustrate dim expression and facilitate neuron identification.

Neuron class and cell type identification

Neuron classes were identified by crossing the gfp reporter alleles with the landmark strain “NeuroPAL” (allele otIs669 or otIs696, for bright reporters and dim reporters, respectively) and following published protocols (Tekieli et al. 2021; Yemini et al. 2021) (also see “lab resources” at hobertlab.org). For neuron identification of the eat-4(syb4257), unc-46(syb7278), and unc-47(syb7566) reporter alleles in hermaphrodites, the reporter alleles were also crossed into the fosmid-based reporter transgenes of the same gene [eat-4(otIs518), unc-46(otIs568), and unc-47(otIs564)] as a “first-pass” to identify potential non-overlapping expression of the two alleles. For tph-1(syb6451) analysis, an eat-4 fosmid-based reporter (otIs518) was also used. For identification of VC4, VC5, HSN, and uv1, an ida-1p::mCherry integrant (LX2478, lin-15(n765ts) X; vsls269[ida-1::mCherry]) was also used in some cases (Fernandez et al. 2020). For phasmid neurons, dye-filling with DiD (Thermo Fisher Scientific) was sometimes used to confirm neuron ID. For glial expression, a panglial reporter otIs870[mir-228p::3xnls::TagRFP] was used. For hypodermal cells identification, a dpy-7p::mCherry reporter stIs10166 [dpy-7p::his-24::mCherry + unc-119(+)] was used. (Liu et al. 2009)

Results

Comparing CeNGEN scRNA data to reporter gene data

To investigate the neurotransmitter identity of neurons throughout the entire C. elegans nervous system of both sexes, we consider here the expression pattern of the following 15 genetic loci (see also Fig. 1A):

eat-4/VGLUT: expression of the vesicular glutamate transporter is alone sufficient to define glutamatergic neuron identity (Lee et al. 1999; Serrano-Saiz et al. 2013).

unc-17/VAChT: expression of the vesicular acetylcholine transporter, located in an operon together with the acetylcholine-synthesizing gene cha-1/ChAT (Alfonso et al. 1994), defines cholinergic neurons (Duerr et al. 2001; Duerr et al. 2008; Pereira et al. 2015).

unc-25/GAD, unc-47/VGAT and its sorting co-factor unc-46/LAMP: expression of these three genes defines neurons that synthesize and release GABA (Mcintire et al. 1993; Mcintire et al. 1997; Jin et al. 1999; Schuske et al. 2007; Gendrel et al. 2016). Additional neurons that we classify as GABAergic are those that do not synthesize GABA (unc-25/GAD-negative), but take up GABA from other neurons (based on anti-GABA antibody staining) and are expected to release GABA based on unc-47/VGAT expression (Gendrel et al. 2016). unc-47/VGAT expression without any evidence of GABA synthesis or uptake (unc-25/GAD- and anti-GABA-negative) is indicative of an unknown transmitter being present in these cells and utilizing unc-47/VGAT for vesicular secretion.

tph-1/TPH and bas-1/AAAD: the co-expression of these two biosynthetic enzymes, together with the co-expression of the monoamine vesicular transporter cat-1/VMAT, defines all serotonin-synthesizing and -releasing neurons (Fig. 1A)(Horvitz et al. 1982; Duerr et al. 1999; Sze et al. 2000; HARE AND LOER 2004).

cat-2/TH and bas-1/AAAD: the co-expression of these two biosynthetic enzymes, together with the co-expression of the monoamine vesicular transporter cat-1/VMAT, defines all dopamine-synthesizing and -releasing neurons (Fig. 1A)(Sulston et al. 1975; Duerr et al. 1999; LINTS AND EMMONS 1999; HARE AND LOER 2004).

tdc-1/TDC: defines, together with cat-1/VMAT, all tyramine-synthesizing and -releasing neurons (Fig. 1A)(Alkema et al. 2005).

tbh-1/TBH: expression of this gene, in combination with that of tdc-1/TDC and cat-1/VMAT, defines octopamine-synthesizing and -releasing neurons (Fig. 1A) (Alkema et al. 2005).

cat-1/VMAT: expression of this vesicular monoamine transporter defines all four above-mentioned monoaminergic neurons (serotonin, dopamine, tyramine, octopamine)(Duerr et al. 1999), but as described and discussed below, it may also define additional sets of monoaminergic neurons.

hdl-1/AAAD: hdl-1, a previously uncharacterized gene, encodes the only other AAAD with sequence similarity to the bas-1 and tdc-1 AAAD enzymes that produce other bona fide monoamines (Fig. S1)(HARE AND LOER 2004). hdl-1 expression may therefore, in combination with cat-1/VMAT, identify neurons that produce and release trace amines of unknown identity.

snf-3/BGT1/SLC6A12: this gene encodes the functionally validated orthologue of the vertebrate betaine uptake transporter SLC6A12 (i.e. BGT1)(Peden et al. 2013). In combination with the expression of cat-1/VMAT, which synaptically transports betaine (Hardege et al. 2022), snf-3 expression may identify neurons that utilize betaine as a synaptically released neurotransmitter to gate betaine-gated ion channels, such as ACR-23 (Peden et al. 2013) or LGC-41 (Hardege et al. 2022).

mod-5/SERT: this gene codes for the functionally validated orthologue of the vertebrate serotonin uptake transporter SERT (Ranganathan et al. 2001), which defines neurons that take up 5-HT independently of their ability to synthesize 5-HT and, depending on their expression of cat-1/VMAT, may either re-utilize serotonin for synaptic signaling or serve as 5-HT clearance neurons.

oct-1/OCT: this gene encodes the closest representative of the OCT subclass of SLC22 organic anion transporters (Zhu et al. 2015), several members of which are selective uptake transporters of tyramine (Breidert et al. 1998; Berry et al. 2016). Its expression or function in the nervous system had not previously been analyzed in C. elegans.

For all these 15 genetic loci, we compared scRNA transcriptome data from the CeNGEN scRNA atlas (at all 4 available stringency levels (Taylor et al. 2021)) to previously published reporter and antibody staining data. As shown in Fig. 1B and Table S1, such comparisons reveal the following: (1) scRNA data support the expression of genes in the vast majority of neurons in which those genes were found to be expressed with previous reporter gene approaches. In most cases, this is true even at the highest threshold levels for scRNA detection. (2) Vice versa, reporter gene expression supports scRNA transcriptome data for a specific neurotransmitter system in the great majority of cells. (3) In spite of this congruence, there were several discrepancies between reporter data and scRNA data. Generally, while valuable, scRNA transcriptome data cannot be considered the final word for any gene expression pattern assignments. Lack of detection of transcripts could be a sensitivity issue and, conversely, the presence of transcripts does not necessarily indicate that the respective protein is generated, due to the possibility of posttranscriptional regulation.

Hence, to consolidate and further improve neurotransmitter identity assignment throughout the entire C. elegans nervous system, and to circumvent potential limitations of multicopy, fosmid-based reporter transgenes on which previous neurotransmitter assignments have been based, we engineered and examined expression patterns of 16 knock-in reporter alleles of neurotransmitter synthesis, vesicular transport, and uptake loci listed above (Fig. 1, Fig. 2). For unc-17 and eat-4, we knocked-in a t2a::gfp::h2b (his-44) cassette right before the stop codon of the respective gene. For unc-25, we created two knock-in alleles with the t2a::gfp::h2b (his-44) cassette tagging isoforms a.1/c.1 and b.1 separately. For tdc-1, a gfp::h2b::t2a cassette was knocked into the N-terminus of the locus because of different C-terminal splice variants. The self-cleaving T2A peptide frees up GFP::H2B, which will be transported to the nucleus, thereby facilitating cell identification. For unc-46, unc-47, tph-1, bas-1, tbc-1, cat-1, cat-2, snf-3, and oct-1, we knocked-in a sl2::gfp::h2b cassette at the C-terminus of the locus. Both types of reporter cassettes should capture posttranscriptional, 3’UTR-mediated regulation of each locus, e.g. by miRNAs and RNA-binding proteins (not captured by CeNGEN scRNA data). Since the reporter is targeted to the nucleus, this strategy circumvents shortcomings associated with interpreting antibody staining patterns or dealing with too densely packed cytosolic signals. For mod-5, we analyzed an existing reporter allele (Maicas et al. 2021). For all our neuronal cell identification, we utilized the neuronal landmark strain NeuroPAL (Tekieli et al. 2021; Yemini et al. 2021).

Expression of a reporter allele of eat-4/VGLUT, a marker for glutamatergic identity, in the hermaphrodite

37 of the 38 previously reported neuron classes that express an eat-4 fosmid-based reporter (Serrano-Saiz et al. 2013) showed eat-4 transcripts in the CeNGEN scRNA atlas (Taylor et al. 2021) at all 4 thresholds of stringency, and 1/38 (PVD neuron) showed it in 3 out of the 4 threshold levels (Fig. 1B, Table S1). However, scRNA transcripts were detected at all 4 threshold levels in three additional neuron classes, RIC, PVN, and DVA, for which no previous reporter data provided support. In a recent publication, we had already described that the eat-4 reporter allele syb4257 is expressed in RIC (Reilly et al. 2022)(confirmed in Fig. 4A). We now also confirm expression of this reporter allele, albeit at low levels, in DVA and PVN (Fig. 4B, Table S2).

Summary of neurotransmitter usage and atlases.

Refer to Table 1, Table 2, and Supplementary Tables S2 and S3 for individual gene expression, rationale for neurotransmitter assignments, and more detailed notes.

(A) ACh = acetylcholine; Glu = glutamate; DA = dopamine; 5-HT = 5-hydroxytryptamine, or serotonin; 5-HTP = 5-hydroxytryptophan; PEOH? = the neuron has the potential to use β-hydroxyphenethylamine, or phenylethanolamine; bas-1-depen-unknown MA? = the neuron has the potential to use bas-1-dependent unknown monoamines (histamine, tryptamine, PEA; also see Fig. S1); unknown MA? = the neuron has the potential to use non-canonical monoamines; (up) = neurotransmitter uptake; (syn) = neurotransmitter synthesis; * = dim and variable expression of respective identity gene(s) is detected. Variability could be due to one of the following reasons: (1) the gene is expressed in some but not all animals; (2) the gene is expressed in every animal but the level of expression is below detection threshold in some. Variability is detected only at low fluorescent intensity; at higher intensities, expression remains consistent. Results for anti-GABA staining in SMD and anti-5-HT staining in VC4, VC5, I5, URX, and PVW (male) are variable based on previous reports (see text for citations).

(B) Information from A shown in the context of neuron positions in worm schematics. Note "unknown monoamine" here includes both "bas-1-dependent-unknown MA" and "unknown MA" in A. Neurons marked with "u" can uptake given neurotransmitters but not exclusively; some may also synthesize them, e.g., ADF can both synthesize and uptake 5-HT.

Expression of eat-4/VGLUT and unc-17/VAChT reporter alleles in the adult hermaphrodite.

Neuronal expression of eat-4(syb4257) and unc-17(syb4491) was characterized with landmark strain NeuroPAL (otIs696 and otIs669, respectively). Only selected neurons are shown for illustrating updates from previous reports. See Table S2 for a complete list of neurons.

(A) Dim expression of eat-4(syb4257) in head neurons ASK and ADL is consistent with previous fosmid-based reporter expression. RIC expression is consistent with previous observation using the same reporter allele (Reilly et al. 2022). In addition, dim expression is detected in pharyngeal neuron M5 (also in grayscale inset), previously not detected with eat-4 GFP fosmid-based reporter (otIs388) but visible with eat-4 mCherry fosmid-based reporter (otIs518).

(B) Previously uncharacterized eat-4 expression in PDE and DVA neurons is detected with the eat-4(syb4257) reporter allele. Variable expression in PHA is also occasionally detected. No expression is detected in PVQ. Expression in PVN is detected in both sexes but at a much higher level in the male.

(C) In the head, prominent expression of unc-17(syb4491) in RIP and dim expression in AWA and AFD neurons are detected. There is no visible expression in RIB, FLP, or AVJ. Consistent with previous reports, AIM expresses unc-17 only in males and not hermaphrodites. In addition, very dim expression of AVG can be detected occasionally in hermaphrodites (representative image showing an animal with no visible expression) and slightly stronger in males (representative image showing an animal with visible expression). Inset, grayscale image showing dim expression for AWA and AFD and no expression for RIB.

(D) In the tail, PVN expresses unc-17(syb4491) in both sexes, consistent with previous reports. Scale bars, 10 μm in color images in A, C, and D; 5 μm in B and all grayscale images. Quantification in B is done by normalizing fluorescent intensity of eat-4 GFP to that of the blue channel in the NeuroPAL background. Statistics, Mann-Whitney test.

Another neuron found to have some eat-4 transcripts, but only with the two lower threshold sets, is the I6 pharyngeal neuron. Consistent with our previous fosmid-based reporter data, we detected no I6 expression with our eat-4(syb4257) reporter allele. The eat-4 reporter allele also shows expression in the pharyngeal neuron M5, albeit very weakly (Fig. 4A, Table S2), consistent with CeNGEN scRNA data. Weak expression of the eat-4 fosmid-based reporter in ASK and ADL remained weak, but clearly detectable with the eat-4(syb4257) reporter allele (Fig. 4A, Table S2). Extremely dim expression in PHA can be occasionally detected. Whereas the PVQ neuron class display eat-4 scRNA transcripts and was reported to show very dim eat-4 fosmid-based reporter expression, we detected no expression of the eat-4(syb4257) reporter allele in PVQ neurons (Fig. 4B, Table S2). We also did not detect expression of eat-4(syb4257) in the GABAergic AVL and DVB neurons, in which a recent report describes expression using an eat-4 promoter fusion (Li et al. 2023). An absence of eat-4(syb4257) expression in AVL and DVB is also consistent with the absence of scRNA transcripts in these neurons.

A few neurons were found to express eat-4 transcripts by the CeNGEN atlas, but only with lower threshold levels, including, for example, the RMD, PVM, and I4 neurons (Fig. 1B, Table S1). We failed to detect reporter allele expression in RMD or PVM neurons, but occasionally observed very dim expression in I4. Lastly, we identified a novel site of eat-4 expression in the dopaminergic PDE neuron (Fig. 4B, Table S2). While such expression was neither detected with previous reporters nor scRNA transcripts, we detected it very consistently but at relatively low levels.

Expression of a reporter allele of unc-17/VAChT, a marker for cholinergic identity, in the hermaphrodite

41 of previously described 52 neuron classes that show unc-17 fosmid-based reporter expression (Pereira et al. 2015) showed transcripts in the CeNGEN scRNA atlas at 4 out of 4 threshold levels, another 7 neuron classes at 3 out of 4 threshold levels and 1 at the lowest 2 threshold levels (Taylor et al. 2021). Only one neuron class, RIP, displayed scRNA levels at all 4 thresholds, but showed no corresponding unc-17 fosmid-based reporter expression (Fig. 1B, Table S1). Using the unc-17(syb4491) reporter allele (Fig. 1A), we confirmed expression in RIP (Fig. 4C, Table S2). Of the additional neuron classes that show unc-17 expression at the lower stringency transcript detection levels (Fig. 1B, Table S1), we were able to detect unc-17 reporter allele expression only in AWA (Fig. 4C, Table S2).

Conversely, a few neurons display weak expression with previous multicopy, fosmid-based reporter constructs (RIB, AVG, PVN)(Pereira et al. 2015), but show no CeNGEN scRNA support for such expression (Taylor et al. 2021). The unc-17(syb4491) reporter allele confirmed weak but consistent expression in the PVN neurons as well as variable, borderline expression in AVG (Fig. 4C,D). However, we failed to detect unc-17(syb4491) reporter allele expression in the RIB neurons.

We detected a novel site of unc-17 expression, albeit dim, in the glutamatergic AFD neurons (Fig. 4C, Table S2). This expression was not reported with previous fosmid-based reporter or CeNGEN scRNA data. scRNA transcript reads for cha-1/ChAT, the ACh-synthesizing choline acetyltransferase, were also detected in AFD and PVN (Table S1). Taken together, these observations are consistent with the expectation that although available scRNA data capture the majority of gene expression, it may not necessarily have the required depth to detect lowly expressed genes.

Lastly, another notable observation is the lack of any unc-17 reporter expression or CeNGEN scRNA transcripts in the interneuron AVJ, but presence of CeNGEN scRNA transcript reads for cha-1/ChAT (Table S1), which shares exons with the unc-17/VAChT locus (Alfonso et al. 1994). Although no reporter data is available for cha-1/ChAT, such interesting mismatch between available unc-17 and cha-1/ChAT expression data could provide a hint to potential non-vesicular cholinergic transmission in the AVJ neurons in C. elegans, potentially akin to reportedly non-vesicular release of acetylcholine in the visual system of Drosophila (YANG AND KUNES 2004).

Expression of reporter alleles for GABAergic pathway genes in the hermaphrodite

Expression of unc-25/GAD

The most recent analysis of GABAergic neurons identified GABA-synthesizing cells by anti-GABA staining and an SL2-based unc-25/GAD reporter allele that monitors expression of the rate-limiting step of GABA synthesis, generated by CRISPR/Cas9 engineering (Gendrel et al. 2016). The CeNGEN scRNA atlas shows robust support for these assignments at all 4 threshold levels (Fig. 1B, Table S1). unc-25 scRNA signals were detected at several orders of magnitude lower levels in 3 additional neuron classes, but only with the least robust threshold level.

In this study we generated another unc-25/GAD reporter allele, using a t2a::gfp::h2b cassette (ot1372) (Fig. 2). This allele showed the same expression pattern as the previously described SL2-based unc-25(ot867) reporter allele (Fig. 5A, Table S2). This includes a lack of expression in a number of neurons that stain with anti-GABA antibodies (SMD, AVA, AVB, AVJ, ALA, and AVF) and GLR glia, corroborating the notion that these neurons take up GABA from other cells (indeed, a subset of those cells do express the GABA uptake reporter SNF-11; (Gendrel et al. 2016)).

Expression of GABAergic reporter alleles in the adult hermaphrodite.

(A) Expression of the unc-25/GAD reporter allele ot1372 is detected in the head, ventral nerve cord, and tail neurons. The expression pattern of this new T2A-based reporter allele is similar to that of a previously described SL2-based reporter allele, unc-25(ot867)(Gendrel et al. 2016).

(B) Expression of unc-47/VGAT reporter allele syb7566. Left, the expression pattern of the reporter allele largely matches that of a previously described unc-47 mCherry fosmid-based reporter (otIs564) in the head. Right, a close-up view for the characterization of the reporter allele expression with landmark strain NeuroPAL (otIs669). In the head, consistent with previous reports of the unc-47 fosmid-based reporter (otIs564), dim expression of unc-47(syb7566) in SMD, ALA, and very dim and variable expression in IL1 is detected in both sexes, and unc-47(syb7566) is expressed in ADF only in the male and not hermaphrodite. In addition, the reporter allele is also expressed at a very dim level in the pharyngeal neuron I1 (also in inset) whereas no expression is detected in M1. In the tail, consistent with previous reports of the fosmid, sexually dimorphic expression of the unc-47(syb7566) reporter allele is also detected in PDB, AS11, PVN, and PHC only in the male and not the hermaphrodite. In addition, we also detected very dim expression of PLM in both sexes, confirming potential dim expression of the unc-47 mCherry fosmid-based reporter that was only readily visible after anti-mCherry staining in the past (Serrano-Saiz et al. 2017b). Scale bars, 5 μm for insets and 10 μm for all other images.

(C) Expression of unc-46/LAMP reporter allele syb7278 is largely similar to that of the previously described unc-46 mCherry fosmid-based reporter (otIs568). We also observed expression of both the reporter allele and fosmid-based reporter in PVW, PVN, and very dimly in PDA. Scale bars, 10 μm.

We carefully examined potential expression in the AMsh glia, which were reported to generate GABA through unc-25/GAD (Duan et al. 2020; Fernandez-Abascal et al. 2022). We did not detect visible unc-25(ot867) or unc-25(ot1372) reporter allele expression in AMsh, consistent with the failure to directly detect GABA in AMsh through highly sensitive anti-GABA staining (Gendrel et al. 2016). Furthermore, since these reporters do not capture an alternatively spiced isoform b.1, we generated another reporter allele, unc-25(ot1536), to specifically target this isoform. However, we did not observe any discernible fluorescent reporter expression from this allele. Hence, it is unlikely that an alternative isoform could contribute to expression in additional cell types.

Expression of unc-47/VGAT

While promoter-based transgenes for the vesicular transporter for GABA, unc-47/VGAT, had shown expression that precisely match that of unc-25/GAD (Eastman et al. 1999), we had noted in our previous analysis of the GABA system that a fosmid-based reporter showed much broader expression in many additional neuron classes that showed no sign of GABA usage (Gendrel et al. 2016). In several of these neuron classes both the fosmid-based reporter and the CeNGEN scRNA data indicate very robust expression (e.g. AIN, SIA, SDQ), while in many others scRNA transcripts are only evident at looser thresholds and, correspondingly, fosmid-based reporter expression in these cells is often weak (Table S1) (Gendrel et al. 2016). To investigate this matter further, we CRISPR/Cas9-engineered a gfp-based reporter allele for unc-47, syb7566, and first crossed it with an mCherry-based unc-47 fosmid-based reporter (otIs564) as a first-pass assessment for any obvious overlaps and mismatch of expression patterns between the two (Fig. 5B, left side panels). The vast majority of neurons exhibited overlapping expression between syb7566 and otIs564. There were also many notable similarities in the robustness of expression of the fosmid-based reporter and the reporter allele (Table S1). In a few cases where the fosmid-based reporter expression was so dim that it is only detectable via antibody staining against its fluorophore (mCherry) (Gendrel et al. 2016; Serrano-Saiz et al. 2017b), the reporter allele expression was readily visible (Table S1).

The very few mismatches of expression of the fosmid-based reporter and the reporter allele included the pharyngeal neuron M1, which expresses no visible unc-47(syb7566) reporter allele but weak fosmid-based reporter expression, and the pharyngeal neuron I1, which expresses dim syb7566 but no fosmid-based reporter (Fig. 5B, right side panels). Similarly, AVJ shows very dim and variable unc-47(syb7566) reporter allele expression but no fosmid-based reporter expression. Since AVJ stains with anti-GABA antibodies (Gendrel et al. 2016), this neuron classifies as an uptake and recycling neuron. Other neurons previously shown to stain with anti-GABA antibodies and to express the unc-47 fosmid-based reporter (ALA and SMD)(Gendrel et al. 2016) still show expression of the unc-47 reporter allele. In contrast, the anti-GABA-positive AVA and AVB neurons express no unc-47 and therefore possibly operate solely in GABA clearance (see Discussion).

In conclusion, while the reporter allele of unc-47/VGAT, in conjunction with CeNGEN scRNA data, corroborates the notion that unc-47/VGAT is expressed in all GABA-synthesizing and most GABA uptake neurons, there is a substantial number of unc-47-positive neurons that do not show any evidence of GABA presence. This suggests that UNC-47/VGAT may transport another unidentified neurotransmitter (see Discussion)(Gendrel et al. 2016).

Expression of unc-46/LAMP

In all GABAergic neurons, the UNC-47/VGAT protein requires the LAMP-like protein UNC-46 for proper localization (Schuske et al. 2007). A previously analyzed fosmid-based reporter confirmed unc-46/LAMP expression in all “classic” GABAergic neurons (i.e. anti-GABA and unc-25/GAD-positive neurons), but also showed robust expression in GABA- and unc-47-negative neurons, such as RMD (Gendrel et al. 2016). This non-GABAergic neuron expression is confirmed by CeNGEN scRNA data (Taylor et al. 2021)(Table S1). We generated an unc-46/LAMP reporter allele, syb7278, and found its expression to be largely similar to that of the fosmid-based reporter and to the scRNA data (Fig. 5C, Table S1), therefore corroborating the non-GABAergic neuron expression of unc-46/LAMP. We also detected previously unreported expression in the PVW and PVN neurons in both the reporter allele and fosmid-based reporter (Fig. 5C), thereby further corroborating CeNGEN data. In addition, we also detected very dim expression in PDA (Fig. 5C), which shows no scRNA transcript reads (Table S1). With one exception (pharyngeal M2 neuron class), the sites of non-GABAergic neuron expression of unc-46/LAMP expression do not show any overlap with the sites of unc-47/VGAT expression, indicating that these two proteins have functions independent of each other.

Expression of reporter alleles for serotonin biosynthetic enzymes, tph-1/TPH and bas-1/AAAD, in the hermaphrodite

tph-1/TPH and bas-1/AAAD code for enzymes required for serotonin (5-HT) synthesis (Fig. 1A). scRNA transcripts for tph-1 and bas-1 are detected in previously defined serotonergic neurons at all 4 threshold levels (HSN, NSM, ADF) (Fig. 1, Table S1). In addition to these well characterized sites of expression, several of the individual genes show scRNA-based transcripts in a few additional cells: tph-1 at all 4 threshold levels in AFD and MI. Neither of these cells display scRNA transcripts for bas-1/AAAD, the enzyme that metabolizes the TPH-1 product 5-HTP into 5-HT (Fig. 1A). To further investigate these observations, we generated reporter alleles for both tph-1 and bas-1 (Fig. 2). Expression of the tph-1 reporter allele syb6451 confirmed expression in the previously well-described neurons that stained positive for 5-HT, namely NSM, HSN, and ADF, matching CeNGEN data. While expression in AFD (seen at all 4 threshold levels in the CeNGEN scRNA atlas) could not be confirmed with the reporter allele, expression in the pharyngeal MI neurons could be confirmed (Fig. 6A, Table S2).

Expression of tph-1/TPH, bas-1/AAAD, and cat-1/VMAT reporter alleles in the adult hermaphrodite.

(A) Dorsoventral view of a hermaphrodite head and midbody expressing tph-1(syb6451). tph-1 expression is detected robustly in the MI neuron and dimly and variably in VC4 & VC5. Neuron identities for MI and VC4 & VC5 were validated using otIs518[eat-4(fosmid)::sl2::mCherry::h2b] and vsls269[ida-1::mCherry], respectively, as landmarks. Inset, grayscale image highlighting dim expression in VC4.

(B) Neuronal expression of bas-1(syb5923) characterized with the landmark NeuroPAL (otIs669) strain in the head and midbody regions of young adult hermaphrodites. Dorsoventral view of the adult head shows bas-1/AAAD expression in left-right neuron pairs, including previously reported expression in NSM, CEP, ADF, and ADE (HARE AND LOER 2004). Additionally, we observed previously unreported expression in the URB neurons. Non-neuronal bas-1/AAAD expression is detected in other non-neuronal cell types as reported previously (Yu et al. 2023) (also see Fig. 14, S3).

(C) Lateral views of young adult hermaphrodite head and midbody expressing cat-1/VMAT (syb6486). Previously unreported cat-1/VMAT expression is seen in RIR, CAN, AUA, ASI (also in inset), and variably, ALA. Non-neuronal expression of cat-1/VMAT is detected in a single midbody cell in the gonad (also see Fig. S3), marked with an asterisk. Scale bars, 10 μm for all color images; 5 μm for the inset in grayscale.

We detected co-expression of the bas-1 reporter allele, syb5923, with tph-1(syb6451) in NSM, HSN, and ADF, in accordance with previous reporter and scRNA data (Fig. 6B, Table S2). However, bas-1(syb5923) is not co-expressed with tph-1 in MI (Fig. 6A,B), nor is there CeNGEN-transcript evidence for bas-1/AAAD in MI (Fig. 1, Table S1). Hence, TPH-1-synthesized 5-HTP in MI is not metabolized into 5-HT, consistent with the lack of 5-HT-antibody staining in MI (Horvitz et al. 1982; Sze et al. 2000).

We also detected tph-1(syb6451) reporter allele expression in the serotonergic VC4 and VC5 neurons (Fig. 6A, Table S2), consistent with scRNA data (Fig. 1, Table S1) and previous reporter transgene data (Mondal et al. 2018). This suggests that these neurons are capable of producing 5-HTP. However, there is no bas-1(syb5923) expression in VC4 or VC5, consistent with previous data showing that serotonin is taken up, but not synthesized by them (Duerr et al. 2001) (more below on monoamine uptake; Table 1, 2).

As expected from the role of bas-1/AAAD in dopamine synthesis (HARE AND LOER 2004), bas-1(syb5923) is also expressed in dopaminergic neurons PDE, CEP, and ADE. In addition, it is also expressed weakly in URB, consistent with scRNA data. We did not detect visible expression in PVW or PVT, both of which showed very low levels of scRNA transcripts (Fig. 1, Table S1). Expression of bas-1/AAAD in URB may suggest that URB generates a non-canonical monoamine (e.g. tryptamine, phenylethylamine, or histamine), but since URB expresses no vesicular transporter (cat-1/VMAT, see below), we consider it unlikely that any such monoamine would be secreted via canonical vesicular synaptic release mechanisms.

Expression of a reporter allele of cat-2/TH, a dopaminergic marker, in the hermaphrodite

The CeNGEN scRNA atlas shows transcripts for the rate-limiting enzyme of dopamine synthesis encoded by cat-2/TH (Fig. 1B, Table S1) at all 4 threshold levels in all 3 previously described dopaminergic neuron classes in the hermaphrodite, ADE, PDE, and CEP (Sulston et al. 1975; Sulston et al. 1980; LINTS AND EMMONS 1999). At lower threshold levels, transcripts can also be detected in the OLL neurons. A CRISPR/Cas9-engineered reporter allele for cat-2/TH, syb8255, confirmed expression in ADE, PDE and CEP in adult hermaphrodites (Fig. 7A, Table S2). As expected and described above, all three neuron classes also expressed bas-1/AAAD (Fig. 6B) and cat-1/VMAT (Fig. 6C, see below) (Table S2). We did not detect visible expression of cat-2(syb8255) in OLL. The OLL neurons also display no scRNA transcripts (nor reporter allele expression) for bas-1/AAAD or cat-1/VMAT. No additional sites of expression of cat-2(syb8255) were detected in the adult hermaphrodite.

Expression of cat-2/TH, tdc-1/TDC, and tbh-1/TBH reporter alleles in the adult hermaphrodite.

Neuronal expression was characterized with landmark strain NeuroPAL (otIs669). Lateral views of young adult hermaphrodites expressing reporter alleles for (A) cat-2(syb8255), (B) tbh-1(syb7786), and (C) tdc-1(syb7768). (A) cat-2/TH expression in CEPs, ADE, and PDE match previously reported dopamine straining expression (Sulston et al. 1975).

(B) and (C) Head areas are shown, no neuronal expression was detected in other areas. tdc-1 expression matches previous analysis (Alkema et al. 2005). We detected previously unreported expression of tbh-1 in all six IL2 neurons at low levels. Scale bars, 10 μm.

Expression of reporter alleles of tdc-1/TDC and tbh-1/TBH, markers for tyraminergic and octopaminergic neurons, in the hermaphrodite

The invertebrate analogs of adrenaline and noradrenaline, tyramine and octopamine, are generated by tdc-1 and tbh-1 (Fig. 1A)(Alkema et al. 2005). Previous work had identified expression of tdc-1 in the hermaphrodite RIM and RIC neurons and tbh-1 in the RIC neurons (Alkema et al. 2005). Transcripts in the CeNGEN atlas match those sites of expression for both tdc-1 (scRNA at 4 threshold levels in RIM and RIC neurons) and tbh-1 (scRNA at 4 threshold levels in RIC neurons) (Fig. 1B, Table S1). Much lower transcript levels are present in a few additional, non-overlapping neurons (Fig. 1B). CRISPR/Cas9-engineered reporter alleles confirmed tdc-1 expression in RIM and RIC and tbh-1 expression in RIC (Fig. 7B,C, Table S2). In addition, we also detected dim expression of tbh-1(syb7786) in all six IL2 neurons, corroborating scRNA transcript data (Fig. 7C, Table S2). However, IL2 neurons do not exhibit expression of the reporter allele of tdc-1, which acts upstream of tbh-1 in the octopamine synthesis pathway, or that of cat-1/VMAT, the vesicular transporter for octopamine (Fig. 6C, see below). Hence, the IL2 neurons are unlikely to produce or use octopamine as neurotransmitter, but they may synthesize another monoaminergic signal (Table 2).

Expression of a reporter allele of cat-1/VMAT, a marker for monoaminergic identity, in the hermaphrodite

As the vesicular monoamine transporter, cat-1/VMAT is expected to be expressed in all neurons that synthesize serotonin, dopamine, tyramine, and octopamine (Fig. 1A). Both scRNA data and a CRISPR/Cas9-engineered reporter allele, syb6486, confirm expression in all these cells (Fig. 6C, Table S2). In addition, based on antibody staining and previous fosmid-based reporters, cat-1/VMAT is known to be expressed in neurons that are serotonin-positive (VC4, VC5, and RIH) (Duerr et al. 1999; Duerr et al. 2001; Serrano-Saiz et al. 2017b). Again, both scRNA data, as well as a CRISPR/Cas9-engineered reporter allele, syb6486, confirm expression in these cells (Fig. 6C, Table S2).

In addition to these canonical monoaminergic neurons, the CeNGEN scRNA data shows cat-1/VMAT expression at all 4 threshold levels in RIR, CAN, AVM and, at much lower threshold, 8 additional neuron classes (Fig. 1B, Table S1). Our cat-1/VMAT reporter allele, syb6486, corroborates expression in RIR and CAN, but not in AVM (Fig. 6C, Table S2). We also observed expression of the cat-1 reporter allele in two of the neuron classes with scRNA transcripts at the lowest threshold level, ASI and variably, AVL (Fig. 6C, Table S1). Interestingly, AVL does not express any other monoaminergic pathway genes (Table S2), therefore it may be transporting a new amine yet to be discovered. This scenario also applies for two male-specific neurons (more below). As previously mentioned, we detected no cat-1/VMAT expression in the tph-1-positive MI or the cat-2/TH-positive OLL neurons.

The cat-1/VMAT reporter allele revealed expression in an additional neuron class, the AUA neuron pair (Fig. 6C, Table S2). Expression in this neuron is not detected in scRNA data; such expression may be consistent with previous CAT-1/VMAT antibody staining data (Duerr et al. 1999). These authors found the same expression pattern as we detected with cat-1/VMAT reporter allele, except for the AIM neuron, which Duerr et al. identified as CAT-1/VMAT antibody-staining positive. However, neither our reporter allele, nor a fosmid-based cat-1/VMAT reporter, nor scRNA data showed expression in AIM, and we therefore think that the neurons identified by Duerr et al as AIM may have been the AUA neurons instead (Serrano-Saiz et al. 2017b). Additionally, a cat-1-positive neuron pair in the ventral ganglion, unidentified but mentioned by Duerr and colleagues (Duerr et al. 1999), is likely the tyraminergic RIM neuron pair, based on our reporter allele and CeNGEN scRNA data.

Expression of reporter alleles of monoamine uptake transporters in the hermaphrodite

In addition to or in lieu of synthesizing monoamines, neurons can uptake them from their surroundings. To investigate the cellular sites of monoamine uptake in more detail, we analyzed fluorescent protein expression from engineered reporter alleles for the uptake transporters of 5-HT (mod-5/SERT(vlc47)), the predicted uptake transporter for tyramine (oct-1/OCT(syb8870)), and that for betaine (snf-3/BGT1(syb7290)).

Serotonin/5-HT uptake

Using a promoter-based transgene and antibody staining, previous work had shown expression of the serotonin uptake transporter mod-5 in NSM, ADF, RIH, and AIM (Jafari et al. 2011; Maicas et al. 2021). This matched the observations that RIH and AIM do not synthesize 5-HT (i.e. do not express tph-1), but stain positive with a 5-HT antibody (Jafari et al. 2011). In mod-5 mutants or wildtype worms treated with serotonin reuptake inhibitors (such as the SSRI fluoxetine), RIH and AIM lose 5-HT immunoreactivity (Jafari et al. 2011). We analyzed a CRISPR-based reporter allele, mod-5(vlc47)(MAICAS et al. 2021), and confirmed expression in the four neuron classes NSM, ADF, RIH, and AIM (Fig. 8). Because only NSM, ADF, and RIH, but not AIM, express the reporter allele of the monoamine transporter CAT-1/VMAT (Fig. 6), we agree with previous studies that AIM likely functions as a serotonin uptake/clearance neuron (Table 1, 2; see also Discussion). In addition, we also detected dim expression in the phasmid neuron class PHA and very dim, variable signal in URX (Fig. 12A,B,E), consistent with scRNA data (Table S1). The results for anti-5-HT-staining from previous reports are variable in a few neurons, possibly due to differences in staining methods (including URX, I5, VC4, VC5 and PVW (LOER AND KENYON 1993; RAND AND NONET 1997; Duerr et al. 1999; Serrano-Saiz et al. 2017b). In light of its mod-5 reporter expression, URX may acquire 5-HT via mod-5, akin to AIM (Table 1, 2).

Expression of mod-5/SERT and oct-1/OCT reporter alleles in adult animals.

Neuronal expression was characterized with landmark strain NeuroPAL (otIs669) and DiD-filling.

(A,C) In adult hermaphrodites, mod-5(vlc47) is expressed in sex-shared neurons NSM, ADF, RIH, AIM, consistent with previous reports(Jafari et al. 2011; Maicas et al. 2021). In addition, we also observed expression in the phasmid neuron PHA and dim and variable expression in URX. There is no visible expression in the ventral nerve cord (VNC).

(B,D) In adult males, mod-5(vlc47) is visibly expressed in NSM, RIH, AIM, as well as the male-specific neurons CEM, PGA, R3B, R9B, and CP1 to CP6. Expression in ADF is often not detected (see F).

(E) DiD-filling confirms mod-5(vlc47) expression in phasmid neuron class PHA, and not PHB, in young adults in both sexes (L4 male image is to facilitate neuron ID in adults, because the positions of the two neuron classes can change in males during the L4 to adult transition).

(F) Expression of mod-5(vlc47) in ADF is stronger in hermaphrodites than in males. In comparison, expression is not sexually dimorphic for the reporter alleles of either the 5-HT-synthsizing enzyme tph-1 or the vesicular acetylcholine transporter unc-17.

(G) In the tail region of wild type males, male-specific neurons PGA, R1B, R3B, and R9B are stained positive for 5-HT. In a mod-5(n3314) mutant background, staining is completely lost in PGA (41/41 stained animals) and significantly affected for R9B (completely lost in 31/41 animals and much dimmer in the rest), while it remains in all 41 stained animals for R1B and R3B. The staining for CP1 to CP6 are also not affected in mod-5 mutant animals (remaining in 41/41 stained animals; image showing CP5 and CP6).

(H,I) In adult animals, oct-1(syb8870) is expressed in the tyraminergic neuron class RIM in both sexes. Expression is not observed in any other neurons.

(J,K) Outside the nervous system, oct-1(syb8870) is expressed in body wall muscle (BWM) throughout the worm (J) as well as hypodermal cells and selected head glia (K). Expression is also observed in gonadal cells in the male vas deferens (K). A pan-glial reporter otIs870[mir-228p::3xnls::TagRFP] and a dpy-7p::mCherry reporter stIs10166 [dpy-7p::his-24::mCherry + unc-119(+)] were used for glial and hypodermal identification, respectively.

Scale bars, 10 μm.

In the hermaphrodite-specific neurons HSN, VC4, and VC5, we did not observe expression of the mod-5 reporter allele (Table 1, 2). Since VC4 and VC5 do not express the complete synthesis pathway for 5-HT, we infer that the anti-5-HT staining in these neurons is a result of alternative 5-HT uptake or synthesis mechanisms. A similar scenario holds for the pharyngeal neuron I5 which was previously reported to stain weakly for 5-HT (Serrano-Saiz et al. 2017b).

Tyramine uptake

Biochemical studies in vertebrates have shown that the SLC22A1/2/3 (aka OCT-1/2/3) organic anion transporter can uptake monoaminergic neurotransmitters (NIGAM 2018), with SLC22A2 being apparently selective for tyramine (Berry et al. 2016). oct-1 is the ortholog of the OCT subclass of SLC22 family members (Zhu et al. 2015), but neither its expression nor function in the nervous system had been previously reported. We tagged the endogenous oct-1 locus with an sl2::gfp::h2b cassette (syb8870) and, within the nervous system, observed exclusive expression in the RIM neuron (Fig. 8H,I), indicating that RIM is likely capable of uptaking tyramine in addition to synthesizing it via tdc-1/TDC. This is consistent with RIM being the only neuron showing oct-1 scRNA transcripts at all 4 threshold levels in the CeNGEN atlas (Table S1).

Betaine uptake

Notably, four CAT-1/VMAT- expressing neuron classes, CAN, AUA, RIR, and ASI do not express biosynthetic enzymes for synthesis of the four conventional monoaminergic transmitters known to be employed in C. elegans (5-HT, dopamine, octopamine, or tyramine). Hence, these neuron classes might instead uptake some of these transmitters. We considered the putative neurotransmitter betaine as a possible candidate, since CAT-1/VMAT is also able to package betaine (Peden et al. 2013; Hardege et al. 2022). Betaine is synthesized endogenously, within the nervous system mostly in the cat-1/VMAT- positive RIM neuron (Hardege et al. 2022), but it is also available in the bacterial diet of C. elegans (Peden et al. 2013). In vertebrates, dietary betaine is taken up by the betaine transporter BGT1 (i.e. SLC6A12). To test whether cat-1/VMAT-positive neurons may acquire betaine via BGT1-mediated uptake, we CRISPR/Cas9-engineered a reporter allele for snf- 3/BGT1, syb7290. We detected expression in the betaine-synthesizing (and also tyraminergic) RIM neuron (Fig. 9, Table 1, 2). In addition, snf-3 is indeed expressed in all the four cat-1/VMAT-positive neuron classes that do not synthesize a previously known monoaminergic transmitter (CAN, AUA, and variably, RIR and ASI)(Fig. 9A,B). These neurons may therefore take up betaine and synaptically release it via CAT-1/VMAT. The snf-3(syb7290) reporter allele is also expressed in the serotonergic neuron NSM (albeit variably) (Table 1, 2), thus NSM could also be a betaine uptake neuron. In addition, we also detected snf-3(syb7290) expression in several other neurons that do not express cat-1(syb6486) (Table S1). Expression was also observed in a substantial number of non-neuronal cell types (Fig. 9E-G, Table 2, S1). These neurons and non-neuronal cells may serve to clear betaine (see Discussion, Neurotransmitter synthesis versus uptake). snf-3(syb7290) is not expressed in the inner and outer labial neuron classes as previously suggested (Peden et al. 2013); these cells were likely misidentified in the previous study and are in fact inner and outer labial glial cells (as discussed further below).

Expression of snf-3/BGT1/SLC6A12 in adult animals.

Neuronal expression was characterized with landmark strain NeuroPAL (otIs669) and DiD-filling.

(A,B) In the adult hermaphrodite, neuronal expression of snf-3(syb7290) is detected in cat-1/VMAT-positive neurons AUA, CAN, and dimly and variably, RIR and ASI (confirmed with DiD-filling). In addition, it is also expressed in cat-1/VMAT-negative neurons AIB, RIM, RMH, SMD, VA12, DA9, PDA, PHC, PVN as labeled, as well as more neurons listed in Table S1. In the midbody, expression is not in PDE (dopaminergic, cat-1-positive) but in its associated glial cells. It is also detected in multiple vulval support cells (B) and some epithelial cells near the somatic gonad.

(C) In the adult male, in addition to its expression in sex-shared neurons as in hermaphrodites, snf-3(syb7290) is also expressed in male-specific neuron class PDC, as well as in PHD and variably in PVV.

(D) Similarly to its expression in hermaphrodites, snf-3(syb7290) is detected in CAN and PDE-associated glial cells, but not PDE neurons, in males.

(E) In the male tail, snf-3(syb7290) is expressed in a number of glial cells including the spicule sockets and/or sheath cells (dorsal view). It is also detected in the somatic gonad (ventral view).

(F) snf-3(syb7290) is broadly expressed in most if not all glia in both sexes. Glial cell type is determined by cell location and the appearance of their nuclei in Normarski. To confirm they are not neurons, a pan-neuronal marker (UPN, or “uber pan-neuronal”, a component in NeuroPAL) is used to determine non-overlapping signals between the two reporters. Head expression in the male is very similar to that in the hermaphrodite and thus not shown. (G) snf-3(syb7290) is broadly expressed in hypodermal and seam cells in both sexes. Scale bars, 10 μm. Asterisks, non-neuronal expression.

Together with the expression pattern of the uptake transporters, all cat-1/VMAT-positive neurons in the hermaphrodite can be matched with an aminergic neurotransmitter. We nevertheless wondered whether another presently unknown monoaminergic transmitter, e.g., histamine or other trace amine, could be synthesized by a previously uncharacterized AAAD enzyme encoded in the C. elegans genome, hdl-1 (Fig. S1A)(HARE AND LOER 2004). We CRISPR/Cas9-engineered an hdl-1 reporter allele, syb1048, but detected no expression of this reporter in the animal (Fig. S1C,D). Attempts to amplify weak expression signals by insertion of Cre recombinase into the locus failed [hdl-1(syb4208)](see Methods). CeNGEN scRNA data also shows no strong transcript expression in the hermaphrodite nervous system and only detected notable expression in sperm (Taylor et al. 2021).

Reporter alleles and NeuroPAL-facilitated neuron class-identification reveal novel expression patterns of neurotransmitters in the male-specific nervous system

No comprehensive scRNA atlas has yet been reported for the nervous system of the male. Based on the expression of fosmid-based reporters, we had previously assembled a neurotransmitter atlas of the C. elegans male nervous system in which individual neuron classes are notoriously difficult to identity (Serrano-Saiz et al. 2017b). We have since established a NeuroPAL landmark strain that permits more reliable identification of gene expression patterns in both the hermaphrodite and male-specific nervous system (Tekieli et al. 2021; Yemini et al. 2021). We used NeuroPAL to facilitate the analysis of the expression profiles of our CRISPR/Cas9-engineered reporter alleles in the male, resulting in updated expression profiles for 11 of the 16 reporter alleles analyzed. As in the hermaphrodite, reasons for the updates vary. In addition to the improved accuracy of neuron identification provided by NeuroPAL, in some cases there are true differences of expression patterns between the fosmid-based reporters and reporter alleles. We elaborate on these updates for individual reporter alleles below.

Expression of reporter alleles of Glu/ACh/GABA markers in the male-specific nervous system

We analyzed eat-4/VGLUT (syb4257), unc-17/VAChT (syb4491), unc-25/GAD (ot1372), and unc-47/VGAT (syb7566) expression in the male-specific nervous system using NeuroPAL landmark strains (otIs696 for eat-4 and otIs669 for all others). Of all those reporter alleles, unc-25/GAD (ot1372) was the only one with no updated expression. Specifically, in addition to confirming presence of expression of the unc-25(ot1372) reporter allele in CP9, EF1/2, EF3/4, we also confirmed its lack of expression in anti-GABA-positive neurons R2A, R6A, and R9B (Gendrel et al. 2016; Serrano-Saiz et al. 2017b)(Fig. 11A, Table S3).

In the preanal ganglion, we observed weak expression of unc-17(syb4491) in DX3/4 (Fig. 10B, Table S3), hence assigning previously unknown neurotransmitter identity to these neurons. Related to DX3/4, we also confirmed expression of unc-17 in DX1/2 in the dorsorectal ganglion, consistent with fosmid-based reporter data (Table S3) (Serrano-Saiz et al. 2017b). In the lumbar ganglion, we detected novel expression of unc-17(syb4491) in 5 pairs of type B ray neurons, namely R1B, R4B, R5B, R7B, and R9B (Fig. 10B, Table S3). Expression in all these neurons is low, possibly explaining why it is not observed with an unc-17 fosmid-based reporter (Serrano-Saiz et al. 2017b).

Expression of eat-4/VGLUT and unc-17/VAChT reporter alleles in the adult male.

Neuronal expression of eat-4(syb4257) and unc-17(syb4491) was characterized with landmark strain NeuroPAL (otIs696 and otIs669, respectively). Only selected neurons are shown to illustrate updates from previous studies. See Table S3 for a complete list of neurons.

(A) eat-4(syb4257) expression. Top, long panels: CA1, CA2, and CA3 show visible, albeit very dim, novel expression of eat-4 (also expressed in CA4). Bottom panels: CA7 strongly expresses eat-4(syb4257), whereas CP7 does not. Neuron IDs for these two neurons were previously switched (Serrano-Saiz et al. 2017b).

(B) unc-17(syb4491) expression. Top, long panels: ventral view of a male ventral nerve cord showing high levels of expression in CA1, CA2, and CA3 and previously unreported low levels of expression in CP1, CP2, and CP3. Middle panels: low levels of expression in CA7 and CP7. There is no visible expression in CP9. Bottom panels: lateral view of a male tail showing previously unreported dim expression in R1B, R4B, R5B, R7B, and R9B; ventral view of the pre-anal ganglion showing expression in DX3/4. Scale bars, 10 μm.

(C) The updated neurotransmitter atlas underscores the molecular diversity of the male-specific ventral cord neuron class CA and CP. Based on their expression patterns for neurotransmitter genes, these neurons can be potentially grouped into the following 4 CA and 5 CP sub-classes.

See Table S3 and Figs. 10-12 for all genes mentioned in the following. CAs: (1) CA1 to CA4: express unc-17, unc-47, very weak eat-4; (2) CA5, 6, 8: express unc-17 and unc-47; (3) CA7: expresses unc-17, strong eat-4, and no unc-47; (4) CA9: expresses weak unc-17. CPs: (1) CP0: only expresses weak eat-4; (2) CP1 to CP4: express unc-17, unc-47, cat-1, tph-1, bas-1, and stain for 5-HT; (3) CP5, 6: express eat-4, unc-47, cat-1, tph-1, bas-1, and stain for 5-HT; (4) CP7, 8: express very weak unc-17 and unc-47; (5) CP9: expresses strong unc-25 and unc- 47, and stain for GABA. The proposed sub-classification for these neuron classes indicates different functions of individual CA and CP neurons; it also provides useful resources for ongoing single-cell sequencing efforts for male-specific neurons.

Expression of GABAergic reporter alleles in the adult male.

Neuronal expression of unc-25(ot1372) and unc-47(syb7566) reporter alleles was characterized with landmark strain NeuroPAL (otIs669). Only selected neurons are shown to illustrate updates from previous reports. See Table S3 for a complete list of neurons.

(A) unc-25(ot1372) is expressed in male-specific CP9 and EF neurons as well as a few sex-shared neurons, all consistent with previous reports (Gendrel et al. 2016; Serrano-Saiz et al. 2017b).

(B) unc-47(syb7566) shows expression in male head neuron classes MCM and CEM, the former previously undetected and the latter consistent with fosmid-based reporter otIs564.

(C) unc-47(syb7566) shows expression in a number of ventral cord CA and CP neurons, largely consistent with reported otIs564 fosmid-based reporter expression except for no visible expression of syb7566 in CA7 (due to its initial confusion with CP7, described in Fig. 10) and presence of very dim expression in CP7. The syb7566 reporter allele is also not visible in CA9. Scale bars, 10 μm.

In the ventral nerve cord, we found additional, very weak expression of eat-4(syb4257) in CA1 to CA4 (Fig. 10A, Table S3), as well as weak expression of unc-17(syb4491) in CP1 to CP4 (Fig. 10B, Table S3), all undetected by previous analysis of fosmid-based reporters (Serrano-Saiz et al. 2017b). Conversely, two neurons lack previously reported expression of fosmid-based reporters; CP9 does not show visible unc-17(syb4491) expression (Fig. 10B) and neither does CA9 show visible expression of unc-47(syb7566) expression (Fig. 11C). We also realized that the neuron identifications of CA7 and CP7 were previously switched (Serrano-Saiz et al. 2017b), due to lack of proper markers for those two neurons. With NeuroPAL, we are now able to clearly distinguish the two and update their classic neurotransmitter reporter expression: CA7 expresses high levels of eat-4(syb4257) (Fig. 10A, Table S3), very low levels of unc-17(syb4491) (Fig. 10B), and no unc-47(syb7566) (Fig. 10C); CP7 expresses no eat-4(syb4257) (Fig. 10A, Table S3), very low levels of unc-17(syb4491) (Fig. 8B), and very low levels of unc-47(syb7566) as well (Fig. 11C). Taken together, the analysis of reporter alleles reveals a remarkable diversity of CA and CP neurons, summarized in Fig. 8C.

In the head, we detected expression of unc-47(syb7566) in the male-specific neuron class MCM (Fig. 11B, Table S3), previously not observed with fosmid-based reporters. Consistent with fosmid-based reporter data, the other male-specific neuron class, CEM, shows expression of unc-17(syb4491) (Table S3) and unc-47(syb7566) (Fig. 11B, Table S3) reporter alleles.

Expression of reporter alleles for monoaminergic neurotransmitter pathway genes in the male-specific nervous system

We analyzed the expression of reporter alleles for the following genes involved in monoamine biosynthesis and uptake in the male-specific nervous system: cat-1/VMAT (syb6486), tph-1/TPH (syb6451), cat-2/TH (syb8255), bas-1/AAAD (syb5923), tdc-1/TDC (syb7768), tbh-1/TBH (syb7786), mod-5/SERT (vlc47), oct-1/OCT (syb8870), and snf-3/BGT1 (syb7290). As in the hermaphrodite nervous system, we used the NeuroPAL reporter landmark (otIs669) for neuron ID (Tekieli et al. 2021). We found novel expression patterns in all male-specific ganglia (Fig. 12, 13, Table S3).

Expression of the cat-1/VMAT, tph-1/TPH, and bas-1/AAAD reporter alleles in the adult male.

Neuronal expression was characterized with landmark strain NeuroPAL (otIs669).

(A) Novel cat-1(syb6486) expression is seen in male-specific neurons PDC, PVY, PVX, R2A, and R4B. Consistent with previous reports, it is also expressed in HOA, PGA, R5A, R7A, R9A, R1B, and R8B. Its expression in ventral cord neurons CP1 to CP6 is consistent with earlier studies.

(B) tph-1(syb6451) is expressed in male-specific head neuron class CEM and sex-shared neurons ADF, NSM, and MI. Similar to its expression in hermaphrodites, tph-1 in MI was previously undetected. In the tail, in addition to previously determined expression in R1B, R3B, and R9B, tph-1(syb6451) is also expressed at very low levels in R4B and R7B. Ventral cord expression of tph-1(syb6451) in CP1 to CP6 is consistent with previous reports and thus not shown here.

(C) bas-1(syb5923) is expressed in previously identified NSM, ADE, PDE, and CEP neurons. In addition, we detected weak expression in URB as in the hermaphrodite. We also updated bas-1/AAAD expression in 39 male-specific neurons and 1 more with variable expression (see Table S3 for complete list). Neurons are also shown in grayscale for clearer visualization in some cases. Scale bars, 10 μm. Asterisks, non-neuronal expression, also see Fig. 14 and Fig. S3.

Serotonin/5-HT synthesis

Serotonergic identity had been assigned to several male-specific neurons before (CP1 to CP6, R1B, R3B, R9B)(LOER AND KENYON 1993), and we validated these assignments with our reporter alleles (Fig. 12, Table S3). In addition, we detected previously unreported expression of tph-1 (Fig. 12B) in the male-specific head neuron class CEM, as well as in a subset of B-type ray sensory neurons, R4B and R7B. However, not all of the neurons display additional, canonical serotonergic neuron features: While R4B and R7B express bas-1(syb5923) (with R4B expressing it variably) to generate 5-HT, neither neuron was detected by anti-5-HT staining in the past. On the other hand, R9B and CEM stain positive for 5-HT (Serrano-Saiz et al. 2017b), but they do not express bas-1(syb5923), indicating that they may be producing 5-HTP rather than 5-HT (see more below on serotonin uptake). In addition, R4B and R9B, but not R7B or CEM, express cat-1(syb6486) for vesicular release of 5-HT.

In the ventral nerve cord, consistent with previous fosmid-based reporter data (Serrano-Saiz et al. 2017b), we observed the expression of cat-1(syb6486) and tph-1(syb6451) in CP1 to CP6 (Fig. 12A,B; Table S3). Additionally, we also detected novel expression of bas-1(syb5923) in CP1 to CP4 and strongly in CP5 and CP6 (Fig. 12C, Table S3). This updated expression supports the serotonergic identities of these neurons, which had been determined previously based only on their expression of cat-1/VMAT reporters and positive staining for 5-HT (LOER AND KENYON 1993; Serrano-Saiz et al. 2017b).

Dopamine synthesis

We found that the expression of the dopamine-synthesizing cat-2(syb8255) reporter allele precisely matched previous assignments of dopaminergic identity (Sulston et al. 1975; Sulston et al. 1980; LINTS AND EMMONS 1999), i.e. expression was detected exclusively in R5A, R7A, and R9A (Fig. 13A, Table S3), in addition to all sex-shared dopaminergic neurons. All these neurons show matching expression of bas-1/AAAD, the other essential enzyme for dopamine synthesis, and cat-1/VMAT, the vesicular transporter for dopamine (Fig. 13A,C; Table S3).

Expression of cat-2/TH, tdc-1/TDC and tbh-1/TBH reporter alleles in the adult male.

Neuronal expression was characterized with landmark strain NeuroPAL (otIs669).

(A) cat-2(syb8255) is expressed in male-specific neurons R4A, R7A, and R9B. This expression, as well as its expression in sex-shared neurons PDE, CEP, and ADE, is consistent with previous reports (Sulston et al. 1975; Sulston et al. 1980; LINTS AND EMMONS 1999).

(B) tdc-1(syb7768) is expressed in sex-shared neurons RIM and RIC and male-specific neurons HOA, R8A, and R8B, all consistent with previous studies (Serrano-Saiz et al. 2017b). We also detected weak expression in R7A.

(C) tbh-1(syb7786) is expressed in RIC, consistent with its previously reported expression in hermaphrodites. As in hermaphrodites, we also detected tbh-1(syb7786) in IL2 neurons of the male. In male-specific neurons, previously unreported expression is detected in CEM, HOB, and all type B ray neurons except for R6B. Intriguingly, this expression pattern resembles that of pkd-2 and lov-1, both genes essential for male mating functions (BARR AND STERNBERG 1999; Barr et al. 2001). Inset, grayscale image showing dim expression for IL2 neurons. Scale bars, 10 μm. Asterisks, non-neuronal expression, also see Fig. 14 and Fig. S3.

Tyramine & Octopamine synthesis

Reporter alleles for the two diagnostic enzymes, tdc- 1/TDC and tbh-1/TBH, confirm the previously reported assignment of HOA as tyraminergic (Serrano-Saiz et al. 2017b), based on the presence of tdc-1(syb7768) but absence of tbh- 1(syb7786) expression (Fig. 11B,C). The tdc-1 reporter allele reveals a novel site of expression in R7A. Due to lack of tbh-1 expression, R7A therefore classifies as another tyraminergic neuron. Both HOA and R7A also co-express cat-1/VMAT for vesicular release of tyramine.

We detected no neurons in addition to the sex-shared RIC neuron class that shares all features of a functional octopaminergic neuron, i.e. co-expression of tbh-1/TBH, tdc-1/TDC, and cat-1/VMAT. While one male-specific neuron, R8B, shows an overlap of expression of tdc- 1(syb7768) and tbh-1(syb7786), indicating that these neurons can synthesize octopamine, R8B does not express cat-1(syb6486), indicating that these neurons cannot engage in vesicular release of octopamine.

Curiously, while there are no other male-specific neurons that co-express tdc-1 and tbh- 1, several male-specific neurons express tbh-1, but not tdc-1 (Fig. 13B,C; Table 2, Table S3). The absence of the TDC-1/AAAD protein, which produces tyramine, the canonical substrate of the TBH-1 enzyme (Fig. 1A), indicates that TBH-1 must be involved in the synthesis of a compound other than octopamine. Moreover, bas-1/AAAD is expressed in several of the tbh- 1(+); tdc-1(-) neurons (R1B, R2B, R3B, R4B, and R7B) (Fig. 12C, Table 2, Table S3). Rather than using L-Dopa or 5-HTP as substrate, BAS-1/AAAD may decarboxylate aromatic amino acids, which then may serve as a substrate for TBH-1. We consider the trace amine phenylethanolamine (PEOH) as a candidate end product (see Discussion).

Other monoaminergic neurons

In the preanal ganglion, we detected novel expression of the cat-1(syb6486) reporter allele in the cholinergic PDC, PVX, and PVY neurons (Fig. 12A). Intriguingly, just as the sex-shared neuron AVL (Fig. 6C), these neurons express no other serotonergic, dopaminergic, tyraminergic, or octopaminergic pathway gene. However, we did find PDC (but not PVX or PVY) to express the betaine uptake transporter reporter allele snf- 3(syb7290) (Fig. 9; more below). PVX and PVY may synthesize or uptake another aminergic transmitter. Such presumptive transmitter is not likely to be synthesized by hdl-1/AAAD since we detected no expression of the hdl-1 reporter allele syb4208 in the male nervous system (Fig. S1C,D).

The expression pattern of the bas-1/AAAD, which had not been previously analyzed in the male-specific nervous system, reveals additional novelties. In addition to the “canonical” serotonergic and dopaminergic neurons described above, we detected bas-1(syb5923) reporter allele expression in a substantial number of additional neurons, including the tyraminergic HOA and R7A neurons, but also the DVE, DVF, R2A, R3A, R6A, R8A, R2B, R6B, R7B, PCB and SPC neurons (Fig. 12C, Table S3). As described above, a subset of the neurons co-express tbh-1(syb7786) (most B-type ray neurons), a few co-express tdc- 1(syb7768) (HOA and several A-type ray neurons), and several co-express neither of these two genes. Only a subset of these neurons express cat-1(syb6486). Taken together, this expression pattern analysis argues for the existence of additional monoaminergic signaling system(s) (Table 2).

Serotonin/5-HT uptake

In the male-specific nervous system, we detected mod-5(vlc47) expression in CEM, PGA, R3B, R9B, and ventral cord neurons CP1 to CP6 (Fig. 8D). We found that anti-5-HT staining in CP1 to CP6, R1B, and R3B is unaffected in mod-5(n3314) mutant animals, consistent with these neuron expressing the complete 5-HT synthesis machinery (i.e. tph-1 and bas-1)(Table 2, Fig. 8B,D,G). Hence, like several other monoaminergic neurons, these serotonergic neurons both express, synaptically release, and re-uptake 5-HT. In contrast, anti-5-HT staining is lost from the R9B and PGA neurons of mod-5(n3314) mutant animals, indicating that the presence of 5-HT in these neurons depends on 5-HT uptake, consistent with them not expressing the complete 5-HT synthesis pathway (Table 2, Fig. 8B,D,G). Since R9B and PGA express cat-1/VMAT, these neurons have the option to utilize 5-HT for synaptic signaling after mod-5-dependent uptake.

Tyramine and betaine uptake

We did not observe oct-1(syb8870) reporter allele expression in male-specific neurons. As in the hermaphrodite nervous system, we detected snf-3(syb7290) in a number of neurons that do not express CAT-1/VMAT (Table S1), including in male-specific neurons PHD, and variably, PVV (Fig. 9C). As mentioned earlier, the male-specific neuron PDC expresses both cat-1(syb6486) and snf-3(syb7290), making it a likely betaine-signaling neuron.

Sexually dimorphic neurotransmitter expression in sex-shared neurons

eat-4/VGLUT

We had previously noted that a fosmid-based eat-4/VGLUT reporter is upregulated in the sex-shared neuron PVN, specifically in males (Serrano-Saiz et al. 2017b). Since PVN is also cholinergic (Fig. 4D)(Pereira et al. 2015), this observation indicates a sexually dimorphic co-transmission configuration. As described above (Fig. 4B, Table S2), our eat-4 reporter allele revealed low levels of eat-4/VGLUT expression in hermaphrodites PVN, but in males the eat-4 reporter alleles shows strongly increased expression, compared to hermaphrodites. Hence, rather than being an “on” vs. “off” dimorphism, dimorphic eat-4/VGLUT expression in male PVN resembles the “scaling” phenomenon we had described previously for eat-4/VGLUT in male PHC neurons, compared to hermaphrodite PHC neurons (Serrano-Saiz et al. 2017a). Both PHC and PVN display a substantial increase in the amount of synaptic output of these neurons in males compared to hermaphrodites (Cook et al. 2019), providing a likely explanation for such scaling of gene expression. The scaling of eat-4/VGLUT expression in PVN is not accompanied by scaling of unc-17/VAChT expression, which remains comparable in both sexes (Fig. 4D).

We also examined AIM, another neuron class that was previously reported to be sexually dimorphic in that AIM expresses eat-4/VGLUT fosmid-based reporters in juvenile stages in both sexes, whereas upon sexual maturation its neurotransmitter identity is switched from being glutamatergic to cholinergic only in adult males and not hermaphrodites (Pereira et al. 2015; Pereira et al. 2019). With the eat-4(syb4257) reporter allele, we also detected a downregulation of eat-4 expression to low levels in young adult males and almost complete elimination in 2-day-old adult males, while expression in hermaphrodites stays high.

unc-17/VAChT

The unc-17/VAChT reporter allele syb4491 confirms that cholinergic identity is indeed male-specifically turned on in the AIM neurons (Fig. 4C), thereby confirming the previously reported neurotransmitter switch (Pereira et al. 2015). The fosmid-based unc- 17 reporter also showed sexually dimorphic expression in the AVG neurons (Serrano-Saiz et al. 2017b). This is also confirmed with the unc-17 reporter allele, which shows dim and variable expression in hermaphrodites and slightly stronger, albeit still dim, AVG expression in males (Fig. 4C, showing a hermaphrodite representing animals with no visible expression and a male with representative dim expression).

unc-47/VGAT

unc-47(syb7566) confirms previously reported sexually dimorphic expression of unc-47/VGAT in several sex-shared neurons, including ADF, PDB, PVN, PHC, AS10, and AS11 (Fig. 5B, right side panels) (Serrano-Saiz et al. 2017b). The assignment of AS10 was not definitive in our last report (we had considered either DA7 or AS10), but with the help of NeuroPAL the AS10 assignment could be clarified. In all these cases expression was only detected in males and not hermaphrodites. It is worth mentioning that expression of the mCherry-based unc-47/VGAT fosmid-based reporter (otIs564) in some of these neurons was so dim that it could only be detected through immunostaining against the mCherry fluorophore and not readily visible with the fosmid-based reporter by itself (Serrano-Saiz et al. 2017b). In contrast, the unc-47/VGAT reporter allele is detected in all cases except the PQR neuron class. In addition, we also detected dim unc-47/VGAT expression in the PLM neurons in both sexes (Fig. 5B).

mod-5/SERT

Expression of the mod-5(vlc47) reporter allele is sexually dimorphic in the pheromone-sensing ADF neurons, with higher levels in hermaphrodites compared to males (Fig. 8F). Notably, the serotonin-synthesizing enzyme (tph-1) and vesicular acetylcholine transporter (unc-17) do not exhibit this dimorphism in ADF (Fig. 8F). This suggests that the sex difference specifically involves serotonin signaling mechanisms, particularly serotonin uptake rather than synthesis.

We had previously reported that the PVW neuron stains with anti-5-HT antibodies exclusively in males but did not detect expression of a fosmid-based reporter for the serotonin-synthesizing enzyme TPH-1 (Serrano-Saiz et al. 2017b). We confirmed the lack of tph-1 expression with our new tph-1 reporter allele in both males and hermaphrodites, and also found that hermaphrodite and male PVW does not express the reporter allele for the other enzyme in the 5-HT synthesis pathway, bas-1. Because of very dim cat-1::mCherry fosmid-based reporter expression that was only detected upon anti-mCherry antibody staining, we had assigned PVW as a 5-HT-releasing neuron (Serrano-Saiz et al. 2017b). However, we failed to detect expression of our new cat-1/VMAT reporter allele in PVW. Neither did we detect expression of the mod-5(vlc47) reporter allele. Taken together, PVW either synthesizes or uptakes 5-HT by unconventional means, akin to the pharyngeal I5 neuron.

In conclusion, although there are some updates in the levels of dimorphic gene expression (PVN and ADF neuron classes), our analysis with reporter alleles does not reveal pervasive novel sexual dimorphism in sex-shared neurons compared to those that we previously identified in (Serrano-Saiz et al. 2017b). These sexual dimorphisms are summarized in Table S4.

Neurotransmitter pathway genes in glia

In vertebrates, glia can produce various signaling molecules, including neurotransmitters (Araque et al. 2014; SAVTCHOUK AND VOLTERRA 2018). There is some limited evidence for neurotransmitter synthesis in C. elegans glia. In males, it had been reported that the socket glia of spicule neurons synthesize and utilize dopamine, based on their expression of cat-2/TH and bas-1/AAAD (LINTS AND EMMONS 1999; HARE AND LOER 2004; Leboeuf et al. 2014). We confirmed this notion with cat-2/TH and bas-1 reporter alleles (Fig. 14A). Additionally, we detected expression of cat-1/VMAT reporter allele expression in these cells (Fig. 14A), indicating that these glia secrete dopamine by canonical vesicular transport. We also observed bas-1(syb5923) reporter allele expression in cells that are likely to be the spicule sheath glia (Fig. 14A), as well as in additional glia cell types in the head and tail (Fig. 14B).

Expression of neurotransmitter pathway genes in non-neuronal cell types.

Multiple neurotransmitter pathway genes show expression in glial cells (A,B) and other non-neuronal cell types (C-E). Also see Fig. S3 for whole-worm views that capture more non-neuronal expression.

(A) bas-1(syb5923), cat-2(syb8255), and cat-1(syb6486) reporter alleles exhibit expression in the male spicule glial cell types, largely consistent with previous reports.(LINTS AND EMMONS 1999; HARE AND LOER 2004; Leboeuf et al. 2014)

(B) top 6 panels: bas-1(syb5923) is expressed in additional, multiple glial cell types in the male tail. Left 3 panels: bas-1(syb5923) crossed into a pan-glial reporter otIs870[mir-228p::3xnls::TagRFP], confirming its expression in glial cells; right 3 panels: bas-1(syb5923) shows no overlap with the pan-neuronal marker component in NeuroPAL (otIs669). Bottom 2 panels: bas-1(syb5923) also shows expression in at least two glial cells in the head. A hermaphrodite head is shown here. Expression is similar in the male.

(C) In the hermaphrodite vulval region, tdc-1(syb7768) is expressed in uv1, consistent with previous reports (Alkema et al. 2005). This expression in uv1 is not observed for either cat-1(syb6486) or oct-1(syb8870). An ida-1p::mCherry integrant vsls269[ida-1::mCherry] was used for identifying uv1.

(D) Detection of eat-4(syb4257) expression in muscle cells in both sexes, most prominently in the head.

(E) cat-1(syb6486), tdc-1(syb7768), and tbh-1(syb7786) are expressed in the male somatic gonad. All three have expression in the vas deferens; additionally, cat-1 and tbh-1 are also expressed in the seminal vesicle.

We detected no expression of other vesicular transporters or neurotransmitter biosynthetic synthesis machinery in glia of either sex. This observation contrasts previous reports on GABA synthesis and release from the AMsh glia cell type (Duan et al. 2020; Fernandez-Abascal et al. 2022). We were not able to detect signals in AMsh with anti-GABA staining, nor with an SL2 or T2A-based GFP-based reporter allele for any unc-25 isoform. (Gendrel et al. 2016)(M. Gendrel, pers. comm.; this paper).

There is, however, evidence for neurotransmitter uptake by C. elegans glial cells, mirroring this specific function of vertebrate glia (HENN AND HAMBERGER 1971). We had previously shown that one specific glia-like cell type in C. elegans, the GLR glia, take up GABA via the GABA uptake transporter SNF-11 (Gendrel et al. 2016). We did not detect unc-47/VGAT fosmid-based reporter expression in the GLRs (Gendrel et al. 2016) and also detected no expression with our unc-47/VGAT reporter allele. Hence, these glia are unlikely to release GABA via classic vesicular machinery. Other release mechanisms for GABA can of course not be excluded. Aside from the snf-11 expression in GLR glia (Gendrel et al. 2016), we detected expression of the putative tyramine uptake transporter oct-1/OCT in a number of head glial cells (Fig. 8K), as well as broad glial expression of the betaine uptake transporter snf-3/BGT1 in the head, midbody, and tail (Fig. 9E,F). These results indicate tyramine and betaine clearance roles for glia.

Neurotransmitter pathway gene expression outside the nervous system

We detected expression of a few neurotransmitter pathway genes in cells outside the nervous system. The most prominent sites of reporter allele expression are located within the somatic gonad. We detected expression of tdc-1(syb7768) and tbh-1(syb7786) reporter alleles in the gonadal sheath of hermaphrodite as well as tdc-1(syb7768) expression in the neuroendocrine uv1 cells (Fig. 14C; Fig. S3), as previously reported (Alkema et al. 2005). Intriguingly, while cat-1(syb6486) is expressed in a midbody gonadal cell posterior to the vulva, likely the distal valve (Fig. 6C, Fig. S3), we observed no expression of cat-1(syb6486) in the gonadal sheath or the uv1 cells (Fig. 14C). This suggests alternative release mechanisms for tyramine and octopamine. A vertebrate homolog of the putative tyramine uptake transporter, oct-1, has been found to be located presynaptically and to co-purify with synaptosomes (Berry et al. 2016; Matsui et al. 2016), therefore indicating that this transporter may have the potential to also act in tyramine release, at least in vertebrate cells. However, we observed no expression of our oct-1 reporter allele in uv1 or gonadal sheath cells.

In the male, tdc-1(syb7768), tbh-1(syb7786), cat-1(syb6486), and oct-1(syb8870) animals also show reporter expression in the somatic gonad: while all four genes are expressed in the vas deferens, cat-1 and tbh-1, but not tdc-1 or oct-1, are expressed in the seminal vesicle (Fig. 14C, Fig. 8K). A similar pattern of cat-1(+); tbh-1(+); tdc-1(-); oct-1(-) is detected in several male-specific neurons and may indicate the usage of a novel transmitter (e.g. PEOH, see Discussion) by these cells. snf-3/BGT1 is also expressed in male somatic gonad cells, indicating that these cells could also use betaine for signaling (Fig. 9E).

The AAADs tdc-1, as well as bas-1, are also prominently expressed in the intestine, where bas-1 has been shown to be involved in generating 5-HT-derived glucosides (Yu et al. 2023). bas-1, but not tdc-1, is also expressed in the hypodermis and seam cells, as is the betaine uptake transporter snf-3 (Fig. S1G, S3). The tph-1 reporter allele expresses in a subset of pharyngeal non-neuronal cells during the L1 to L4 larval stages of development (Fig. S2), which is consistent with low levels of tph-1 transcripts detected in pharyngeal muscles in the CeNGEN scRNA dataset. Additionally, we observed previously uncharacterized eat-4/VGLUT expression in muscle cells in both sexes (Fig. 14D).

Discussion

Using CRISPR/Cas9-engineered reporter alleles we have refined and extended neurotransmitter assignment throughout all cells of the C. elegans male and hermaphrodite. We conclude that in both hermaphrodites and males, about one quarter of neurons are glutamatergic (eat-4/VGLUT-positive), a little more than half are cholinergic (unc-17/VAChT-positive), around 10% are GABAergic (unc-25/GAD-positive), and about another 10% are monoaminergic (cat-1/VMAT-positive). We compiled comprehensive lists for gene expression and neuron identities, which are provided in Table S2 for hermaphrodites and Table S3 for males. Figure 3 presents a summary of neurotransmitter usage and atlases showing neuron positions in worm schematics. Additionally, we summarize our rationale for assigning neurotransmitter usage and updates to previously reported data in Tables 1, 2, and S5. Given the complexity and nuances in determining neurotransmitter usage, we refer the reader to all the individual tables for a comprehensive description of the subject matter, rather than encouraging sole reliance on the summary in Figure 3.

Neurotransmitter synthesis versus uptake

Direct detection of neurotransmitters through antibody staining has shown that at least two neurotransmitters, GABA and 5-HT, are present in some neurons that do not express the synthesis machinery for these transmitters (Table 1, 2). Instead, these neurons acquire GABA and 5-HT through uptaking them via defined uptake transporters, SNF-11/BGT1 for GABA (Mullen et al. 2006) and MOD-5/SERT for 5-HT (Ranganathan et al. 2001; Jafari et al. 2011).

A combination of CeNGEN scRNA transcriptome and our reporter allele data corroborates the absence of synthesis machinery in these presumptive uptake neurons (Table 1, 2). One interesting question that relates to these uptake neurons is whether they serve as “sinks” for clearance of a neurotransmitter or whether the taken-up neurotransmitter is subsequently “recycled” for synaptic release via a vesicular transporter. Previous data, as well as our updated expression profiles, provide evidence for both scenarios: ALA and AVF do not synthesize GABA via UNC-25/GAD, but they stain with anti-GABA antibodies in a manner that is dependent on the uptake transporter SNF-11 (Gendrel et al. 2016). ALA expresses unc-47, hence it is likely to synaptically release GABA, but AVF does not, and it is therefore apparently involved only in GABA clearance. Similarly, RIH, AIM, and PGA express the 5-HT uptake transporter mod-5/SERT and stain for 5-HT in a MOD-5-dependent manner (Jafari et al. 2011)(this study), but only RIH, not AIM or PGA, expresses the vesicular transporter cat-1/VMAT, suggesting RIH is likely a serotonergic signaling neuron whereas AIM and PGA are clearance neurons.

Some neurons do not obviously fall into the synthesis or uptake category, most notably, the anti-GABA-antibody-positive AVA and AVB neurons (both of which conventional cholinergic neurons). None of these neurons express unc-25/GAD, nor the snf-11/BGT1 uptake transporter, yet unc-25/GAD is required for their anti-GABA-positive staining (Gendrel et al. 2016). This suggests that GABA may be acquired by these neurons through non-canonical uptake or synthesis mechanisms. Also, the AVA and AVB neurons do not express UNC-47 (Gendrel et al. 2016; Taylor et al. 2021)(this study); hence, it is not clear if or how GABA is released from them. A member of the bestrophin family of ion channels has been shown to mediate GABA release from astrocyte glia in vertebrates (Lee et al. 2010) and, more recently, in C. elegans (CHENG et al. 2024; GRAZIANO et al. 2024). However, while there are more than 20 bestrophin channels encoded in the C. elegans genome (HOBERT 2013), they do not appear to be expressed in the AVA or AVB neurons (Taylor et al. 2021).

The co-expression of a specific uptake transporter and a vesicular transporter also leads us to predict the usage of betaine as a potential neurotransmitter. Betaine is known to be synthesized in C. elegans, but is also taken up via its diet (Peden et al. 2013; Hardege et al. 2022). Betaine has documented effects on animal behavior and acts via activation of several betaine-gated ion channels (Peden et al. 2013; Hardege et al. 2022). Expression of biosynthetic enzymes suggests betaine production in at least the RIM neuron class, which also expresses the vesicular transporter cat-1/VMAT, capable of transporting betaine (Hardege et al. 2022). The expression of the betaine uptake transporter snf-3/BGT1 in CAN, AUA, RIR, ASI, and male-specific neuron PDC, coupled with their co-expression of cat-1/VMAT, suggests that several distinct neuron classes in different parts of the nervous system may uptake betaine and engage in vesicular betaine release via CAT-1/VMAT to gate betaine-activated ion channels, such as ACR-23 (Peden et al. 2013) or LGC-41 (Hardege et al. 2022). Additionally, we detected the snf-3/BGT1 reporter allele in several other neuron classes that do not co-express cat-1/VMAT. This indicates that these neurons could function as betaine clearance neurons.

Lastly, based on sequence similarity and expression pattern, we predict that the ortholog of the OCT subclass of SLC22 family, oct-1, could serve as a tyramine uptake transporter in C. elegans. We identified RIM to be the only neuron expressing an oct-1 reporter allele, suggesting that like several other monoaminergic neuron classes, RIM both synthesizes its monoaminergic transmitter, tyramine, and reuptakes it after release.

Evidence for usage of currently unknown neurotransmitters

Novel amino acid transmitters?

unc-47/VGAT is expressed in a substantial number of non-GABAergic neurons (95 out of 302 total neurons in hermaphrodites, plus 61 out of 93 male-specific neurons). However, expression in many of these non-GABAergic neurons is low and variable and such expression may not lead to sufficient amounts of a functional gene product. Yet, in some neurons (e.g. the SIA neurons) expression of unc-47 is easily detectable and robust (based on fosmid-based reporter, reporter allele, and scRNA data), indicating that VGAT may transport another presently unknown neurotransmitter (Gendrel et al. 2016). In vertebrates, VGAT transports both GABA and glycine, and the same is observed for UNC-47 in vitro (Aubrey et al. 2007). While the C. elegans genome encodes no easily recognizable ortholog of known ionotropic glycine receptors, it does encode anion channels that are closely related by primary sequence (HOBERT 2013). Moreover, a recently identified metabotropic glycine receptor, GPR158 (Laboute et al. 2023), has a clear sequence ortholog in C. elegans, F39B2.8. Therefore, glycine may also act as a neurotransmitter in C. elegans. VGAT has also been shown to transport β-alanine (Juge et al. 2013), another potential, but as yet unexplored, neurotransmitter in C. elegans. However, it needs to be pointed out that most of the additional unc-47-positive neurons do not co-express the LAMP-type UNC-46 protein, which is important for sorting UNC-47/VGAT to synaptic vesicles in conventional GABAergic neurons (Schuske et al. 2007). In vertebrates, the functional UNC-46 ortholog LAMP5 is only expressed and required for VGAT transport in a subset of VGAT-positive, GABAergic neurons (Tiveron et al. 2016; Koebis et al. 2019), indicating that alternative vesicular sorting mechanisms may exist for UNC-47/VGAT.

Novel monoaminergic transmitters?

Three neuron classes (AVL, PVX, and PVY) express cat-1/VMAT but do not express the canonical synthesis machinery for 5-HT, tyramine, octopamine, or dopamine. Neither do they show evidence for uptake of known monoamines. There are also several cat-1/VMAT-positive male-specific neurons that express only a subset of the biosynthetic machinery involved in the biosynthesis of known aminergic transmitters in the worm. That is, some neurons express cat-1/VMAT and bas-1/AAAD, but none of the previously known enzymes that produce the substrate for BAS-1, i.e. CAT-2 or TPH-1 (Fig. 1A). In these neurons, BAS-1/AAAD may decarboxylate an unmodified (i.e. non-hydroxylated) aromatic amino acid as substrate to produce, for example, the trace amine phenylethylamine (PEA) from phenylalanine (Table 2, Fig. S1A). A subset of these neurons (all being B-type ray sensory neurons) co-express tbh-1, which may use PEA as a substrate to produce the trace amine, phenylethanolamine (PEOH). PEOH is a purported neurotransmitter in Aplysia (Saavedra et al. 1977) and the vertebrate brain (SAAVEDRA AND AXELROD 1973) and can indeed be detected in C. elegans extracts (F. Schroeder, pers. comm.).

bas-1/AAAD may also be responsible for the synthesis of histamine, an aminergic neurotransmitter that can be found in extracts of C. elegans (PERTEL AND WILSON 1974). The only other AAAD that displays reasonable sequence similarity to neurotransmitter-producing AAADs is the hdl-1 gene (HARE AND LOER 2004; HOBERT 2013)(Fig. S1B), for which we, however, did not detect any expression in the C. elegans nervous system (Fig. S1D). Since there are neurons that only express bas-1/AAAD, but no enzyme that produces canonical substrates for bas-1/AAAD (tph-1/TPH, cat-2/TH; Fig. 1A) and since at least a subset of these neurons express the monoamine transporter cat-1/VMAT, bas-1/AAAD may be involved in synthesizing another currently know bioactive monoamine.

Conversely, based on the expression of tph-1, but concurrent absence of bas-1/AAAD, the pharyngeal MI neuron, hermaphrodite VC4 and VC5, and male neurons CEM and R9B may produce 5-HTP (Fig. S1A, Table 2). 5-HTP may either be used directly as a signaling molecule or it may be metabolized into some other serotonin derivative, an interesting possibility in light of serotonin-derivatives produced elsewhere in the body (Yu et al. 2023).

Additionally, 3 neuron classes (IL2, HOB, and R5B) express tbh-1 but lack expression of any other genes in canonical monoaminergic pathways, including bas-1 (Table 2). This observation further suggests the presence of non-canonical mechanisms for monoaminergic synthesis. Taken together, monoaminergic pathway genes are expressed in unconventional combinations in several neuron classes, pointing towards the existence of yet undiscovered monoaminergic signaling systems.

Neurons devoid of canonical neurotransmitter pathway genes may define neuropeptide-only neurons

We identified neurons that do not express any conventional, well-characterized vesicular neurotransmitter transporter families, namely UNC-17/VAChT, CAT-1/VMAT (the only SLC18 family members), UNC-47/VGAT (only SLC32 family member), or EAT-4/VGLUT (an SLC17 family member). Six sex-shared neurons (AVH, BDU, PVM, PVQ, PVW, RMG) and one male-specific neuron (SPD) fall into this category. Most of these neurons exhibit features that are consistent with them being entirely neuropeptidergic. First, electron microscopy has revealed a relative paucity of clear synaptic vesicles in most of these neurons (White et al. 1986; Cook et al. 2019; Witvliet et al. 2021). Second, not only do these neurons express a multitude of neuropeptide-encoding genes (Taylor et al. 2021), but they also display a dense interconnectivity in the “wireless” neuropeptidergic connectome (Ripoll-Sanchez et al. 2022).

That said, electron microscopy shows that some of the neurons devoid of conventional neurotransmitter pathway genes generate synapses with small, clear synaptic vesicles, indicative of the use of non-peptidergic transmitters (e.g. the sex-shared RMG and PVM neurons or the male-specific SPD neurons) (White et al. 1986; Cook et al. 2019; Witvliet et al. 2021). It is therefore conceivable that either conventional neurotransmitters utilize non-conventional neurotransmitter synthesis and/or release pathways, or that completely novel neurotransmitter systems remain to be discovered. Although the C. elegans genome does not encode additional members of the SLC18A2/3 (cat-1/VMAT, unc-17/VAChT) or SLC32A1 (unc-47/VGAT) family of vesicular neurotransmitter transporters, it does contain a number of additional members of the SLC17A6/7/8 (VLGUT) family (HOBERT 2013). These may serve as non-canonical vesicular transporters of more uncommon neurotransmitters or, alternatively, may be involved in modulating release of glutamate (Serrano-Saiz et al. 2020; Choi et al. 2021). Uncharacterized paralogs of bona fide neurotransmitter uptake transporters (SLC6 superfamily) may also have functions in neurotransmitter release rather than uptake. However, based on CeNGEN scRNA data, no robust or selective expression of these SLC17 or SLC6 family members is observed in these “orphan neurons”.

Co-transmission of multiple neurotransmitters

Our analysis expands the repertoire of neurons that co-transmit multiple neurotransmitters (Fig. 3). Neurotransmitter co-transmission has been observed in multiple combinations in the vertebrate brain (WALLACE AND SABATINI 2023). In C. elegans, the most frequent co-transmission configurations are a classic, fast transmitter (acetylcholine or glutamate) with a monoamine. Co-transmission of two distinct monoaminergic systems also exists. In several cases, however, it is not clear whether the second neurotransmitter is indeed used for communication or whether its presence is merely a reflection of this neuron being solely an uptake neuron. For example, the glutamatergic AIM neuron stains positive for serotonin, which it uptakes via the uptake transporter MOD-5, but it does not express the vesicular monoamine transporter cat-1/VMAT (Fig. 3, 6, 8, Table 1, 2).

Co-transmission of small, fast-acting neurotransmitters (glutamate, GABA, acetylcholine) does exist, but it is rare (Fig. 3). The most prominent co-transmission configuration is acetylcholine with glutamate, but acetylcholine can also be co-transmitted with GABA. There are no examples of co-transmission of glutamate and GABA, as observed in several regions of the vertebrate brain (WALLACE AND SABATINI 2023).

Interestingly, co-transmission appears to be much more prevalent in the male-specific nervous system, compared to the sex-shared nervous system (Fig. 3, Table S3). Remarkably, several male-specific neuron classes may utilize three co-transmitters. Such extensive co-transmission may relate to male-specific neurons displaying a greater degree of anatomical complexity compared to the hermaphrodite nervous system, both in terms of branching patterns and extent of synaptic connectivity (Jarrell et al. 2012; Cook et al. 2019). Given that all co-transmitting neurons display multiple synaptic outputs (Cook et al. 2019), it appears possible that each individual neurotransmitter secretory system is distributed to distinct synapses. Based on vertebrate precedent (WALLACE AND SABATINI 2023), co-release from the same vesicle is also possible.

Sexual dimorphisms in neurotransmitter usage

The observation of sexual dimorphisms in neurotransmitter abundance in specific regions of the mammalian brain has been one of the earliest molecular descriptors of neuronal sex differences in mammals (Mccarthy et al. 1997). However, it has remained unclear whether such differences are the result of the presence of sex-specific neurons or are indications of distinctive neurotransmitter usage in sex-shared neurons. Using C. elegans as a model, we have been able to precisely investigate (a) whether sex-specific neurons display a bias in neurotransmitter usage and (b) whether there are neurotransmitter dimorphisms in sex-shared neurons (Pereira et al. 2015; Gendrel et al. 2016; Serrano-Saiz et al. 2017b)(this paper). We found that male-specific neurons display a roughly similar proportional usage of individual neurotransmitter systems and note that male specific neurons display substantially more evidence of co-transmission, a possible reflection of their more elaborate morphology and connectivity. We also confirmed evidence for sexual dimorphisms in neurotransmitter usage in sex-shared neurons (Table S4), which are usually correlated with sexual dimorphisms in synaptic connectivity of these sex-shared neurons (Cook et al. 2019).

Neurotransmitter pathway genes in glia and gonad

Neurotransmitter uptake is a classic function of glial cells across animal phylogeny (HENN AND HAMBERGER 1971), and such uptake mechanisms are observed in C. elegans as well. Previous reports demonstrated glutamate uptake by CEPsh (Katz et al. 2019) and GABA uptake by GLR glia (Gendrel et al. 2016). We now add to this list betaine uptake by most glia, as inferred from the expression pattern of SNF-3/BGT1 (Fig. 9, Table S1).

Studies in vertebrates have also suggested that specific glial cell types synthesize and release several neurotransmitters (Araque et al. 2014; SAVTCHOUK AND VOLTERRA 2018). For example, astrocytes were recently shown to express VGLUT1 to release glutamate (DE Ceglia et al. 2023). Evidence of neurotransmitter synthesis and release also exists in C. elegans. Previous work indicated that glia associated with male-specific spicule neurons synthesize (through cat-2/TH and bas-1/AAAD) the monoaminergic transmitter dopamine to control sperm ejaculation (Leboeuf et al. 2014). Our identification of cat-1/VMAT expression in these glia indicate that dopamine is released via the canonical vesicular monoamine transporter. We also detected expression of bas-1/AAAD in additional male and hermaphrodite glia, indicating the production of other signaling substances released by these glia. bas-1 has indeed recently been shown to be involved in the synthesis of a class of unconventional serotonin derivates (Yu et al. 2023).

We observed no additional examples of neurotransmitter synthesis and release by glia, based on the apparent absence of detectable expression of neurotransmitter-synthesizing enzymes or any vesicular transporter (unc-17/VAChT, unc-47/VGAT, eat-4/VGLUT, cat-1/VMAT). Both observations are particularly notable in the context of previous reports on GABA synthesis and release from the AMsh glia cell type (Duan et al. 2020; Fernandez-Abascal et al. 2022). We were not able to detect AMsh with anti-GABA staining, nor with reporter alleles of unc-25/GAD. However, since very low levels of unc-25 are observed in AMsh scRNA datasets (Taylor et al. 2021; Purice et al. 2023), the abundance of GABA in AMsh may lie below conventional detection levels.

Outside the nervous system, the most prominent and functionally best characterized usage of neurotransmitters lies in the hermaphrodite somatic gonad, which has been shown to synthesize octopamine and use it to control oocyte quiescence (Alkema et al. 2005; Kim et al. 2021). Intriguingly, we also detected tbh-1, tdc-1, and cat-1 expression in the somatic gonad of the male, specifically the vas deferens, which is known to contain secretory granules that are positive for secretory molecular markers (Nonet et al. 1993). The presence of octopamine is unexpected because, unlike oocytes, sperm are not presently known to require monoaminergic signals for any aspect of their maturation. It will be interesting to assess sperm differentiation and function of tbh-1 or tdc-1 mutant animals. The usage of monoaminergic signaling systems in the gonad is not restricted to C. elegans and has been discussed in the context of sperm functionality and oocyte maturation in vertebrates (Mayerhofer et al. 1999; Ramirez-Reveco et al. 2017; Alhajeri et al. 2022).

Comparing approaches and caveats of expression pattern analysis

Our analysis also provides an unprecedented and systematic comparison of antibody staining, CeNGEN scRNA transcript data, reporter transgene expression, and knock-in reporter allele expression. The bottom-line conclusions of these comparisons are: (1) Reporter alleles reveal more sites of expression than fosmid-based reporters. It is unclear whether this is due to the lack of cis-regulatory elements in fosmid-based reporters or issues associated with the multicopy-nature of these reporters (e.g. RNAi-based gene silencing of multicopy arrays or squelching of regulatory factors). Another factor to consider is that neuron identification for most fosmid-based reporters was carried out prior to the introduction of NeuroPAL. Consequently, errors occasionally occurred, as exemplified by the misidentification of neuron IDs for CA7 and CP7 in previous instances (Serrano-Saiz et al. 2017b). (2) The best possible reporter approaches (i.e. reporter alleles) show very good overlap with scRNA data, thereby validating each approach. However, our comparisons also show that no single approach is perfect. CeNGEN scRNA data can miss transcripts and can also show transcripts in cells in which there is no independent evidence for gene or protein expression. Conversely, antibody staining displays vagaries related to staining protocols and protein localization, which can be overcome with reporter approaches, but the price to pay with reporter alleles is that if they are based on SL2 or T2A strategies, they may fail to detect additional levels of posttranslational regulation, which may result in protein absence even in the presence of transcripts. The existence of such mechanisms may be a possible explanation for cases where the expression of synthesis and/or transport machinery expression does not match up (e.g. tdc-1(-); tbh-1(+) neurons).

Our detailed analysis of reporter allele expression has uncovered several cases where expression of a neurotransmitter pathway gene in a given neuron class appears very low and variable from animal to animal. Such variability only exists when expression is dim, thus one possible explanation for it is that expression levels merely hover around an arbitrary microscopical detection limit. However, we cannot rule out the other possibility that this may also reflect true on/off variability of gene expression. Taking this notion a step further, we cannot exclude the possibility that expression observed with reporter alleles misses sites of expression. This possibility is raised by our inability to detect unc-25/GAD reporter allele expression in AMsh glia (Duan et al. 2020; Fernandez-Abascal et al. 2022) or eat-4 reporter allele expression in AVL and DVB neurons, in which some (but not other) multicopy reporter transgenes revealed expression of the respective genes (Li et al. 2023). Functions of these genes in the respective cell types were corroborated by cell-type specific RNAi experiments and/or rescue experiments; whether there is indeed very low expression of these genes in those respective cells or whether drivers used in these studies for knock-down and/or rescue produce very low expression in other functionally relevant cells remains to be resolved.

Conclusions

In conclusion, we have presented here the most complete neurotransmitter map that currently exists for any animal nervous system. Efforts to map neurotransmitter usage on a system-wide level are well underway in other organisms, most notably, Drosophila melanogaster (DENG et al. 2019; ECKSTEIN et al. 2024). The C. elegans neurotransmitter map presented here comprises a critical step toward deciphering information flow in the nervous system and provides valuable tools for studying the genetic mechanisms underlying cell identity specification. Moreover, this neurotransmitter map opens new opportunities for investigating sex-specific neuronal differentiation processes, particularly in the male-specific nervous system, where a scarcity of molecular markers has limited the analysis of neuronal identity control. Lastly, our analysis strongly suggests that additional neurotransmitter systems remain to be identified.

While the gene expression patterns delineated here enable informed predictions about novel neuronal functions and neurotransmitter identities, further investigations involving genetic perturbations, high-resolution imaging, complementary functional assays, and analyses across developmental stages are needed to shed further light on neurotransmitter usage. Nonetheless, this comprehensive neurotransmitter map provides a robust foundation for deciphering neural information flow, elucidating developmental mechanisms governing neuronal specification, exploring sexual dimorphisms in neuronal differentiation, and potentially uncovering novel neurotransmitter systems awaiting characterization.

Resource availability

Lead contact

Further information and requests for resources and reagents should be directed to and will be fulfilled by the Lead Contact, Oliver Hobert (or38@columbia.edu).

Materials availability

All newly generated strains are available at the Caenorhabditis Genetics Center (CGC).

Data and code availability

Any additional information required to analyze the data reported in this paper is available from the lead contact upon request.

Supporting information